Thymic selection of the T cell receptor repertoire is biased toward autoimmunity in females

Version 1

Abstract

Women represent about 80% of patients with autoimmune diseases. This may partly result from sex-based differences in T cell receptor (TCR) selection during thymocyte development, potentially influenced by hormones and the lower expression of the Autoimmune Regulator (AIRE) transcription factor in females. To investigate this, we analyzed sex-specific differences in TCR generation and selection. We examined TCR repertoires in double-positive thymocytes and single-positive thymic cells, including CD8⁺ and CD4⁺ effector T cells and regulatory T cells (Tregs), derived from male and female organ donors. Minimal sex-based differences were observed in V and J gene usage, and there were no notable differences in TCR repertoire diversity, complementarity-determining region 3 (CDR3) length, amino acid composition, or network structure. No TCR sequences were exclusive to either sex. However, female effector T cells exhibited a significantly higher prevalence of TCRs specific to self-antigens implicated in autoimmunity compared to males, while female Tregs showed a reduced frequency of such TCRs. These differences were not observed for TCRs targeting self-antigens unrelated to autoimmunity or antigens associated with cancer or viruses. Our findings highlight a sex-specific imbalance in thymic selection of TCRs with autoimmunity-associated specificities, providing mechanistic insight into the increased susceptibility of women to autoimmune diseases.

Introduction

Several studies have highlighted a sex imbalance in diseases involving the immune system. This is particularly the case for autoimmune diseases (AIDs), where 80% of those affected are women and the severity between the sexes vary depending on the pathology (1–3). In addition, men are more severely affected by infectious disease such as COVID-19 and tuberculosis (4, 5). The response to preventive immunotherapies such as vaccination, and to curative treatments, like immune checkpoint inhibitors or anti-TNF alpha, is also sex dependent (6). All of these observations suggest intrinsic biological differences in immune responses between males and females.

A variety of factors have been identified that could help explain these observed sex-based differences. Among these, sex hormones have been shown to influence immune responses by acting on immune cells that express specific receptors to these hormones (7–9). Sex hormones could directly influence the adaptive immune response at the level of thymocyte differentiation and selection by acting on the expression of Autoimmune Regulator (AIRE) in thymic epithelial cells. AIRE drives the thymic expression of otherwise tissue-specific antigens, contributing to both the negative selection of effector cells and the positive selection of Tregs that recognize such antigens (10). Testosterone increases AIRE expression in medullary thymic epithelial cells, whereas estradiol decreases it (11, 12). Furthermore, a recent study has shown that the transcriptomic profiles of these specialized antigen-presenting cells in the thymus differ between males and females (13). This clearly suggests that there could be differences in the selection of the T cell receptor (TCR) repertoire between males and females.

Despite these findings, little is known about potential differences in the TCR repertoire between the sexes. The generation of the TCR is a key process in T-cell development that takes place in the thymus. The TCR consists of two chains, alpha (TRA) and beta (TRB), each of which resulting from a random genetic rearrangement involving the V (variable), D (diversity) [for the TRB], and J (joining) genes. The combination of these gene segments is accompanied by insertions and deletions, creating a highly diverse region, the Complementarity Determining Region 3 (CDR3) (14). It is this region of the TCR that mostly interacts with the peptide, a critical interaction for antigen recognition and subsequent T cell activation. The ability to specifically recognize diverse antigens via the CDR3 is central to the efficiency and specificity of the adaptive immune response.

Because TCRs are randomly generated, many could be useless while some could be harmful by recognizing self-antigens and causing autoimmunity. The current paradigm is that during thymocyte development, thymocytes are selected based on the avidity of their TCRs for antigens presented by cortical and medullary thymic epithelial cells. Thymocytes expressing TCRs with too low or too high avidity are eliminated (negative selection), while those with an intermediate avidity are positively selected and mature into single-positive cells, either CD4+ or CD8+. Thymocytes with the highest of these intermediate affinities develop in Tregs. Since AIRE plays a major role in these processes, sex differences in AIRE expression could have a major impact on thymic selection.

The TCR repertoire of female and male peripheral T cells have shown that their diversity, particularly that of the TRB repertoire, is influenced not only by the age of the individuals but also by their sex (15–18). However, another study showed no association between the sex of individuals and the diversity of their TCR repertoire (19). Nevertheless, the study of peripheral blood TCRs reflects that of an immune system that has gone through multiple immune responses. For example, differences observed between males and females for responses to self-antigens associated with autoimmunity could reflect either the cause (the TCRs drive the disease) or the consequence (the TCRs are expanded because of the disease).

To assess whether there are significant differences in the generation and thymic selection of the TCR repertoire between males and females, we therefore focused on analyzing the TCR repertoires of thymocytes. These cells are the most recent product of T cell development and have not yet been involved in immune responses. Their repertoires thus represent the mechanisms of TCR generation for the analysis of CD4/CD8 double positive (DP) cells, and of TCR selection for CD4 and CD8 simple positive (SP) T cells. We generated a unique data set comprising the sequences of the TRA and TRB chains of DP cells, CD8 and CD4 SP effector cells, and CD4 SP Tregs from organ donors, both infants and adults. The analysis of TRA and TRB repertoires using state-of-the-art strategies revealed no sex differences in repertoire generation (DP cell repertoire), but a bias towards autoimmunity in the selection of the CD8 SP T cells in females.

Results

Minimal sex-based differences in TCR V and J gene usage across thymic cell subsets

The first level of diversity in generating TCRs is the usage of V and J genes (20). We analyzed the usage of TRA and TRB V and J genes and their combinations in each thymocyte population (Figure 1).

Schematic overview of the generation of thymic TCR dataset and analytical pipeline.

Top Panel: Generation of the thymic TCR dataset. From deceased human thymuses of males and females, we isolated key T-cell subtypes through cell sorting. These subtypes included double-positive (DP) cells (CD3+ CD4+ CD8+), single-positive (SP) CD8+ cells (CD3+ CD4-CD8+), SP CD4+ cells (CD3+ CD4+ CD8-), and further separated SP CD4+ cells into T effector (Teff, CD3+ CD4+ CD8-CD25-) and Treg (CD3+ CD4+ CD8-CD25+) cells. TCR libraries were generated from the RNA of each cell population using rapid amplification of cDNA-ends by PCR (5’RACE PCR). Following sequencing, data preprocessing involved quality sequencing checks, contig alignment, and quality control. The final dataset comprised 20 DP samples (male-to-female ratio of 1:1), 21 SP CD8+ samples (1.1:1), 6 SP CD4+ samples (1:5), 16 SP CD4 Teff samples (1.67:1), and 14 SP CD4 Treg samples (1.33:1). Males are depicted in violet and females in orange. Bottom Panel: Analytical pipeline. We compared the TCR repertoires of males and females across various dimensions. General aspects of the TCR repertoire were evaluated, including diversity, gene usage, CDR3aa length distribution, and aa usage within the CDR3 region. Additionally, we analyzed the probability of sequence generation, the TCR repertoire structure based on CDR3aa sequence similarity, identification of differentially expressed TRB CDR3aa motifs between sex, and TRB CDR3aa sequence specificity.

Males and females had a relatively similar V gene usage as attested by the absence of a clear separation in the Principal Component Analysis (PCA), for both TRA and TRB, and across all studied cell subtypes (Figure 2A). However, we observed some individual differences in V and J gene usage between males and females, with some being specific to certain cell subtypes (Supplemental Figures 2-5), and others observed across multiple cell subtypes (Supplemental Figures 2-5). For example, TRBV6-5 was more highly expressed in females compared to males in DP cells (Supplemental Figure 2C) and CD8 SP cells (Supplemental Figure 3C).

Comparable overall TCR gene usage between males and females.

(A) Principal Component Analysis (PCA) derived from the distribution of TRAV (left) and TRBV (right) gene usage frequencies across sex groups (males vs females), showing results for DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and CD4 Treg (N = 14) cells (displayed from top to bottom). Each point represents an individual. Ellipses indicate 95% confidence intervals. (B) Heatmap showing the Jensen-Shannon Divergence (JSD) score between samples, derived from the distributional usage of TRAV-TRAJ (left) and TRBV-TRBJ (right) gene associations in DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and CD4 Treg (N = 14) cells (displayed from top to bottom). Hierarchical clustering was performed using the Euclidean distance and the complete linkage method. Males are shown in violet; females in orange.

Next, we compared the usage of TRAV-TRAJ and TRBV-TRBJ gene combinations between all individuals using the Jensen-Shannon Divergence (JSD) score (Figure 2B). The hierarchical clustering showed no clear separation between male and female individuals for any of the chain or cell subtype, indicating no significant differences in the gene combination usage between males and females (Figures 2B). A similar observation was made when directly analyzing the frequency of these V-J gene combinations (Supplemental Figures 2E-5E, 2F-5F).

Comparable TCR repertoire diversity between males and females

A high diversity of the TCR repertoire is essential for generating a broad potential reactivity against a variety of antigens. We thus compared the TCR repertoire diversity between males and females. To achieve this, we analyzed Rényi diversity curves (from zero to infinite 𝛼 parameters) that measure various aspects of TCR repertoire diversity. At different 𝛼 values, the Rényi index places more or less weight on the contribution of the most frequent clonotypes. Higher 𝛼 values emphasize on the most abundant clonotypes, thereby revealing the impact of expanded clones on the overall diversity.

Rényi curves showed comparable diversities between sexes in DP and CD4 Teff cells, for both TRA and TRB, and a difference in CD8 and CD4 Treg cells (Supplemental Figure 6). In CD8 cells, the curves begin to diverge at low 𝛼 values (Supplemental Figure 6), with an earlier inflection in males, indicating a less balanced distribution of TCR clones. This suggests that in males, fewer clones dominate the repertoire at lower diversity thresholds compared to females. In Tregs, while the inflection point is similar, the difference seems to be characterized by higher diversity and richness in females compared to males (Supplemental Figure 6). However, no significant differences were observed for the Shannon diversity index (𝛼 = 1), the Simpson index (𝛼 = 2), and the Berger-Parker index (𝛼 = infinite), for both chains and all cell subtypes (Figure 3B). These results indicate that the thymic TCR repertoire diversity is similar between males and females.

Comparable thymic TCR repertoire diversity between males and females.

Boxplots display Shannon (A), Simpson (B), and Berger-Parker (C) index values for TRA (left) and TRB (right), across thymic T cell subtypes, displayed from top to bottom, in DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and CD4 Treg (N = 14). Each point represents the median value from 50 rarefactions per sample. Statistical analysis (Wilcoxon test) showed no significant sex bias in TCR repertoire diversity (p > 0.05). Males are shown in violet; females in orange.

Comparable CDR3aa length distribution in male and female TCR repertoires

Most of the repertoire diversity is generated by the random addition of nucleotides within the VDJ junctions, generating the hypervariable CDR3 region that is interacting with the peptide-MHC complex, and is mostly driving the TCR specificity. The CDR3 length directly influences the TCR’s ability to interact with the peptide (21). We thus studied the distribution of CDR3 aa (CDR3aa) sequence lengths between males and females to identify potential TCR generation and/or selection biases (Figure 1).

Our results showed that males and females have a comparable distribution of CDR3aa lengths for both TRA (Supplemental Figure 7) and TRB (Figure 4A). CDR3aa sequences in TRA were shorter than those in TRB, as previously described in peripheral blood T cells (22, 23). Although there were subtle differences in certain CDR3aa lengths between males and females, these variations were minimal, involving sequences that represent less than 2% of the total TCR repertoire (Figure 4A, Supplemental Figure 7). Altogether, the overall distribution of CDR3aa length is similar between males and females.

CDR3aa length and amino acid composition of TRB CDR3s in males and females.

(A) Distribution of TRB CDR3aa length usage in DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and Treg (N = 14) SP cells (displayed from top to bottom). Stars indicate statistical differences between males and females based on the p-value of the Wilcoxon test (*: p < 0.05, **: p < 0.01). (B) Amino acid (aa) usage between males and females represented as the log2 fold change of the median usage of females over males for each aa in the p108 to p114 CDR3aa region for TRB. A line at log2 fold change = 0 indicates the direction of the difference in usage frequency. Bars are colored according to the hydropathy class of the aa (neutral in gray, hydrophilic in blue-green and hydrophobic in gold). Stars indicate statistical differences of usage between males and females based on the p-value of the Wilcoxon test (*: p < 0.05, **: p < 0.01). (C) Cumulative hydrophobic aa usage for each individual at p109 and p110 positions in TRB across thymic T cell subtypes, including DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and CD4 Treg (N = 14). Statistical analysis showed no significant sex bias (Wilcoxon test: p > 0.05 for the two positions). Males are depicted in violet; females in orange.

Minimal differences in CDR3aa usage in thymic TCR repertoires of males and females

We analyzed the aa usage in the FG loop of the CDR3, which is in contact with the peptide-MHC complex (24). We classified the aa according to their hydropathy property: hydrophilic, hydrophobic, or neutral (25). We observed significant differences in the usage of some aa between males and females for both TRA (Supplemental Figure 8) and TRB (Figure 4B). In DP cells, the differences preferentially involve hydrophobic aa: females exhibit higher alanine (A) usage in TRA and lower in TRB compared to males (Figure 4B, Supplemental Figure 8). Additionally, females show increased usage of phenylalanine (F) and isoleucine (I) in TRB (Figure 4B). Furthermore, other aa with varying hydropathy characteristics are differentially used between males and females in CD4 Teff and Treg populations (Figure 4B, Supplemental Figure 8).

We then analyzed in more detail the usage of aa at positions p109 and p110 as the presence of hydrophobic aa at these two positions in TRB has been associated with a stronger recognition of self-antigens (26–28). We did not observe any significant differences between males and females in the usage of hydrophobic aa at these two positions, across all cell subtypes in TRB (Figure 4C).

Slight sex-based variations in thymic TCR generation probability distribution

Another approach to evaluate biases in the generation and selection of the TCR repertoire is to compare the probability of generation (Pgen) of CDR3 nucleotide sequences (Figure 1). Pgen represents the likelihood of a sequence being produced, according to models of V(D)J recombination (14, 29). For DP and CD8 cells only and for each TCR chain, a sequence generation model was created using 100,000 non-productive nucleotide sequences. Based on this model, the Pgen of all productive TCR sequences composing each repertoire was calculated.

When comparing the distribution of Pgen between males and females, we observe a very slight difference of distribution characterized by a nearly null Kolmogorov-Smirnov score (DK-S< 0.05) (Figure 5). This difference is more pronounced in the TRB compared to the TRA chain (Figure 5). As this small difference could reflect differences linked to variation among individuals irrespective of their sex, we generated 20,000 permuted mixed-sex groups with our dataset. We found the same significant distribution difference of Pgen sequences between these mixed groups population (DP – TRA: DK-S = 0.0183 ± 0.0145 (μ ± sd), DP – TRB: DK-S = 0.0190 ± 0.0161, CD8 – TRA: DK-S = 0.0227 ± 0.0153, and CD8 – TRB: DK-S = 0.0259 ± 0.0140). However, the difference observed in the Pgen distribution between males and females is higher than that between the mixed groups in the DP TRB.

Probabilities of generation of TCRs in males and females.

The figure shows the distribution of Pgen (probability of generation) sequences between males and females for TRA and TRB in DP and CD8 cells. A V(D)J recombination model was created using 400,000 non-productive random sequences derived from the nonproductive sequences of all individuals for each chain, both for DP and CD8 cells (65). This model was then used to calculate the Pgen of sequences for each individual (66). The overall distribution comparison between males and females was tested using the Kolmogorov-Smirnov test, with the D value and associated p-value indicated. Males are depicted in violet and females in orange.

Similar network structure of thymic TCR repertoires in males and females

Next, we evaluated whether thymic selection in males and females is directed towards similar or distinct sequences. For each TCR repertoire, CDR3aa sequences are connected if they differ by one aa, i.e. have a Levenshtein distance of =1 (Figure 1). Such linked sequences have high probability to share the same specificities (30). A random downsampling to the smallest sample was performed to compare sample of similar sizes, with 100 iterations for each, detailed in Supplemental Table. The shapes of the network are depicted in Figure 6A.

Comparable TCR repertoire network structure based on CDR3 amino acid sequence similarity between males and females.

For each sample, 100 random subsamplings were performed on the minimum number of CDR3aa per cell subtype. Two CDR3aa are linked if they have a Levenshtein distance of one. (A) TRA Network structure of subsampled TCR repertoire of a male subject for DP, CD8, CD4 Teff and CD4 Treg SP (from left to right). Each point represents a CDR3aa. (B-C) Comparison of the proportion of linked sequences (B) and network density (C) between males and females samples, in DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and CD4 Treg (N = 14). Each point represents the median value from 100 subsampling iterations for each sample. Statistical analysis using the Wilcoxon test revealed no significant sex differences for these two metrics (p > 0.05). Males are depicted in violet; females in orange.

To analyze more precisely these clustering we used various metrics. When comparing the number of clustered sequences in each individual network between males and females, no significant differences were observed for any cell type in either chain (Figure 6B). Similarly, the degree of sequence similarity, reflected by the density of edges between sequences within an individual’s TCR repertoire, was comparable between males and females (Figure 6C). These findings indicate that there are no significant sex-specific differences in the similarity of CDR3aa sequences, suggesting that males and females produce CDR3aa sequences with the same degree of variability within their TCR repertoires.

Absence of sex-specific CDR3aa motifs in thymic TCR repertoires of males and females

We then sought whether we could detect CDR3aa sequence motifs that were more present in one group compared to the other, in order to evaluate whether the selection of CDR3aa was biased towards particular sequences differently between males and females (Figure 1). To identify sequences with shared characteristics that might indicate a probable common specificity among individuals of the same group, the analysis focused exclusively on CDR3aa sequences of the TRB, as this region has a greater influence on TCR specificity (31, 32). Local motifs were defined using Gliph2 (33), which are a sequences of three to five aa, as well as global motifs, which are a sequence of more than three aa where, at one position, an aa can be substituted by another aa if it has a positive BLOSUM62 score (34).

We found several hundred motifs differentially expressed between males and females for each cell subpopulation. Larger number of motifs were specific to females, ranging from 426 motifs for DP to 278 motifs for CD4 Tregs, compared to males, ranging from 328 motifs for CD4 Teffs to 34 motifs for CD4 Tregs (Figure 7A). The identified motifs are mainly local motifs and mostly restricted to their respective sex group, meaning that motifs identified in females are found almost exclusively in females, and vice versa (Supplemental Figure 9).

Thymic TRB TCR sex-associated motifs.

Different structural motifs found differentially expressed between males and females in our dataset. We distinguish local motifs as distinct aa sequences and global motifs as motif region with one variable aa position maintaining a BLOSUM62 score ≥ 0. (A) Number of male and female associated motifs by cell subset. (B) Euler diagram illustrating the distribution and overlap between all sex associated motifs. Numbers indicate the number of motifs in overlap zones. (C-D) Validation of these sex associated TRB CDR3aa motifs. Heatmap of all these TRB CDR3aa motif usage in the external thymic pediatric dataset (35, 36) (C) and those of TRB CD8 in the peripheral dataset (D). Sex and total CDR3aa number are depicted by sample. Males are depicted in violet and females in orange then local motifs in blue-green and global motifs in magenta.

We observed minimal motif overlap between different cell subtypes (Figure 7B). Only two female associated motifs overlapped: one between CD4 Teffs and CD4 Tregs, and another between CD4 Teffs and CD8 cells (Figure 7B). In contrast, no overlap was found between male-associated motifs (Figure 7B). Additionally, we noted that one motif associated with female CD8 cells motif overlapped with a motif associated with male CD4 Teff cells (Figure 7B).

We next evaluated the usage of the differentially expressed motifs in external datasets. No public thymic TCR dataset reports repertoires according to cell subsets. We thus tested these motifs on TCRs from pediatric bulk thymocytes, which contains TCR data from children aged two to eight months with a male-to-female ratio of 5:3 (35, 36). We calculated the usage of these differentially expressed motifs in each sample of this dataset. We were unable to separate individuals by sex using these motif usages (Figure 7C).

We further evaluated these motifs using a peripheral blood TCR dataset from healthy individuals, where CD8, CD4 Teff, and CD4 Treg cells had been sorted. The usage of these motifs could not distinguish males from females across all cell subsets (Figure 7D, Supplemental Figure 10).

Sex-biased enrichment of TCRs associated with autoimmune diseases and bacterial antigens

We looked for the usage of TCRs with known specificity as represented in IEDB, Mc-PAS and VDJdb public databases (37–39). We compiled the TCRs from these three databases, retaining only those with high sequence reliability scores (see Materials and Methods). We focused on TRB sequences only, due to their greater representation in the databases (31, 40). We identified 55 368 unique TRB CDR3aa sequences with high reliability and specificity assignment scores for at least one specificity.

We classified the TCR specificities based on the antigen they recognize: viral or bacterial peptides, or human peptides overexpressed in cancers, autoimmune diseases (AIDs), or neither of these diseases (Supplemental Figure 11A). We looked for these TRB CDR3aa sequences in our thymocyte dataset. Depending on the individuals and cell subtypes, they represented around 0.82% to 3.58% of the TRB repertoires.

We then compared the distribution of these CDR3aa between males and females for each cell subtype. We observed a significant enrichment of unique TRB CDR3aa sequences specific to self-antigens not associated with pathologies and those associated with cancers in our thymic TCR dataset, for both females and males (Supplemental Figure 11B). This enrichment is observed across all cell subtypes and is approximately 1.25 to 1.5 times higher than the baseline distribution of specificities in the pooled database (Supplemental Figure 11B). Strikingly, in the TCR repertoires with one specificity only, we observed in CD8 SP cells, a significantly higher proportion of unique TRB CDR3aa sequences associated with AIDs of females compared to males (Figure 8A, Supplemental Figure 12). Additionally, in the specific TCR repertoires only, there was a trend towards a higher proportion of thymic sequences specific to bacterial compounds in female CD8 SP cells compared to males and an opposite trend observed for CD4 Tregs SP (Figure 8A, Supplemental Figure 12). These differences observed in CD8 SP cells persist when examining the usage of these sequences (i.e. the cumulative usage frequency), with significantly higher usage of sequences specific to bacterial compounds and those associated with AIDs in females compared to males (Figure 8B, Supplemental Figure 12).

Sex-biased enrichment of TCRs specific for known antigens.

From a pooled and curated database, an exact match with this database infers the specificity of TRB CDR3aa of our thymic dataset. Many specificity groups are defined with the nature of the antigen peptide targeted (bacteria, virus, autoimmune disease [AID], cancer and self-peptide no associated to disease [human]). This analysis compares the distributions of the proportion of unique TRB CDR3aa sequences with a specific specificity (A) and their usage (B) between females and males across cell subtypes, using the log2 fold change of the median values (females over males), following each specificity group. These groups of specificity are additionally classified as microorganism in top (bacteria in gold and virus in light blue) and self at bottom (AID in magenta, cancer in red, human in blue-green). Polyspecific CDR3aa are defined here as CDR3aa capable of recognizing multiple antigens from different organisms (for no self-antigens) or from different specificity groups (e.g. categorized microorganisms, categorized self-antigens, allergens…). Proportion of unique polyspecific CDR3aa in the specific repertoire (C) and their usage (D) is compared between males and females, in DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and CD4 Treg (N = 14) cells. Stars indicate statistical differences between males and females based on the p-value of the Wilcoxon test (*: p < 0.05, **: p < 0.01). Males are depicted in violet and females in orange.

Donor age had no significant effect on the sex-biased enrichment of self-and bacteria-specific TRB CDR3 sequences in CD8 SP and CD4 Treg SP cells (p ≥ 0.52), indicating that the observed differences are age-independent (Supplemental Figure 13). We also investigated for the presence of polyspecific TCRs, which are capable of recognizing multiple antigens from different organisms (41–43). These sequences were enriched across all cell subtypes, compared to their representation in the reference database (Supplemental Figure 11C). However, no significant differences were observed in the proportion or usage of these polyspecific sequences between males and females. Interestingly, an inverse usage trend was noted between CD8 and CD4 Treg SP cells: females exhibited higher usage of these sequences in the CD8 SP repertoire but a lower usage in the CD4 Treg SP repertoire compared to males (Figure 8C-8D).

In conclusion, our analysis of TCR specificity reveals significant sex-specific differences in the thymic selection of CD8 effector T cells specific for autoimmune and bacterial antigens that are overrepresented in female versus males.

Discussion

It is striking that so little is known about possible differences in the TCR repertoire between sexes given the yet unexplained much higher prevalence of autoimmune diseases in females. Exploring the hypothesis that the generation and/or the selection of the TCR repertoires may be different between sexes requires studying the TCR repertoire where they are formed and selected, in the thymus. We thus generated a unique and valuable dataset comprising thymic samples collected from deceased organ donors of various ages, and young children with cardiac surgery. Most importantly, we separated the various thymocytes population to be able to study both TCR repertoire generation at the level of DP cells, and repertoire selection at the level of CD8, CD4 Teff and Treg SP cells. We performed bulk sequencing of the TCRs from these cells, the only method that yet generates enough sequences for sensitive comparison. Indeed, single cell sequencing that analyzes on a few thousand cells would not have allowed making the analyses that can be performed with hundreds of thousand cells.

TCR generation

Our analyses first revealed similarities rather than dissimilarities in TCR generation. There were no significant sex differences for most of the variables that we analyzed comparing males and females DP cells’ repertoires, except for differences in the usage of few V and J genes from both TCR chains. However, these few differences are likely a result of random variation, considering that we performed hundreds of comparisons. This is supported by the fact that there are no significant differences in the usage of TRBV-TRBJ and TRAV-TRAJ gene associations between males and females. We also observed no differences in diversity indexes in DP cells’ repertories between males and females. Altogether, the overall combinatorial diversity of the TCR repertoire remains comparable between sexes.

A detailed analysis of the CDR3aa region in DP cells showed a differential usage of some hydrophobic aa in the FG loop region for both TCR chains. However, no significant sex-based differences were found in the usage of hydrophobic, hydrophilic, or neutral aa at the critical p109 and p110 positions in TRB that have been described as influencing self-antigen recognition due to the impact of hydrophobic aa on TCR self-reactivity (26–28). These findings suggest that the subtle sex-based differences in the FG loop’s aa usage are minor and do not significantly impact the overall characteristics of the CDR3 region in a sex-dependent manner. The Pgen analysis of DP cells revealed only minimal differences in the distribution of Pgen between males and females, which were similar to those observed in control groups. This indicates that the mechanisms governing TCR generation are largely conserved across sexes, with only minor, likely random, deviations.

Additionally, although we identified TRB CDR3aa sequence motifs differentially expressed between males and females at the DP cell stage in our dataset, these motifs could not reliably differentiate individuals by sex in external datasets. This indicates that they are more dataset specific than sex-specific. This might also be explained by the different characteristics of our data and other dataset. For example, in the Arstila dataset, the analyses were performed on unsorted cells from young subjects. Younger individuals typically have a more diverse and richer TCR repertoire, with greater repertoire sharing in the periphery, which could obscure sex-specific patterns (16, 44). This suggests that these motifs might be context-specific or influenced by other factors such as age, genetic background, or environmental exposures. Moreover, we did not observe any sex-specific differences in CDR3 TRB specificity usage in DP cells’ repertoires. In summary, our results from DP cells analyses could not identify any relevant sex differences in TCR generation.

TCR selection

We identified three genes with differential usage between males and females. While some of these differences could be random, resulting from multiple comparisons, the preferential use of TRBV6-5 in females has previously been observed in the peripheral TCR repertoire (45). Notably, this gene is overrepresented in female lupus patients compared to their controls (51,52). Further research is needed to validate this observation and then elucidate its significance.

Our most striking observation relates to significant sex-specific differences in specificities of CD8 T cells, with a biased CD8 selection process in the thymus favoring sequences associated with autoimmune diseases and bacterial antigens in females. The fact that this bias does not affect DP cells indicates that it is related to TCR selection rather than generation. Its relevance is strongly supported by the fact that (i) there is no such bias for other self-antigens that are not associated with autoimmunity and (ii) the bias affects both the effectors and regulators of autoimmunity with a directionality compatible with the increased prevalence of autoimmune disease in females. TCRs specific for autoimmunity related self-antigen are increased in the CD8 effector T cells while decreased in the CD4 Tregs. It is worth nothing that some bacterial antigens, but not viral antigens, are also affected by this bias. This could potentially be explained by the fact that bacterial antigens are expressing numerous mimotopes of self-antigens and are known to contribute to the shaping of the TCR repertoire. The observed bias towards bacterial antigens in female CD8 T cells could likewise provide more effector cells specific for bacterial antigens that are mimics of self-antigens, contributing to autoimmune disease development by molecular mimicry (48). In this line, antigens linked to celiac disease (CeD) and type 1 diabetes (T1D) are highly represented in the database and these two AIDs have been linked to bacterial components that may mimic self-antigens (49–53). Similarly, it is well-established that the gut microbiota influences thymic development of T cells (54). Sexual hormones could influence this biased selection by affecting the composition of the gut microbiota (55) and/or by affecting the selection of the TCRs specific for these antigens.

Finally, we have recently described polyspecific TCRs that can respond to multiple unrelated viral antigens and hypothesized their possible involvement in autoimmunity (43). We observed a tendency towards higher usage of polyspecific sequences in the CD8 SP repertoire of females, although not statistically significant. This may indicate a preferential cross-activation of CD8 T cells in females, potentially enhancing protection against pathogens as well as increasing the risk of heightened inflammatory reactivity and autoimmunity.

Conversely, the frequency of polyspecific TCR sequence usage was lower in female CD4 Tregs SP cells. This could limit CD4 Treg antigen coverage and be detrimental to prevent autoimmunity. However, little is known about polyspecific TCRs and these observations warrant further investigation.

To further evaluate the robustness of our findings, we performed additional analyses to assess whether donor age could confound the observed sex-specific differences. A potential limitation of our study is the relatively wide age range of donors. However, the absence of age-related clustering in repertoire features (data not shown), and the lack of age impact on CD8 and CD4 Treg bacterial-specific TRB usage and CD8 and CD4 Treg self-associated to autoimmune disease-specific TRB usage patterns (Supplemental Figure 13), argue against age as a confounding factor. These results support an intrinsic sex bias in thymic TCR selection rather than an age-related effect.

While our findings provide valuable insights into sex-specific differences in thymic selection of TCR repertoires, our TCR specificity analyses were limited. Indeed, we only studied the TRB CDR3aa region of the TCR, which is the most critical for defining TCR specificity (31, 32) while both the TRA and the TRB chains contribute to TCR specificity (31, 32). This limitation is intrinsic to the limited number of TRA with assigned specificity in databases and to the fact that, given the nature of our studies, there is a need to study large repertoires, which is currently not feasible by single cell TCR sequencing. However, as methods for inferring specificity continue to evolve, they will likely allow more refined analyses of our dataset in the future (56). In addition, the continuous development of the TCR database should enrich these analyses.

Altogether, our analyses of TCR specificity revealed a biased selection in females towards sequences associated with autoimmune diseases, potentially underlying the higher prevalence of autoimmune diseases in females. Future research should focus on validating these observations in larger datasets.

Methods

Samples

Thymus samples were obtained from twenty-two organ donors aged between 73 days and 64 years, without any specific pathologies (Figure 1). Adult samples were collected post-surgery at the Cardiac Surgery Departments of many hospitals in France, after approval by the Agence de Biomédecine and the Ministry of Research (#PFS14-009). Pediatric samples were obtained after total thymectomy during cardiac surgeries at Necker Hospital. The male-to-female across donors is balanced to 1:1, although it ranges from 1.1:1 and 1.33:1 depending on the cell subtypes analyzed. Age distributions of male and female donors were compared using two-sample Kolmogorov–Smirnov tests, with no significant differences observed across thymic subsets (p > 0.46), indicating no age-related confounding (Supplementary Figure 1).

Thymocyte isolation and RNA extraction

Thymic cell suspensions were prepared through automated tissue dissociation using gentleMACS dissociators (Miltenyi®) or by mechanical disruption, followed by filtration through a 70-µm nylon mesh. Cells were stained with anti-CD3 (AF700), anti-CD4 (APC), anti-CD8 (FITC), and for some samples, anti-CD25 (PE) antibodies. Cell sorting was performed by fluorescent activated cell sorting (FACS) on a Becton Dickinson FACSAriaII, achieving a purity of >85%: Populations sorted included DP CD3+ (CD3+CD4+CD8+), SP CD8+ (CD3+CD4-CD8+), SP CD4+ (CD3+CD4+CD8-), with further separation into SP CD4 Teff (CD3+CD4+CD8-CD25-) and SP CD4 Treg (CD3+CD4+CD8-CD25+) for sixteen of samples (Figure 1). For the sorted SP CD4, the Treg percentage varied between 5.8% and 16% across samples, with an average of 9.63%. Six SP CD4 samples were not sorted into Teffs and Tregs, we thus considered them as SP CD4 Teff by assuming a purity of Teff within the total SP CD4 around 90%.

RNA was isolated using the RNAqueous extraction kit (Invitrogen®), according to the manufacturer’s protocol. The RNA were quantified and sample integrity were determined on Nanodrop (ThermoFisher) or on Tapestation 4200 (Agilent, RNA screentape).

TCR library preparation and sequencing

TCR library preparation and sequencing were performed as described by Barennes et al. (57). Briefly, RNA was processed using the SMARTer Human TCR a/b profiling v1 kit (Takarabio) according to the manufacturer’s protocol. Amplicons were purified with AMPure XP beads (Beckman Coulter), quantified, and their integrity assessed using 2100 Bioanalyzer System (Agilent, DNA 1000 kit) or Tapestation 4200 (Agilent, D1000 screentape). Bulk Next-Generation Sequencing was performed using either a Hiseq 2500 (Illumina) with SR-300 protocol plus 10% PhiX on the LIGAN-PM Genomics platform (Lille, France), or a Novaseq 6000 (Illumina) with PE-250 protocol plus 10% PhiX on the LIGAN-PM Genomics platform or ICM platform (Paris, France).

Data processing

Raw FASTQ/FASTA files were aligned to TRA and TRB using MiXCR (v3.0.13, RNA-seq parameters), which corrects for sequencing and PCR errors (58). Samples with fewer than 1,000 CDR3aa sequences, an imbalanced TRA/TRB ratio (<1/9), or a low sequence count relative to sorted cell counts (< 0.9 ratio) were removed. Sequences with CDR3aa lengths outside 6-23 amino acids were also discarded. To standardize clonotype counts and unique number of clonotypes across sequencing techniques, initial counts for each clonotype were reduced by 1 to exclude those with counts of 0 (singletons).

Additionally, due to high similarity between certain TRBV sequences, which could bias the analysis depending on the sequencing method used, TRBV06-2 and TRBV06-3 were merged as TRBV06-2/3, and TRBV12-3 and TRBV12-4 were merged as TRBV12-3/4 (59).

Gene usage analysis

For both TRA and TRB chain across all cell subtypes, we analyzed the V and J gene usage and those of the VJ gene associations (Figure 1).

Where:

Σ𝑖𝜖𝐺𝑐𝑖 represents the sum of the counts of the clonotypes using the interested gene or gene association

Σ𝑗𝜖𝑆𝑐𝑗 represents the total sum of the counts of all clonotypes in the sample

In this notation, 𝐺 is the set of clonotypes using the interested gene, 𝑆 is the set of all clonotypes in the sample, and ci and cj are the counts of the clonotypes i and j, respectively. The usage of the V gene and the VJ gene combinations was investigated through dimensional reduction using Principal Component Analysis (PCA) (MASS package v7.3-60, factoMineR package v2.8, factoextra package v1.0.7). Confidence ellipses were represented with a 5% alpha risk. Additionally, for VJ gene combinations usage, a heatmap was constructed using the Jensen-Shannon Divergence (JSD) distance to measure the distance between samples based on their gene usage distribution (complexHeatmap package v2.16.0, philentropy package v0.7.0). Hierarchical clustering was performed using the Euclidean distance and the complete linkage method.

Diversity analysis

For the TCR repertoire diversity analysis, to correct for sequencing depth bias and differences in cellular richness between individuals, rarefactions were performed fifty times for each sample, using X clonotypes, where X represents the effective diversity number (60).

Effective diversity was calculated as the exponential of the Shannon index (𝑒𝐻) (61). Rényi indices (11 indices from 0 to Infinity) were calculated for each rarefaction (62).

The median Shannon index, Simpson index and Berger-Parker index were specifically compared between sex groups.

Distribution of CDR3aa length analysis

Where

Σ𝑖𝜖𝐿𝑐𝑖 represents the sum of the counts of the CDR3 characterized by the amino acid length of interest.

Σ𝑗𝜖𝑆𝑐𝑗 represents the total sum of counts of all CDR3s per sample.

In this notation, 𝐿 is the set of CDR3s characterized by the amino acid length of interest, 𝑆 is the set of all CDR3s in the sample, and ci and cj are the counts of the CDR3s i and j, respectively.

Usage of amino acid in CDR3 region

For each CDR3aa sequence, alignment was performed following the IMGT convention (63). The aa usage is defined at different scales. Firstly, usage at the sequence level, within the CDR3 loop region (positions p108 to p114), involved in MHC-peptide complex interactions, assessed for CDR3aa sequences 9-21 aa long in both TRA and TRB chains (63). This aa usage was defined as the percentage of their representation within the sequence as follow: Where 𝑛𝑎𝑎 is the count of the specific amino acid within the sequence and L is the total length of the sequence.

Secondly, the CDR3aa usage is calculated at particular positions, the p109 and p110 in the repertoire of each individual. The usage of each aa 𝑎 at position 𝑝 is calculated as the proportion of CDR3 sequences in which 𝑎 appears at 𝑝 and can be defined as follow: Where 𝑛 is the total number of CDR3 sequences and 𝑁𝑎,𝑝 the number of sequences. Additionally, aa were classified into three hydropathy categories (neutral, hydrophobic, and hydrophilic, using the Kyte-Doolitle scale (25, 64).

Probability of generation analysis

The probability of generation analysis of DP and CD8 cell subtypes consisted initially in the creation of a generative model of V(D)J recombination using 400,000 randomly selected non-productive sequences from all samples for both TRA and TRB chains, with the IGOR tool (65). Generation probabilities (Pgen) for all sequences in the dataset were then calculated using the OLGA tool, based on the respective generative models for each cell subtype and TCR chain (66). Comparisons of Pgen values at the nucleotide level between males and females were performed. To ensure statistical robustness, control groups with equivalent male and female sample numbers were created by permuting the samples into two groups 20,000 times.

TCR network structure by sequence similarity

For each sample, 100 random sub-samplings of CDR3aa sequences were performed, using the minimum CDR3aa count per cell subtype. Two CDR3aa were connected if their Levenshtein distance was = 1 (Figure 1) (67). Network analysis focused on two metrics: the proportion of connected sequences and network density, defined as the ratio of actual connections to all possible connections. Median values across 100 subsampling iterations were used for analysis. Levenshtein distances were computed with the stringdist package (v0.9.10), networks were visualized using Cytoscape (v3.8.2), and network density was calculated with igraph package (v1.5.1).

TRB CDR3aa motif search

Two types of structural motifs were searched: local motifs (strict sequential motifs) and global motifs (sequences with substitutions preserving a positive BLOSUM62 score) (Figure 1).

Enriched motifs were selected using gliph2 function with turboGliph package (v0.99.2), with specific parameters defined as follow:

-Input sequences were from the sex group of interest (male or female), with reference sequences matched for TRBV gene usage

-CDR3aa sequences were required to be longer than eight aa

-Local motif lengths were restricted to 3-5 aa

-Only motifs constituted with strictly more than two unique CDR3aa are conserved

-No boost of local motif significance of Fisher test

Additional filters were applied so that: (i) a motif includes public CDR3aa sequences (shared by at least two individuals), (ii) a significant enrichment (Fisher’s test, p < 0.01) and (iii) a usage difference between groups of at least twofold (Wilcoxon test, p < 0.05).

Motif validation was performed on an publicly published pediatric thymic dataset from Arstila team (35, 36) consisting in a bulk TCR dataset from infants aged from seven days to eight months. Here, we removed one boy named “Thymus A” corresponding to a twin of the “Thymus B” and the “Thymus 1” neonate sample so that we analyze height samples with a male-to-female ratio of 5:3. Additional validation was carried out on sorted peripheral blood TCR repertoires (CD8, CD4 Teff and CD4 Treg) from seventy-five healthy volunteers aged from eighteen to eighty-four years (male-to-female ratio of 1.03:1), whose libraries were generated under the same conditions as those of our thymic dataset. They were recruited within the Transimmunom observational trial (NCT02466217) and HEALTHIL-2 (NCT03837093) (68). PBMCs from healthy individuals were isolated using Ficoll density gradient centrifugation and enriched for CD4+ T cells via EasySep™ Human CD4+ T Cell magnetic beads (Stemcell). Further separation into effector T cells (CD4+ CD25-) and Tregs (CD4+ CD25+) was performed using EasySep™ Human Pan-CD25 magnetic beads (Stemcell). Purity was assessed by flow cytometry based on the expression of CD4 and FoxP3, with a purity threshold of >80%. From the frozen PBMC aliquots, CD8+ T cells (CD3+ CD8+) were sorted using a FACS ARIA II cell sorter.

TRB CDR3aa specificity analysis

We established a curated and harmonized TCR specificity database by integrating data from three public TCR repositories: Mc-PAS, IEDB, and VDJdb (38, 39, 69) (Figure 1). Reliability levels for sequences and specificity assignments were added based on the identification methods used in the original studies, ensuring a robust dataset for specificity analysis (70). Specificity groups were categorized by antigen class, including viral, bacterial, yeast, parasitic and self-antigens linked to diseases (e.g. autoimmune conditions, cancer) or unrelated to diseases, as well as antigens from wheat, plants or animals.

This harmonized database was used to identify exact matches with the CDR3aa sequences of our TRB sequences. Enrichment of CDR3aa sequences associated with a given specificity group in male and female thymic repertoires for each cell subtype compared to the specificity distribution in the database was tested. Specificity group distributions between male and female samples were also performed, as was the overall usage of these sequences across the entire TCR repertoire.

Additionally, we focused on polyspecific sequences defined as CDR3aa TRB sequences recognizing at least two different species in case of microorganisms, or belonging to multiple specificity groups for other antigens. The same analyses as described above were conducted for these polyspecific sequences.

Statistical analysis

For V-J gene usage, CDR3aa length usage, CDR3aa usage, network similarity and specificity analyses, the comparison between males and females were statistically analysed using the Wilcoxon test. The statistical tests were performed using ggpubr library (v0.6.0) in R. Significance was denoted by the number of asterisks from one to four, each representing respectively p ≤ 0.05, p ≤ 0.01, p ≤ 0.001, and p ≤ 0.0001. The comparisons of the Rényi curves and Pgen were evaluated using the Kolmogorov-Smirnov test. The comparisons of the proportions of specific sequences in male and female relative to those in the specificity database were tested using the Wilcoxon test.

eLife Assessment

This study provides useful insights into addressing the question of whether the prevalence of autoimmune disease could be driven by sex differences in the T cell receptor (TCR) repertoire, correlating with higher rates of autoimmune disease in females. The authors compare male and female TCR repertoires using bulk RNA sequencing, from sorted thymocyte subpopulations in pediatric and adult human thymuses; however, the results do not provide sufficient analytical rigor and incompletely support the central claims.

Reviewer #1 (Public review):

Summary:

The goal of this paper was to determine whether the T cell receptor (TCR) repertoire differs between a male and a female human. To address this, this group sequenced TCRs from double-positive and single-positive thymocytes in male and female humans of various ages. Such an analysis on sorted thymocyte subsets has not been performed in the past. The only comparable dataset is a pediatric thymocyte dataset where total thymocytes were sorted.

They report on participant ages and sexes, but not on ethnicity, race, nor provide information about HLA typing of individuals. Though the experiments themselves are heroic, they do represent a relatively small sampling of diverse humans. They observed no differences in TCRbeta or TCRalpha usage, combinational diversity, or differences in the length of the CDR3 region, or amino acid usage in the CD3aa region between males or females. Though they observed some TCRbeta CD3aa sequence motifs that differed between males and females, these findings could not be replicated using an external dataset and therefore were not generalizable to the human population.

They also compared TCRbeta sequences against those identified in the past using computational approaches to recognize cancer-, bacterial-, viral-, or autoimmune-antigens. They found very little overlap of their sequences with these annotated sequences (depending on the individual, ranging from 0.82-3.58% of sequences). Within the sequences that were in overlap, they found that certain sequences against autoimmune or bacterial antigens were significantly over-represented in female versus male CD8 SP cells. Since no other comparable dataset is available, they could not conclude whether this is a finding that is generalizable to the human population.

Strengths:

This is a novel dataset. Overall, the methodologies appear to be sound. There was an attempt to replicate their findings in cases where an appropriate dataset was available. I agree that there are no gross differences in TCR diversity between males and females.

Weaknesses:

Overall, the sample size is small given that it is an outbred population. The cleaner experiment would have been to study the impact of sex in a number of inbred MHC I/II identical mouse strains or in humans with HLA-identical backgrounds.

It is unclear whether there was consensus between the three databases they used regarding the antigens recognized by the TCR sequences. Given the very low overlap between the TCR sequences identified in these databases and their dataset, and the lack of replication, they should tone down their excitement about the CD8 T cell sequences recognizing autoimmune and bacterial antigens being over-represented in females.

The dataset could be valuable to the community.

Reviewer #2 (Public review):

Summary:

This study addresses the hypothesis that the strikingly higher prevalence of autoimmune diseases in women could be the result of biased thymic generation or selection of TCR repertoires. The biological question is important, and the hypothesis is valuable. Although the topic is conceptually interesting and the dataset is rich, the study has a number of major issues that require substantial improvement. In several instances, the authors conclude that there are no sex-associated differences for specific parameters, yet inspection of the data suggests visible trends that are not properly quantified. The authors should either apply more appropriate statistical approaches to test these trends or provide stronger evidence that the observed differences are not significant. In other analyses, the authors report the differences between sexes based on a pulled analysis of TCR sequences from all the donors, which could result in differences driven by one or two single donors (e.g., having particular HLA variants) rather than reflect sex-related differences.

Strengths:

The key strength of this work is the newly generated dataset of TCR repertoires from sorted thymocyte subsets (DP and SP populations). This approach enables the authors to distinguish between biases in TCR generation (DP) and thymic selection (SP). Bulk TCR sequencing allows deeper repertoire coverage than single-cell approaches, which is valuable here, although the absence of TRA-TRB pairing and HLA context limits the interpretability of antigen specificity analyses. Importantly, this dataset represents a valuable community resource and should be openly deposited rather than being "available upon request."

Weaknesses:

Major:

(1) The authors state that there is "no clear separation in PCA for both TRA and TRB across all subsets." However, Figure 2 shows a visible separation for DP thymocytes (especially TRA, and to a lesser degree TRB) and also for TRA of Tregs. This apparent structure should be acknowledged and discussed rather than dismissed.

(2) Supplementary Figures 2-5 involve many comparisons, yet no correction for multiple testing appears to be applied. After appropriate correction, all the reported differences would likely lose significance. These analyses must be re-evaluated with proper multiple-testing correction, and apparent differences should be tested for reproducibility in an external dataset (for example, the pediatric thymus and peripheral blood repertoires later used for motif validation).

(3) Supplementary Figure 6 suggests that women consistently show higher Rényi entropies across all subsets. Although individual p-values are borderline, the consistent direction of change is notable. The authors should apply an integrated statistical test across subsets (for example, a mixed-effects model) to determine whether there is an overall significant trend toward higher diversity in females.

(4) Figures 4B and S8 clearly indicate enrichment of hydrophobic residues in female CDR3s for both TRA and TRB (excluding alanine, which is not strongly hydrophobic). Because CDR3 hydrophobicity has been linked to increased cross-reactivity and self-reactivity (see, e.g., Stadinski et al., Nat Immunol 2016), this observation is biologically meaningful and consistent with higher autoimmune susceptibility in females.

(5) The majority of "hundreds of sex-specific motifs" are probably donor-specific motifs confounded by HLA restriction. This interpretation is supported by the failure to validate motifs in external datasets (pediatric thymus, peripheral blood). The authors should restrict analysis to public motifs (shared across multiple donors) and report the number of donors contributing to each motif.

(6) When comparing TCRs to VDJdb or other databases, it is critical to consider HLA restriction. Only database matches corresponding to epitopes that can be presented by the donor's HLA should be counted. The authors must either perform HLA typing or explicitly discuss this limitation and how it affects their conclusions.

(7) Although the age distributions of male and female donors are similar, the key question is whether HLA alleles are similarly distributed. If women in the cohort happen to carry autoimmune-associated alleles more often, this alone could explain observed repertoire differences. HLA typing and HLA comparison between sexes are therefore essential.

(8) In some analyses (e.g., Figures 8C-D) data are shown per donor, while others (e.g., Fig. 8A-B) pool all sequences. This inconsistency is concerning. The apparent enrichment of autoimmune or bacterial specificities in females could be driven by one or two donors with particular HLAs. All analyses should display donor-level values, not pooled data.

(9) The reported enrichment of matches to certain specificities relative to the database composition is conceptually problematic. Because the reference database has an arbitrary distribution of epitopes, enrichment relative to it lacks biological meaning. HLA distribution in the studied patients and HLA restrictions of antigens in the database could be completely different, which could alone explain enrichment and depletions for particular specificities. Moreover, differences in Pgen distributions across epitopes can produce apparent enrichment artifacts. Exact matches typically correspond to high-Pgen "public" sequences; thus, the enrichment analysis may simply reflect variation in Pgen of specific TCRs (i.e., fraction of high-Pgen TCRs) across epitopes rather than true selection. Consequently, statements such as "We observed a significant enrichment of unique TRB CDR3aa sequences specific to self-antigens" should be removed.

(10) The overrepresentation of self-specific TCRs in females is the manuscript's most interesting finding, yet it is not described in detail. The authors should list the corresponding self-antigens, indicate which autoimmune diseases they relate to, and show per-donor distributions of these matches.

(11) The concept of polyspecificity is controversial. The authors should clearly explain how polyspecific TCRs were defined in this study and highlight that the experimental evidence supporting true polyspecificity is very limited (e.g., just a single TCR from Figure 5 from Quiniou et al.).

Minor:

(1) Clarify why the Pgen model was used only for DP and CD8 subsets and not for others.

(2) The Methods section should define what a "high sequence reliability score" is and describe precisely how the "harmonized" database was constructed.

(3) The statement "we generated 20,000 permuted mixed-sex groups" is unclear. It is not evident how this permutation corrects for individual variation or sex bias. A more appropriate approach would be to train the Pgen model separately for each individual's nonproductive sequences (if the number of sequences is large enough).

Author response:

The following is the authors’ response to the original reviews

eLife Assessment

This study provides useful insights into addressing the question of whether the prevalence of autoimmune disease could be driven by sex differences in the T cell receptor (TCR) repertoire, correlating with higher rates of autoimmune disease in females. The authors compare male and female TCR repertoires using bulk RNA sequencing, from sorted thymocyte subpopulations in pediatric and adult human thymuses; however, the results do not provide sufficient analytical rigor and incompletely support the central claims.

The statement in the editorial assessment that our study “does not provide sufficient analytical rigor” surprised us. TCR repertoire analysis is indeed a highly complex domain, both experimentally and computationally. We consider ourselves to be leading experts in this field and have invested a great deal of effort to ensure the rigor and reproducibility of every analytical step.

Specifically, our group has previously benchmarked and published validated methodologies for the following areas: (i) TCR repertoire generation (Barennes et al., Nat Biotechnol 2021), (ii) repertoire analysis (Six et al., Frontiers in Immunol, 2013; Chaara et al., Frontiers in Immunol, 2018; Ritvo et al., PNAS, 2018; Mhanna et al., Diabetes, 2021; Trück et al., eLife, 2021; Quiniou et al., eLife, 2023; Mhanna et al., Cell Rep Methods, 2024; Mhanna et al., Nat Rev Primers Methods, 2024), and (iii) the curation and quality control of public TCR databases (Jouannet et al., NAR Genomics and Bioinformatics 2025). The current study applies these optimized and peer-reviewed pipelines, along with additional internal quality controls that we have been implemented over the years, ensuring the highest possible analytical standards for TCR repertoire studies.

We therefore respectfully feel that the phrase “insufficient analytical rigor” does not accurately reflect the methodological robustness of our work. This perception is also in contrast to the comment made by one of the reviewers, who explicitly noted that “overall, the methodologies appear to be sound.”

We would therefore be grateful if, upon reviewing our detailed point-by-point responses, the editors could reconsider this statement and tone it down in the final editorial summary.

With regard to comment that our results “incompletely support the central claims”, we will leave it to the reader’s judgement. We believe that our work provides a robust and transparent basis for future research into TCR repertoire, autoimmunity, and women’s health.

Reviewer 1 (Public reviews):

Summary

The goal of this paper was to determine whether the T cell receptor (TCR) repertoire differs between a male and a female human. To address this, this group sequenced TCRs from doublepositive and single-positive thymocytes in male and female humans of various ages. Such an analysis on sorted thymocyte subsets has not been performed in the past. The only comparable dataset is a pediatric thymocyte dataset where total thymocytes were sorted.

They report on participant ages and sexes, but not on ethnicity, race, nor provide information about HLA typing of individuals. Though the experiments themselves are heroic, they do represent a relatively small sampling of diverse humans. They observed no differences in TCRbeta or TCRalpha usage, combinational diversity, or differences in the length of the CDR3 region, or amino acid usage in the CD3aa region between males or females. Though they observed some TCRbeta CD3aa sequence motifs that differed between males and females, these findings could not be replicated using an external dataset and therefore were not generalizable to the human population.

They also compared TCRbeta sequences against those identified in the past using computational approaches to recognize cancer-, bacterial-, viral-, or autoimmune-antigens. They found very little overlap of their sequences with these annotated sequences (depending on the individual, ranging from 0.82-3.58% of sequences). Within the sequences that were in overlap, they found that certain sequences against autoimmune or bacterial antigens were significantly over-represented in female versus male CD8 SP cells. Since no other comparable dataset is available, they could not conclude whether this is a finding that is generalizable to the human population.

Strengths:

This is a novel dataset. Overall, the methodologies appear to be sound. There was an attempt to replicate their findings in cases where an appropriate dataset was available. I agree that there are no gross differences in TCR diversity between males and females.

We appreciate the positive feedback from the reviewer regarding these points.

Weaknesses:

Overall, the sample size is small given that it is an outbred population. The cleaner experiment would have been to study the impact of sex in a number of inbred MHC I/II identical mouse strains or in humans with HLA-identical backgrounds.

We respectfully disagree with the reviewer’s statement. We firmly believe that the issue we are dealing with, namely sex-based differences in thymic TCR selection relevant to autoimmunity, should be investigated more thoroughly in the general human population than in inbred mouse models.

While inbred mouse strains, being MHC I/II identical, eliminate the complexity of MHC variation, this comes at the cost of biological relevance. Firstly, a discrepancy in TCR generation or selection may only become apparent under specific MHC contexts, which could easily be overlooked when studying a single inbred strain. Secondly, inbred strains frequently contain fixed genetic variants that may influence thymic selection or immune regulation. This has the potential to introduce confounding effects rather than reducing them and not solving the generalization issue.

We are in full agreement that an HLA-matched human cohort would reduce inter-individual variability. However, such sampling is impossible in practice, as our thymic tissues were obtained from deceased organ donors, a collection effort that was, as the reviewer rightly noted, “heroic”. Despite these inherent limitations, the patterns we observed were consistent across multiple analytical approaches, lending robustness to our findings.

We now explicitly acknowledge this limitation in the Discussion of the revised manuscript and explain why, despite this constraint, our study provides meaningful and biologically relevant insights into human TCR selection and sex-related immune differences.

It is unclear whether there was consensus between the three databases they used regarding the antigens recognized by the TCR sequences. Given the very low overlap between the TCR sequences identified in these databases and their dataset, and the lack of replication, they should tone down their excitement about the CD8 T cell sequences recognizing autoimmune and bacterial antigens being over-represented in females.

The three databases used in this study - McPAS-TCR, IEDB, and VDJdb - provide complementary and partially non-overlapping specificity landscapes. McPAS-TCR is enriched for pathology-associated TCRs, while IEDB and VDJdb contain a higher proportion of viral specificities. Combining them therefore broadens the antigenic spectrum accessible for analysis and represents the most comprehensive approach currently possible to capture the diversity of TCR–antigen annotations.

With regard to the limited overlap between our dataset and these databases, this observation should be interpreted with caution. While the overlap may appear minimal at first glance, it is a biologically significant phenomenon. The public databases collectively contain only a minute fraction of the total universe of TCR specificities, estimated to exceed 1015-21 possible receptors in humans. In this context, the observation of any overlap at all, particularly with coherent biological patterns such as the overrepresentation of autoimmune- and bacterialassociated TCRs in females, is noteworthy.

We have included a short clarification in the Discussion of the revised manuscript to make this point explicit and to further temper the language describing this finding.

The dataset could be valuable to the community.

We thank the reviewer for highlighting the potential value of this dataset to the community. It will be made publicly available on the NCBI website. We would like to clarify that our intention has always been to make this dataset publicly available; therefore, we take back any incorrect suggestions made in the original submission.

Reviewer #1 (Recommendations for the authors):

I would just recommend toning down the excitement about autoimmune TCRs being overrepresented in females. Then the conclusions will be in alignment with their results.

We thank the reviewer for this constructive recommendation. We would like to express our full support for the editorial transparency policies of eLife, which allow readers to access to both the reviewers’ comments and our detailed responses, enabling them to form their own informed opinions regarding our conclusions.

Nevertheless, we have moderated some of our wording.

Reviewer #2 (Public review):

Summary:

This study addresses the hypothesis that the strikingly higher prevalence of autoimmune diseases in women could be the result of biased thymic generation or selection of TCR repertoires. The biological question is important, and the hypothesis is valuable. Although the topic is conceptually interesting and the dataset is rich, the study has a number of major issues that require substantial improvement. In several instances, the authors conclude that there are no sex-associated differences for specific parameters, yet inspection of the data suggests visible trends that are not properly quantified. The authors should either apply more appropriate statistical approaches to test these trends or provide stronger evidence that the observed differences are not significant. In other analyses, the authors report the differences between sexes based on a pulled analysis of TCR sequences from all the donors, which could result in differences driven by one or two single donors (e.g., having particular HLA variants) rather than reflect sex-related differences.

Strengths:

The key strength of this work is the newly generated dataset of TCR repertoires from sorted thymocyte subsets (DP and SP populations). This approach enables the authors to distinguish between biases in TCR generation (DP) and thymic selection (SP). Bulk TCR sequencing allows deeper repertoire coverage than single-cell approaches, which is valuable here, although the absence of TRA-TRB pairing and HLA context limits the interpretability of antigen specificity analyses. Importantly, this dataset represents a valuable community resource and should be openly deposited rather than being "available upon request."

We thank the reviewer for highlighting the potential value of this dataset to the community. It will be made publicly available on the NCBI website. We would like to clarify that our intention has always been to make this dataset publicly available; therefore, we take back any incorrect suggestions made in the original submission.

Weaknesses:

Major:

The authors state that there is "no clear separation in PCA for both TRA and TRB across all subsets." However, Figure 2 shows a visible separation for DP thymocytes (especially TRA, and to a lesser degree TRB) and also for TRA of Tregs. This apparent structure should be acknowledged and discussed rather than dismissed.

We thank the reviewer for this careful observation. Discussing apparent “trends” rather than statistically significant results is indeed a nuanced issue, as over-interpretation of visual patterns is usually discouraged. We agree that, within the specific context of TCR repertoire analyses, visual structures in multivariate projections such as PCA can provide useful contextual information.

However, we have not identified a striking trend in our representation. We therefore chose to avoid overemphasizing these visual impressions in the text.

Supplementary Figures 2-5 involve many comparisons, yet no correction for multiple testing appears to be applied. After appropriate correction, all the reported differences would likely lose significance. These analyses must be re-evaluated with proper multiple-testing correction, and apparent differences should be tested for reproducibility in an external dataset (for example, the pediatric thymus and peripheral blood repertoires later used for motif validation).

As is standard in exploratory immunogenomic studies, including TCR repertoire analyses, our objective was to uncover broad biological patterns rather than to establish definitive statistical associations. In analyses that are discovery-oriented, correction for multiple testing, while essential in confirmatory contexts, is not mandatory and may even obscure meaningful trends by inflating type II error rates. Our objective was therefore to highlight consistent directional patterns across analytical layers, to guide future confirmatory work rather than to make categorical claims.

We also note that this comment somewhat contrasts with the earlier suggestion to discuss trends that are not statistically significant.

With regard to the proposal to verify our observations using an external dataset, we are in full agreement that independent confirmation would be beneficial. However, as reviewer 1 rightly emphasized, the generation of such datasets from sorted human thymocyte subsets is “heroic” and has rarely, if ever, been achieved. We are aware of no existing dataset that provides comparable material or analytical depth.

The available single-cell thymic dataset (Park et al., Science 2020) includes only a few hundred sequences per donor, which is significantly less than the number of sequences in our study. This limited dataset is not adequate for cross-validation or for representing the full complexity of thymic TCR repertoires.

As with the pediatric thymus dataset, the lack of statistical power in the dataset due to the small number of female subjects (only three) means that sex-related differences in V/J usage cannot be evaluated.

Finally, the peripheral blood dataset is not appropriate for validating thymic generation or selection processes, as it reflects post-thymic selection and antigen-driven remodeling, making it impossible to distinguish peripheral effects from thymic influences.

For these reasons, none of the currently available datasets provides a sufficiently clean or powerful framework to test the reproducibility of subtle sex-associated effects on thymic TCR repertoires. Nevertheless, we fully agree that confirmation in an independent and larger cohort will be an important next step to refine these exploratory findings and assess their generalizability to a broader human population.

Supplementary Figure 6 suggests that women consistently show higher Rényi entropies across all subsets. Although individual p-values are borderline, the consistent direction of change is notable. The authors should apply an integrated statistical test across subsets (for example, a mixed-effects model) to determine whether there is an overall significant trend toward higher diversity in females.

We agree that Rényi entropies tend to show a consistent direction of change across subsets, with slightly higher values observed in females. In this section, our objective was to provide a descriptive overview of diversity patterns for each thymic subset. This is because these subsets are biologically distinct and therefore require individual analysis, as we previously demonstrated using the same dataset (Isacchini et al, PRX Life. 2024). Therefore, while a mixed-effects approach could in principle be applied to test for an overall trend, such an analysis would rely on the assumption of a common sex effect across heterogeneous cell types.

It is important to note that the complete dataset has now been made publicly available, enabling interested researchers to perform additional integrative or model-based analyses to further explore these diversity trends.

Figures 4B and S8 clearly indicate enrichment of hydrophobic residues in female CDR3s for both TRA and TRB (excluding alanine, which is not strongly hydrophobic). Because CDR3 hydrophobicity has been linked to increased cross-reactivity and self-reactivity (see, e.g., Stadinski et al., Nat Immunol 2016), this observation is biologically meaningful and consistent with higher autoimmune susceptibility in females.

We thank the reviewer for this insightful comment.

As correctly noted, increased hydrophobicity at specific CDR3β positions has been linked to enhanced cross-reactivity and self-reactivity, as described by Stadinski et al. (Nat Immunol 2016), and we reference this work in the manuscript.

In our analysis corresponding to Figure 4B (TRB), hydrophobicity was quantified at the sequence level by computing, for each unique CDR3β sequence, the overall proportion of hydrophobic amino acids across the CDR3 loop. This approach aligns with that of Lagattuta et al. (Nat Immunol 2022), whose code we adapted to accommodate longer CDR3s. This global hydrophobicity metric captures overall composition, but, by its construction, does not account for positional context, the key mechanism implicated by Stadinski et al.

As outlined in our original Figure 4C, the results were obtained through a position-based amino acid analysis. For each CDR3β sequence, we extracted the amino acid at every IMGTdefined CDR3 position (p104–p118) and quantified, at each position, the percentage of unique sequences containing each amino acid. Positions p109 and p110 correspond to the p6–p7 sites highlighted by Stadinski et al. as functionally relevant for self-reactivity. This analysis evaluates positional composition independently of clonotype frequency, focusing specifically on hydrophobic amino acid classes.

Following the recommendation of the reviewer, the revised manuscript has removed alanine (which is only weakly hydrophobic) has been excluded from the hydrophobic residue set. With this refined definition, we observe a significant enrichment of hydrophobic amino acids at p109 in CD8 T cell repertoires from females, with similar but non-significant trends at p109 in DP and CD4 Teff cells and at p110 in CD8 cells (see new Figure 4C).

As outlined in the revised Methods, Results, and Discussion sections, Figure 4C focuses exclusively on positional hydrophobic amino acid usage. This was previously implicit, although it was noted in the legend and visually represented in the plots.

The majority of "hundreds of sex-specific motifs" are probably donor-specific motifs confounded by HLA restriction. This interpretation is supported by the failure to validate motifs in external datasets (pediatric thymus, peripheral blood). The authors should restrict analysis to public motifs (shared across multiple donors) and report the number of donors contributing to each motif.

We fully agree that donor-specific and HLA-restricted motifs represent a major potential confounder in repertoire-level comparisons. To minimize this potential bias, our analysis was explicitly restricted to public motifs, as clearly stated in the Materials and Methods section:

“Additional filters were applied so that: (i) a motif includes public CDR3aa sequences (shared by at least two individuals); (ii) a significant enrichment is detected (Fisher’s exact test, p < 0.01); and (iii) a usage difference between groups of at least twofold (Wilcoxon test, p < 0.05).”

Accordingly, every motif reported in the manuscript is supported by at least two independent donors, ensuring that no motif reflects an individual- or HLA-specific effect (see Supplementary Figures 10-13[previously Supplementary Figure 9]). We have now added a more explicit mention of the number of donors contributing to each motif in the figure legend and have clarified this point in the revised Methods and Results sections to make this criterion more visible to readers.

When comparing TCRs to VDJdb or other databases, it is critical to consider HLA restriction. Only database matches corresponding to epitopes that can be presented by the donor's HLA should be counted. The authors must either perform HLA typing or explicitly discuss this limitation and how it affects their conclusions.

We respectfully disagree with the assertion that HLA typing is necessary for the type of comparative analysis we have conducted. While it is true that HLA molecules present peptides to TCRs and thereby contribute to the tripartite interaction determining T cell activation, extensive evidence indicates that the CDR3 region, particularly CDR3β, is the dominant determinant of antigen specificity. This finding is supported by structural and computational studies (Madi et al., eLife, 2017; Huang et al., Nat. Biotech., 2020; MayerBlackwell et al., Methods Mol. Biol., 2022) showing that CDR3β residues are responsible for the majority of peptide contacts, whereas CDR1 and CDR2 primarily interact with the MHC framework.

As emphasized in several recent benchmarking studies (e.g., Springer et al., Front Immunol, 2021), CDR3β sequence composition alone captures most of the information required for specificity inference. Consequently, widely used and validated computational tools such as GIANA (Zhang et al. Nat. Commun. 2021), iSMART (Zhang et al. Clin. Cancer Res. 2020), and ATMTCR (Cai et al. Front. Immunol. 2022) rely exclusively on CDR3β aminoacid sequences and still achieve high predictive performance.

Our analysis aligns with this well-established paradigm. While we agree that integrating donor HLA typing would refine epitope-level annotation and reduce potential noise, the absence of HLA data does not invalidate the comparative framework we used, which focuses on relative representation of annotated specificities across groups rather than on individual TCR–HLA–peptide triads.

Although the age distributions of male and female donors are similar, the key question is whether HLA alleles are similarly distributed. If women in the cohort happen to carry autoimmuneassociated alleles more often, this alone could explain observed repertoire differences. HLA typing and HLA comparison between sexes are therefore essential.

To address the issue of any potential differences in HLA background, we examined the subset of adult donors for whom HLA typing information was available (HLA-A, HLA-B, HLADR, and HLA-DQB; n = 16). Within this subset, the distribution of HLA alleles was relatively balanced between males and females (as illustrated by the heatmap showing HLA class II expression patterns and HLA class I family grouping in Author response image 1). This analysis suggests that the sex-associated differences in the repertoire observed in our study are unlikely to be driven solely by unequal representation of autoimmune-associated HLA alleles.

We acknowledge, however, that complete HLA information was not available for all donors, which remains a limitation of the dataset.

In some analyses (e.g., Figures 8C-D) data are shown per donor, while others (e.g., Fig. 8A-B) pool all sequences. This inconsistency is concerning. The apparent enrichment of autoimmune or bacterial specificities in females could be driven by one or two donors with particular HLAs. All analyses should display donor-level values, not pooled data.

While Figures 8A–B present pooled data to summarize global trends, the corresponding donor-level analyses were provided in Supplementary Figures 15B and 16 (previously Supplementary Figures 11B and 12). In these, each individual is shown separately, with each point representing an individual. It is important to note that these donor-resolved plots do not reveal any sample-specific driver: the patterns observed in the pooled data remain consistent across donors, without any single individual accounting for the apparent enrichments. As outlined in the revised manuscript, readers now directed to the relevant supplementary figures for further clarification.

The reported enrichment of matches to certain specificities relative to the database composition is conceptually problematic. Because the reference database has an arbitrary distribution of epitopes, enrichment relative to it lacks biological meaning. HLA distribution in the studied patients and HLA restrictions of antigens in the database could be completely different, which could alone explain enrichment and depletions for particular specificities. Moreover, differences in Pgen distributions across epitopes can produce apparent enrichment artifacts. Exact matches typically correspond to high-Pgen "public" sequences; thus, the enrichment analysis may simply reflect variation in Pgen of specific TCRs (i.e., fraction of high-Pgen TCRs) across epitopes rather than true selection. Consequently, statements such as "We observed a significant enrichment of unique TRB CDR3aa sequences specific to self-antigens" should be removed.

We respectfully disagree with the conclusion that our enrichment analysis lacks biological meaning. Our approach directly involves a direct comparison of the same set of observed TCR sequences between males and females. Consequently, any potential biases related to generation probability (Pgen), which affect all sequences equally, cannot account for the observed sex-specific differences. To summarize, because the comparison is performed on the same set of sequences, changes in the probability of generation across epitopes cannot explain the differences seen between the sexes.

We do agree, however, that the composition of the reference databases may influence apparent enrichment patterns, as these resources contain uneven distributions of epitope categories and often incomplete information regarding HLA restriction. It should be noted that this limitation is inherent to all database-based annotation approaches, a fact which is explicitly acknowledged in the revised Discussion.

The overrepresentation of self-specific TCRs in females is the manuscript's most interesting finding, yet it is not described in detail. The authors should list the corresponding self-antigens, indicate which autoimmune diseases they relate to, and show per-donor distributions of these matches.

We thank the reviewer for this constructive suggestion.

As recommended, we have expanded the description of the self-specific TCRs identified in our dataset and now provide this information in Supplementary Table 2 of the revised manuscript. Specifically, the table lists the corresponding self-antigens and the autoimmune diseases with which they are associated. In our curated database, these annotations primarily correspond to celiac disease and type 1 diabetes, which were the two autoimmune contexts explicitly defined in the manually curated reference datasets.

For the “cancer” specificity group, we have clarified that antigen assignments were established based on (i) annotations available in the original databases (IEDB, VDJdb, McPAS-TCR) and (ii) cross-referencing with additional resources, including the Human Protein Atlas, the Cancer Antigenic Peptide Database (de Duve Institute), and the Cancer Antigen Atlas (Yi et al., iScience 2021), to ensure consistency in the classification of cancer and neoantigen specificities. Please refer to the Materials and Methods section for a full description of the procedure for this specific assignment.

Donor-level distributions of these self-specific matches are now shown in Supplementary Figures 15B and 16 (previously Supplemental Figures 11B and 12), allowing direct visualization of inter-donor variability. Importantly, these plots confirm that the observed enrichment in females is not driven by a single individual, further supporting the robustness of the finding.

The concept of poly-specificity is controversial. The authors should clearly explain how polyspecific TCRs were defined in this study and highlight that the experimental evidence supporting true polyspecificity is very limited (e.g., just a single TCR from Figure 5 from Quiniou et al.).

We certainly agree (and regret) that the concept of TCR polyspecificity remains a subject of debate and often underappreciated in the field of immunology. As Don Mason famously discussed in his seminal essay “A very high cross-reactivity is an essential feature of the TCR” (doi: 10.1016/S0167-5699(98)01299-7) published over 25 years ago, both theoretical and experimental evidence indicates that each TCR can, in principle, recognize millions of distinct peptides, albeit with variable avidity.

Although this principle is widely accepted, it is frequently overlooked in the field of experimental immunology. In this area, anything that deviates from strict monospecificity is often disregarded as noise.

In our own analyses of large-scale TCR repertoires, we have repeatedly observed that many CDR3 sequences are annotated with multiple specificities across different databases, often corresponding to peptides from unrelated organisms. As demonstrated in Quiniou et al. (eLife 2023), such polyreactive TCRs exhibit distinctive features, including biased physicochemical composition, and tend to be enriched in various biological contexts. In our preliminary study of such TCRs, which have the capacity to be specific for multiple viral- and self- epitopes, we hypothesized that they may serve as a first line of defense against pathogens and also be involved in triggering autoimmunity. We therefore consider it important to report this phenomenon rather than omit it, especially given its potential relevance to both protective immunity and autoimmunity.

In the present study, polyspecific TCRs were defined operationally as TRB CDR3aa sequences associated with a minimum of two distinct specificity groups, corresponding either to different microbial species or to multiple antigen categories within the curated database. Therefore, our definition captures broader antigenic groupings rather than epitope-level binding events.

We fully acknowledge that direct experimental evidence for true molecular-level polyspecificity remains limited. Indeed, as the reviewer notes, only a single TCR with multiepitope reactivity has been rigorously demonstrated to date (Quiniou et al.2023). Consequently, our analysis does not make claims about structural promiscuity; instead, it uses database-annotated cross-reactivity as a proxy to explore broader repertoire-level patterns.

As outlined in the Methods section, this definition has been clarified and its discussion expanded in the Discussion to explicitly address these conceptual and methodological nuances.

Minor:

Clarify why the Pgen model was used only for DP and CD8 subsets and not for others.

As noted, computing Pgen values involves two steps: (i) training a generative model of V(D)J recombination using IGoR, and (ii) estimating generation probabilities with OLGA based on that model. Both steps require a significant amount of computing power, especially when applied to large repertoires across multiple subsets. For this reason, we focused the analysis on DP thymocytes, which represent the repertoire prior to thymic selection, and CD8 T cells after CD8 selection.

The Methods section should define what a "high sequence reliability score" is and describe precisely how the "harmonized" database was constructed.

Briefly, the annotated database used in this study was constructed in accordance with the procedure established in our previously published work (Jouannet et al., NAR Genomics and Bioinformatics, 2025). The study integrates three publicly available resources, IEDB, VDJdb, and McPAS-TCR, which were collected as of October 2023. These three datasets were then merged into a single harmonized compendium, undergoing extensive standardization. When entries shared identical information across databases (same V–CDR3–J for both TRA and TRB, same epitope, organism, PubMed ID, and cell subset), only one representative was kept; discrepant or incomplete entries were retained to preserve information. We then assigned a sequence reliability score, the Verified Score (VS), following the verification strategy used by IEDB. The scale ranges from 0 to 2 and reflects the concordance between calculated and curated TRA/TRB CDR3 sequences (2 = both TRA and TRB present are verified, 1.1 = only TRA verified, 1.2 = only TRB verified, 0 = no verified chain). A second score, the Antigen Identification Score (AIS), is used to rank antigen-identification methods on a scale of 0 to 5, according to the strength of the experimental evidence supporting them.

In the present study, “high reliability” refers to sequences with a verified TRB CDR3aa chain (VS ≥ 1.2) and an AIS score corresponding to T cells in vitro stimulation with a pathogen, protein or peptide, or pMHC X-mer sorting (> 3.2, excluding categories 4.1 and 4.2), ensuring that downstream analyses were performed on a rigorously curated and biologically trustworthy dataset. The Methods section now explicitly details these criteria.

The statement "we generated 20,000 permuted mixed-sex groups" is unclear. It is not evident how this permutation corrects for individual variation or sex bias. A more appropriate approach would be to train the Pgen model separately for each individual's nonproductive sequences (if the number of sequences is large enough).

The objective of this analysis was to determine whether the enrichment of TRBV06-5 in females was due to random grouping of individuals or whether it was attributable to sex itself. To do so, we generated all possible perfectly mixed groups of donors (i.e., groups containing an equal number of male and female donors) for the concerned thymocyte subset, and then performed 20,000 random pairwise comparisons between such mixed groups. For each comparison, we tested the TRBV06-5 usage between the two mixed groups. This procedure directly evaluates whether group composition (independent of sex) could spuriously generate differences in TRBV usage. Notably, none of these 20,000 comparisons between the two mixed groups yielded a statistically significant difference in TRBV06-5 usage. In contrast, when comparing the true male and female groups, a significant difference was identified. This demonstrates that the signal we observe is not driven by random donor grouping or individual-level variation, but is specifically associated with sex. It is important to note that this analysis, which is designed to exclude spurious group effects, is rarely performed in published repertoire studies, yet it provides an important internal control for robustness.

Reviewer #2 (Recommendations for the authors):

(1) Data availability "upon request" is unacceptable. All raw and processed data, as well as scripts used for analysis and figure generation, must be publicly deposited before publication.

We would like to clarify that our intention has always been to make this dataset publicly available. It was a mistake to suggest otherwise in the original submission.

(2) At the beginning of the Results section, include a brief description of the dataset: number of donors, sex ratio, age range, number of samples per subset, and sorting strategy. Although Figure 1 shows this, the information should also be mentioned in the main text.

In line with the recommendation, we have now added a summary of the cohort characteristics at the beginning of the Results section. This includes the number of donors, sex ratio, age range, number of samples per subset, and the sorting strategy used. While this information was already included in Figure 1, we concur that including it directly in the main text enhances readability.

(3) Report the number of cells and unique clonotypes analyzed per individual. Rank-frequency plots (in log-log coordinates) would be helpful.

We have now added, for each donor and each subset, the number of cells, and additionally for each chain, the number of total and unique clonotypes analyzed. This information is provided in the revised manuscript in a new supplementary table (Supplemental Table 1).

These plots have been integrated into the revised manuscript as Supplementary Figure 2.

(4) For analysis in Figure 4B, the total fraction of hydrophobic amino acids should be calculated for each patient separately, and values for men and women should be compared (analogously to Figure 4C, but for the whole CDR3 and excluding alanine).

Please note that the TRB CDR3aa composition in Figure 4B has already been quantified at the individual level. For each unique TRB CDR3aa sequence, we computed the proportion of each of the 20 amino acids across the CDR3β loop, then summarized these values per donor (mean per individual). The log2 fold change displayed in Figure 4B (and supplemental Figure 9 for TRA) is calculated from the median donor-level values for females versus males, rather than from pooled CDR3s. It is intended as descriptive, “global” view of amino acid usage within the central CDR3 region. Hydrophobicity was not used directly in the computation, but is indicated only by bar color, based on the Kyte-Doolittle- derived IMGT classification. This provides an observational overview of amino acid composition in the central CDR3 region.

As the mechanistic link between hydrophobicity and self-reactivity described by Stadinski et al. is explicitly position-dependent, we consider positional analyses to be the most appropriate method for formally interrogating this hypothesis, as we did in Figure 4C. Here, our primary focus was on the position-specific usage of hydrophobic amino acids at IMGT positions p109-p110. These positions correspond to the central p6-p7 positions described by Stadinski et al. For each individual, we computed the proportion of unique TRB CDR3aa sequences carrying a hydrophobic amino acid at a given position.

Accordingly, in the revised manuscript we refined the Figure 4C by excluding alanine due to its weak hydrophobic property (as recommended by the reviewer) This positional composition analysis now reveals a statistically significant increase in hydrophobic usage at p109 in female CD8 repertoires, with similar, though non-significant, trends at p109 in DP and CD4Teff ad at p110 in CD8 cells. Figure 4B is therefore retained as an exploratory overview of amino acid composition usage along the CDR3 loop, while Figure 4C is used for the more specific question of hydrophobicity and potential cross-reactivity.

The Methods section has been expanded to provide clearer descriptions of these computations, and the Results and Discussion sections corresponding to Figures 4B-C (and supplemental Figure 9) have been revised to make the rationale, implementation, and interpretation of these hydrophobicity analyses more explicit.

(5) Figure 6 shows a trend toward higher clustering of Treg TCRs in males, which could relate to the lower incidence of autoimmunity in men. The authors could test whether specific Treg clusters are male-specific and shared among male donors.

As shown in Figure 6, a clear trend towards higher similarity among Treg CDR3aa sequences in males is evident, as indicated by the proportion of sequences included in clusters and in the overall similarity density. However, identifying “male-specific clusters” shared across donors is not straightforward in our analytical framework.

In our approach, for each cell subset, CDR3aa sequences were downsampled 100 times to the smallest sample size, and clustering was repeated at each iteration. Therefore, the clusters’ identities are not consistent across iterations. The clusters depend on the specific subset of sequences selected at each downsampling step, as well as on their underlying Pgen distribution. Therefore, it is not possible to reliably assess whether specific clusters are systematically “male-shared”. This is because cluster composition is a function of stochastic resampling rather than of biological structure. For this reason, a comparison of cluster identities across donors would not produce interpretable results.

Version 2

Abstract

Women represent about 80% of patients with autoimmune diseases. This may partly result from sex-based differences in T cell receptor (TCR) selection during thymocyte development, potentially influenced by hormones and the lower expression of the Autoimmune Regulator (AIRE) transcription factor in females.

To investigate this, we analyzed sex-specific differences in TCR generation and selection. We examined TCR repertoires in double-positive thymocytes and single-positive thymic cells, including CD8⁺ and CD4⁺ effector T cells and regulatory T cells (Tregs), derived from male and female organ donors. Minimal sex-based differences were observed in V and J gene usage, and there were no notable differences in TCR repertoire diversity, complementarity-determining region 3 (CDR3) length, amino acid composition, or network structure. No TCR sequences were exclusive to either sex.

However, female effector T cells exhibited a significantly higher prevalence of TCRs specific to self-antigens implicated in autoimmunity compared to males, while female Tregs showed a reduced frequency of such TCRs. These differences were not observed for TCRs targeting self-antigens unrelated to autoimmunity or antigens associated with cancer or viruses.

Our findings identify a sex-specific imbalance in thymic selection of TCRs with autoimmunity-associated specificities, providing mechanistic insight into the increased susceptibility of women to autoimmune diseases.

Introduction

Several studies have highlighted a sex imbalance in diseases involving the immune system. This is particularly evident in the case of autoimmune diseases (AIDs), where 80% of those affected individuals are women, and the severity of the condition varies between the sexes (1–3). In addition, men are more severely affected by infectious diseases such as COVID-19 and tuberculosis (4, 5). The response to preventive immunotherapies such as vaccination, and to curative treatments, like immune checkpoint inhibitors or anti-TNF alpha, is also sex dependent (6). These observations indicate the presence of intrinsic biological differences in immune responses between males and females.

A variety of factors have been identified that could help explain the observed sex-based differences. Among these, sex hormones have been shown to influence immune responses by acting on immune cells that express specific receptors to these hormones (7–9). Sex hormones could directly influence the adaptive immune response at the level of thymocyte differentiation and selection by acting on the expression of Autoimmune Regulator (AIRE) in thymic epithelial cells. AIRE is responsible for the thymic expression of otherwise tissue-specific antigens, contributing to both the negative selection of effector cells and the positive selection of Tregs that recognize such antigens (10). Testosterone increases AIRE expression in medullary thymic epithelial cells, whereas estradiol decreases it (11, 12). Furthermore, a recent study has shown that the transcriptomic profiles of these specialized antigen-presenting cells in the thymus differ between males and females (13). Thissuggests that there may be differences in the selection of the T cell receptor (TCR) repertoire between males and females.

Despite these findings, there is still limited knowledge about potential differences in the TCR repertoire between the sexes. The generation of the TCR is a key process in T-cell development that takes place in the thymus. The TCR consists of two chains, alpha (TRA) and beta (TRB), each of which resulting from a random genetic rearrangement involving the V (variable), D (diversity) [for the TRB], and J (joining) genes. The combination of these gene segments is accompanied by insertions and deletions, creating a highly diverse region, the Complementarity Determining Region 3 (CDR3) (14). It is this region of the TCR that predominantly interacts with the peptide, a critical interaction for antigen recognition and subsequent T cell activation. The ability to specifically recognize diverse antigens via the CDR3 is central to the efficiency and specificity of the adaptive immune response.

TCRs are randomly generated, which means that while some may be useful, others may be harmful by recognizing self-antigens and causing autoimmunity. The current understanding is that during thymocyte development, selection is based on the avidity of their TCRs for antigens presented by cortical and medullary thymic epithelial cells. Thymocytes expressing TCRs with insufficient or excessive avidity are eliminated (negative selection), while those with an intermediate avidity are positively selected and mature into single-positive cells, either CD4+ or CD8+. Thymocytes with the highest of these intermediate affinities develop in Tregs. Given AIRE’s pivotal function in these processes, sex differences in AIRE expression may significantly influence thymic selection.

The TCR repertoire of female and male peripheral T cells have shown that their diversity, particularly that of the TRB repertoire, is influenced not only by the age of the individuals but also by their sex (15–18). However, another study showed no association between the sex of individuals and the diversity of their TCR repertoire (19). Nevertheless, the study of peripheral blood TCRs reflects that of an immune system that has undergone through multiple immune responses. For example, observed differencesbetween males and females in their responses to self-antigens associated with autoimmunity could indicate either the cause (the TCRs drive the disease) or the consequence (the TCRs are expanded because of the disease).

To assess whether there are significant differences in the generation and thymic selection of the TCR repertoire between males and females, we therefore focused on analyzing the TCR repertoires of thymocytes. These cells represent the latest stage in T cell development and have not yet been involved in immune responses. Their repertoires thus represent the mechanisms of TCR generation for the analysis of CD4/CD8 double positive (DP) cells, and of TCR selection for CD4 and CD8 simple positive (SP) T cells. We generated a unique data set comprising the sequences of the TRA and TRB chains of DP cells, CD8 and CD4 SP effector cells, and CD4 SP Tregs from organ donors, both infants and adults. The analysis of TRA and TRB repertoires using state-of-the-art strategies revealed no sex differences in repertoire generation (DP cell repertoire), but an enrichment in females for CD8 SP T cells bearing CDR3 associated with specificities related to autoimmunity.

Results

We analyzed thymic TCRαβ repertoires from 22 human organ donors (11 females, 11 males, aged between 73 days and 64 years) (Figure 1, Supplementary Figure 1). For each donor, CD3+CD4+CD8+ double positive (DP), CD3+CD4-CD8+ CD8 single positive (SP), and CD3+CD4+CD8- CD4 SP thymocytes were sorted by flow cytometry. In 16 donors, CD4 SP cells were further subdivided into CD25- conventional CD4 Teff and CD25+ CD4 Treg subsets, whereas in the remaining six donors, total CD4 SP cells were analyzed as Teff-enriched populations. This resulted in 20 DP, 21 CD8 SP, 22 CD4 Teff SP and 14 CD4 Treg SP samples (Figure 1, Supplementary Figure 1). Bulk TCR sequencing was then performed on each cell subset, with the aim to characterizing the TRA and TRB repertoires. Rank–frequency plots of clonotype abundances displayed comparable heavy-tailed distributions across donors, without any obvious systematic differences between males and females (Supplementary Figure 2).

Schematic overview of the generation of the thymic TCR dataset and the analytical pipeline.

Top Panel: Generation of the thymic TCR dataset. From deceased human thymuses of males and females, we isolated key T-cell subtypes through cell sorting. These subtypes included double-positive (DP) cells (CD3+ CD4+ CD8+), single-positive (SP) CD8+ cells (CD3+ CD4- CD8+), SP CD4+ cells (CD3+ CD4+ CD8-) that were further separated into T effector (Teff, CD3+ CD4+ CD8- CD25-) and Treg (CD3+ CD4+ CD8- CD25+) cells. TCR libraries were generated from the RNA of each cell population using rapid amplification of cDNA ends by PCR (5’RACE PCR). Following sequencing, data preprocessing involved quality sequencing checks, contig alignment, and quality control. The final dataset comprised 20 DP samples (male-to-female ratio of 1:1), 21 SP CD8+ samples (1.1:1), 6 SP CD4+ samples (1:5), 16 SP CD4 Teff samples (1.67:1), and 14 SP CD4 Treg samples (1.33:1). Males are depicted in violet and females in orange. Bottom Panel: Analytical pipeline. We compared the TCR repertoires of males and females across various dimensions. We evaluated general aspects of the TCR repertoire were evaluated, including diversity, gene usage, CDR3aa length distribution, and aa usage within the CDR3 region. Additionally, we analyzed the probability of sequence generation and the TCR repertoire structure based on CDR3aa sequence similarity. We identified differentially expressed TRB CDR3aa motifs between sexes and analyzed TRB CDR3aa sequence specificity. Created with BioRender.com

Minimal sex-based differences in TCR V and J gene usage across thymic cell subsets

The first level of diversity in generating TCRs is the usage of V and J genes (20). We analyzed the usage of TRA and TRB V and J genes and their combinations in each thymocyte population (Figure 1).

Males and females had a relatively similar V gene usage, as evidenced by the lack of a clear separation in the Principal Component Analysis (PCA), for both TRA and TRB, and across all cell subtypes studied (Figure 2A). However, we observed some individual differences in V and J gene usage between males and females, with some being specific to certain cell subtypes (Supplemental Figures 3-6), and others observed across multiple cell subtypes (Supplemental Figures 3-6). For example, TRBV6-5 was found to be more highly expressed in females compared to males in DP cells (Supplemental Figure 3C) and CD8 SP cells (Supplemental Figure 4C).

Comparable overall TCR gene usage between males and females.

(A) Principal Component Analysis (PCA) derived from the distribution of TRAV (left) and TRBV (right) gene usage frequencies across sex groups (males vs females), showing results for DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and CD4 Treg (N = 14) cells (displayed from top to bottom). Each point on the graph represents an individual. Ellipses indicate 95% confidence intervals. (B) Heatmap showing the Jensen-Shannon Divergence (JSD) score between samples, derived from the distributional usage of TRAV-TRAJ (left) and TRBV-TRBJ (right) gene associations in DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and CD4 Treg (N = 14) cells (displayed from top to bottom). Hierarchical clustering was performed using the Euclidean distance and the complete linkage method. Males are shown in violet; females in orange.

In the following step, we compared the usage of TRAV-TRAJ and TRBV-TRBJ gene combinations across all individuals using the Jensen-Shannon Divergence (JSD) score (Figure 2B). The hierarchical clustering showed no clear separation between male and female individuals for any of the chain or cell subtype, indicating no significant differences in the gene combination usage between males and females (Figure 2B). A similar observation was made when directly analyzing the frequency of these V-J gene combinations (Supplemental Figures 3E-6E, 3F-6F).

Comparable TCR repertoire diversity between males and females

A high diversity of the TCR repertoire is essential for generating a broad potential reactivity against a variety of antigens. We thus compared the TCR repertoire diversity between males and females. To achieve this objective, we analyzed Rényi diversity curves (from zero to infinite 𝛼 parameters) that measure various aspects of TCR repertoire diversity. At different 𝛼 values, the Rényi index places more or less weight on the contribution of the most frequent clonotypes. Higher 𝛼 values highlight the most prevalent clonotypes, thereby revealing the effect of expanded clones on overall diversity.

Rényi curves showed comparable diversities between sexes in DP and CD4 Teff cells, for both TRA and TRB, and a difference in CD8 and CD4 Treg cells (Supplemental Figure 7). In CD8 cells, the curves begin to diverge at low 𝛼 values (Supplemental Figure 7), with an earlier inflection in males, indicating a less balanced distribution of TCR clones. This suggests that in males, fewer clones dominate the repertoire at lower diversity thresholds compared to females. In Tregs, while the inflection point is similar, the difference seems to be characterized by higher diversity and richness in females compared to males (Supplemental Figure 7). However, no significant differences were observed for the Shannon diversity index (𝛼 = 1), the Simpson index (𝛼 = 2), and the Berger-Parker index (𝛼 = infinite), for both chains and all cell subtypes (Figure 3B). These results indicate that the diversity of thymic TCR repertoire is similar between males and females.

Comparable thymic TCR repertoire diversity between males and females.

Boxplots display Shannon (A), Simpson (B), and Berger-Parker (C) index values for TRA (left) and TRB (right), across thymic T cell subtypes, displayed from top to bottom, in DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and CD4 Treg (N = 14). Each point on the graph represents the median value from 50 rarefactions per sample. Statistical analysis (Wilcoxon test) showed no significant sex bias in TCR repertoire diversity (p > 0.05). Males are shown in violet; females in orange.

Comparable CDR3aa length distribution in male and female TCR repertoires

Most of the repertoire diversity is generated by the random addition of nucleotides within the VDJ junctions, resulting in the hypervariable CDR3 region that interacts with the peptide-MHC complex, and is mostly responsible for the TCR specificity. The CDR3 length has been shown to directly influence the TCR’s ability to interact with the peptide (21). We thus studied the distribution of CDR3 aa (CDR3aa) sequence lengths between males and females to identify potential TCR generation and/or selection biases (Figure 1).

Our results showed that males and females have a comparable distribution of CDR3aa lengths for both TRA (Supplemental Figure 8) and TRB (Figure 4A). CDR3aa sequences in TRA were shorter than those in TRB, as previously described in peripheral blood T cells (22, 23). Although there were subtle differences in certain CDR3aa lengths between males and females, these variations were minimal, involving sequences that represent less than 2% of the total TCR repertoire (Figure 4A and Supplemental Figure 8). Altogether, the overall distribution of CDR3aa length is comparable between males and females.

CDR3aa length and amino acid composition of TRB CDR3s in males and females.

(A) Distribution of TRB CDR3aa length usage in DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and Treg (N = 14) SP cells (displayed from top to bottom). Stars indicate statistical differences between males and females based on the p-value of the Wilcoxon test (*: p < 0.05, **: p < 0.01). (B) The data on amino acid (aa) usage between males and females is presented as the log2 fold change of the median per-donor usage in females over males for each aa in the p108 to p114 CDR3aa region for TRB. A line at log2 fold change = 0 is indicative of the direction of the difference in usage frequency. The bars are colour-coded to the hydropathy class of the aa as defined by the Kyte-Doolittle-based IMGT classification (neutral aa by gray, hydrophilic aa by blue-green and hydrophobic aa by gold). Stars indicate statistical differences of usage between males and females based on the p-value of the Wilcoxon test (*: p < 0.05, **: p < 0.01). (C) Position-specific usage of hydrophobic aa (excluding alanine, due to its weak hydrophobicity) at IMGT positions p109 and p110 in TRB across thymic T cell subtypes, including DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and CD4 Treg (N = 14). For each donor, the values represent the proportion of unique TRB CDR3aa sequences carrying a hydrophobic amino acid at the indicated position. Stars indicate significant sex differences based on the p-value of the Wilcoxon test (*: p < 0.05), with a significant increase in hydrophobic usage at p109 in female CD8 SP cells. Males are depicted in violet; females in orange.

Subtle differences in the composition of CDR3aa in the thymic TCR repertoires of males and females

We analyzed the aa usage within the FG loop of the CDR3, which is in contact with the peptide-MHC complex (24). The classification of aa was determined based on their hydropathy properties, categorized as follows: hydrophilic, hydrophobic, or neutral (25). We observed significant differences in the usage of some aa between males and females for both TRA (Supplemental Figure 9) and TRB (Figure 4B). In DP cells, these differences preferentially involved hydrophobic aa: females exhibited higher alanine (A) usage in TRA and lower in TRB compared to males (Figure 4B and Supplemental Figure 9). Additionally, females showed increased usage of phenylalanine (F) and isoleucine (I) in TRB (Figure 4B). Furthermore, other aa with varying hydropathy properties were differentially used between males and females in CD4 Teff and Treg populations (Figure 4B and Supplemental Figure 9).

We then examined the usage of aa specifically at positions p109 and p110 of the TRB CDR3aa region, as hydrophobic aa at these two positions have been associated with a stronger recognition of self-antigens (26–28). For each individual, we calculated the proportion of unique TRB CDR3aa carrying a hydrophobic aa (excluding alanine, which is only weakly hydrophobic) at these two positions. This positional analysis revealed a significant increase in the usage of hydrophobic amino acid at position p109 in female CD8 cells compared with males, with similar trends observed at p109 in DP and CD4 Teff subsets and at p110 in CD8 cells (Figure 4C). Taken together, these positional patterns suggest a subtle sex-biased enrichment of hydrophobic amino acids at key TRB CDR3 positions that could be compatible with a slightly increased potential for self-reactivity and cross-reactivity in female CD8 T cells (28).

Slight sex-based variations in thymic TCR generation probability distribution

Another approach to evaluating biases in the generation and selection of the TCR repertoire is to compare the probability of generation (Pgen) of CDR3 nucleotide sequences (Figure 1). Pgen represents the likelihood of a sequence being produced, according to models of V(D)J recombination (14, 29). For DP and CD8 cells only and for each TCR chain, a sequence generation model was created using 100,000 non-productive nucleotide sequences. Using this model, the Pgen of all productive TCR sequences composing each repertoire was calculated.

When comparing the distribution of Pgen between males and females, we observe a very slight difference of distribution, characterized by a nearly null Kolmogorov-Smirnov score (DK-S < 0.05) (Figure 5). This difference is more pronounced in the TRB compared to the TRA chain (Figure 5). As this minor discrepancy might be indicative of variations among individuals irrespective of their sex, we have generated 20,000 permuted mixed-sex groups utilizing our dataset. Our findings revealed a significant distribution difference of Pgen sequences between these mixed groups population (DP – TRA: DK-S = 0.0183 ± 0.0145 (μ ± sd), DP – TRB: DK-S = 0.0190 ± 0.0161, CD8 – TRA: DK-S = 0.0227 ± 0.0153, and CD8 – TRB: DK-S = 0.0259 ± 0.0140). However, the observed difference in the Pgen distribution between males and females is higher than that between the mixed groups in the DP TRB.

Probabilities of generation of TCRs in males and females.

The figure shows the distribution of Pgen (probability of generation) sequences between males and females for TRA and TRB in DP and CD8 cells. A V(D)J recombination model was created using 400,000 non-productive random sequences derived from the nonproductive sequences of all individuals for each chain, both for DP and CD8 cells (65). This model was then used to calculate the Pgen of sequences for each individual (66). The overall distribution comparison between males and females was tested using the Kolmogorov-Smirnov test, with the D value and associated p-value indicated. Males are depicted in violet and females in orange.

Similar network structure of thymic TCR repertoires in males and females

Next, we evaluated whether thymic selection in males and females is directed towards similar or distinct sequences. For each TCR repertoire, CDR3aa sequences are connected if they differ by one aa, i.e. have a Levenshtein distance of =1 (Figure 1). Such linked sequences have a high probability of sharing the same specificities (30). A random downsampling to the smallest sample was performed to compare samples of similar sizes. This process was repeated 100 times, and the results are detailed in Supplemental Table 1. The shapes of the network are depicted in Figure 6A.

Comparable TCR repertoire network structure based on CDR3 amino acid sequence similarity between males and females.

For each sample, 100 random subsamplings were performed on the minimum number of CDR3aa per cell subtype. Two CDR3aa are linked if they have a Levenshtein distance of one. (A) TRA Network structure of subsampled TCR repertoire of a male subject for DP, CD8, CD4 Teff and CD4 Treg SP (from left to right). Each point on the graph represents a CDR3aa. (B-C) Comparison of the proportion of linked sequences (B) and network density (C) between male and female samples, in DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and CD4 Treg (N = 14). Each point on the graph represents the median value from 100 subsampling iterations for each sample. Statistical analysis using the Wilcoxon test revealed no significant sex differences for these two metrics (p > 0.05). Males are depicted in violet; females in orange.

To analyze these clusterings with greater precision, we used various metrics. When comparing the number of clustered sequences in each individual network between males and females, no significant differences were observed for any cell type in either chain (Figure 6B). Similarly, the degree of sequence similarity, as indicated by the density of edges between sequences within an individual’s TCR repertoire, was comparable between males and females (Figure 6C). These findings indicate that there are no significant sex-specific differences in the similarity of CDR3aa sequences, suggesting that males and females produce CDR3aa sequences with the same degree of variability within their TCR repertoires.

Absence of sex-specific CDR3aa motifs in thymic TCR repertoires of males and females

We then sought to determine whether we could detect CDR3aa sequence motifs that were more prevalent in one group compared to the other, in order to evaluate whether the selection of CDR3aa was biased towards particular sequences differently between males and females (Figure 1). To identify sequences with shared characteristics that might indicate a probable common specificity among individuals of the same group, the analysis focused exclusively on CDR3aa sequences of the TRB, as this region has been shown to have a greater influence on TCR specificity (31, 32). Local motifs were defined using Gliph2 (33), which are a sequences of three to five aa, as well as global motifs, which are sequences of more than three aa where, at one position, an aa can be substituted by another aa if it has a positive BLOSUM62 score (34).

We identified several hundred motifs that were differentially expressed between males and females for each cell subpopulation. A greater number of motifs were specific to females, ranging from 426 motifs for DP to 278 motifs for CD4 Tregs, compared to males, ranging from 328 motifs for CD4 Teffs to 34 motifs for CD4 Tregs (Figure 7A). The identified motifs are mainly local motifs and largely confined to their respective sex group, indicating that motifs identified in females are predominantly found in females, and vice versa (Supplemental Figures 10-13).

Thymic TRB TCR sex-associated motifs.

Different structural motifs found differentially expressed between males and females in our dataset. We distinguish local motifs as distinct aa sequences and global motifs as motif regions with one variable aa position maintaining a BLOSUM62 score ≥ 0. (A) Number of male and female associated motifs by cell subset. (B) Euler diagram illustrating the distribution and overlap between all sex-associated motifs. The numbers indicate the number of motifs in overlap zones. (C-D) Validation of these sex-associated TRB CDR3aa motifs. The heatmap illustrates the usage of all the TRB CDR3aa motif in the external thymic pediatric dataset (35, 36) (C) and those of TRB CD8 in the peripheral dataset (D). Sex and total CDR3aa number are depicted by sample. Males are depicted in violet and females in orange then local motifs in blue-green and global motifs in magenta.

As illustrated in Figure 7B, we observed minimal motif overlap between different cell subtypes. Only two female associated motifs overlapped: one between CD4 Teffs and CD4 Tregs, and another between CD4 Teffs and CD8 cells (Figure 7B). In contrast, no overlap was found between male-associated motifs (Figure 7B). Furthermore, it was observed that there was an overlap between one motif associated with female CD8 cells and a motif associated with male CD4 Teff cells (Figure 7B).

We then proceeded to evaluate the usage of the differentially expressed motifs in external datasets. There is no publicly available thymic TCR dataset that reports repertoires according to cell subsets. We thus tested these motifs on TCRs from pediatric bulk thymocytes, which contains TCR data from children aged between two and eight months, with a male-to-female ratio of 5:3 (35, 36). We calculated the usage of these differentially expressed motifs in each sample of this dataset. We were unable to separate individuals by sex using these motif usages (Figure 7C).

We then evaluated these motifs using a peripheral blood TCR dataset from healthy individuals, where CD8, CD4 Teff, and CD4 Treg cells had been sorted. The usage of these motifs could not distinguish males from females across all cell subsets (Figure 7D and Supplemental Figure 14).

Sex-biased enrichment of TCRs associated with autoimmune diseases and bacterial antigens

We sought to identify the usage of TCRs with known specificity as represented in the IEDB, Mc-PAS and VDJdb public databases (37–39). We compiled the TCRs from these three databases, retaining only those with high sequence reliability scores (see Materials and Methods). We focused on TRB sequences exclusively, due to their greater representation in the databases (31, 40). We identified 55,368 unique TRB CDR3aa sequences with high reliability and specificity assignment scores for at least one specificity.

We classified the TCR specificities based on the antigen they recognize: viral or bacterial peptides, or human peptides overexpressed in cancers, autoimmune diseases (AIDs), or neither of these diseases (Supplemental Figure 15A). We sought to identify the TRB CDR3aa sequences in our thymocyte dataset. Depending on the individuals and cell subtypes, they represented between 0.82% and 3.58% of the TRB repertoires.

We then compared the distribution of these CDR3aa between males and females for each cell subtype. We observed a significant enrichment of unique TRB CDR3aa sequences specific to self-antigens not associated with pathologies and those associated with cancers in our thymic TCR dataset, for both females and males (Supplemental Figure 15B). This enrichment is observed across all cell subtypes and is approximately 1.25 to 1.5 times higher than the baseline distribution of specificities in the pooled database (Supplemental Figure 15B). Strikingly, in the TCR repertoires with one specificity only, we observed a significantly higher proportion of unique TRB CDR3aa sequences associated with AIDs in female CD8 SP cells compared to male CD8 SP cells (Figure 8A). Furthermore, in the specific TCR repertoires only, there was a trend towards a higher proportion of thymic sequences with specificity associated to bacterial compounds in female CD8 SP cells compared to males, and an opposite trend was observed for CD4 Tregs SP (Figure 8A). These differences observed in CD8 SP cells persisted when examining the usage of these sequences (i.e. cumulative usage frequency), with significantly higher usage of sequences with specificity associated to bacterial compounds and those associated with AIDs in females compared to males (Figure 8B). Of particular note are the donor-resolved analyses (Supplementary Figures 15B and 16), in which each individual is represented separately. These confirm that these patterns are not driven by a single donor. Donor age had no significant effect on the sex-biased enrichment of self- and bacteria-specific TRB CDR3 sequences in CD8 SP and CD4 Treg SP cells (p ≥ 0.52), indicating that the observed differences are age-independent (Supplemental Figure 17). In addition, we also investigated for the presence of polyspecific TCRs, which are capable of recognizing multiple antigens from different organisms (41–43). These sequences were found to be enriched across all cell subtypes, in comparison to their representation in the reference database (Supplemental Figure 15C). However, no significant differences were observed in the proportion or usage of these polyspecific sequences between males and females. Interestingly, an inverse usage trend was noted between CD8 and CD4 Treg SP cells: females exhibited higher usage of these sequences in the CD8 SP repertoire but a lower usage in the CD4 Treg SP repertoire compared to males (Figure 8C-8D).

Sex-biased enrichment of TCRs specific for known antigens.

From a pooled and curated database, an exact match with this database infers the specificity of TRB CDR3aa of our thymic dataset. Many specificity groups are defined according to the nature of the antigen peptide targeted (bacteria, virus, autoimmune disease [AID], cancer and self-peptide no associated to disease [human]). This analysis compares the distributions of the proportion of unique TRB CDR3aa sequences with a specific specificity (A) and their usage (B) between females and males across cell subtypes, using the log2 fold change of the median values (females over males), following each specificity group. These groups of specificity are additionally classified as microorganism in the top section (bacteria in gold and virus in light blue) and self at the bottom (AID in magenta, cancer in red, human in blue-green). Polyspecific CDR3aa are defined here as CDR3aa capable of recognizing multiple antigens from different organisms (for no self-antigens) or from different specificity groups (e.g. categorized microorganisms, categorized self-antigens, allergens…). The proportion of unique polyspecific CDR3aa in the specific repertoire (C) and their usage (D) is compared between males and females, in DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and CD4 Treg (N = 14) cells. Stars indicate statistical differences between males and females based on the p-value of the Wilcoxon test (*: p < 0.05, **: p < 0.01). Males are depicted in violet and females in orange.

In conclusion, our analysis of TCR specificity has identified sex-specific differences in the thymic selection of CD8 effector T cells associated to autoimmune and bacterial antigens that are overrepresented in female versus males.

Discussion

It is noteworthy that there is a paucity of research on potential differences in the TCR repertoire between the sexes, despite the much higher prevalence of autoimmune diseases in females, a topic that remains to be fully elucidated. To explore the hypothesis that the generation and/or the selection of the TCR repertoires may be different between sexes requires, it is necessary to study the TCR repertoire where they are formed and selected, in the thymus. We thus generated a unique and valuable dataset comprising thymic samples collected from deceased organ donors of various ages, and young children who have undergone cardiac surgery. Firstly, we separated the various thymocytes populations in order to study both TCR repertoire generation at the level of DP cells, and repertoire selection at the level of CD8, CD4 Teff and Treg SP cells. We performed bulk sequencing of the TCRs from these cells, as this is the only method that yet generates enough sequences for sensitive comparison at this time. Single-cell sequencing is an analyzing technique that it is limited in its scope. It is only capable of analyzing a few thousand cells and would not have enabled the analyses that can be performed with hundreds of thousands of cells.

TCR generation

Our analyses initially revealed similarities rather than dissimilarities in TCR generation. Our detailed analyses revealed no significant sex differences in relation to the majority of the examined variables when comparing males and females DP cells’ repertoires, except for differences in the usage of small number of V and J genes from both TCR chains. However, these minor discrepancies are likely attributable to random variation, considering the extensive number of comparisons conducted. This finding is further supported by the observation that there are no significant differences in the usage of TRBV-TRBJ and TRAV-TRAJ gene associations between males and females. We also observed no differences in diversity indexes in DP cells’ repertories between males and females. Altogether, the overall combinatorial diversity of the TCR repertoire remains comparable between sexes.

A detailed analysis of the CDR3aa region in DP cells showed that some hydrophobic aa in the FG loop region are differently found for both TCR chains. However, no significant sex-based differences were found in the usage of hydrophobic, hydrophilic, or neutral aa at the critical p109 and p110 positions in TRB that have been described as influencing self-antigen recognition due to the impact of hydrophobic aa on TCR self-reactivity (26–28). These findings suggest that the subtle sex-based differences in the FG loop’s aa usage are minor and do not significantly impact the overall characteristics of the CDR3 region in a sex-dependent manner. The Pgen analysis of DP cells revealed only minimal differences in the distribution of Pgen between males and females, which were similar to those observed in control groups. This indicates that the mechanisms governing TCR generation are largely consistent across sexes, with only minor, likely random, deviations.

Additionally, although we identified TRB CDR3aa sequence motifs that were differentially expressed between males and females at the DP cell stage in our dataset, these motifs could not reliably differentiate individuals by sex in external datasets. This indicates that they are more dataset specific than sex-specific. This issue may also be explained by the different characteristics of our data and that of the other dataset. For example, in the Arstila dataset, the analyses were performed on unsorted cells from young subjects. Younger individuals tend to have a more diverse and richer TCR repertoire, with greater repertoire sharing in the periphery, which could obscure sex-specific patterns (16, 44). This suggests that these motifs might be context-specific or influenced by other factors such as age, genetic background, or environmental exposures. Moreover, we did not observe any sex-specific differences in CDR3 TRB specificity usage in DP cells’ repertoires. In summary, our results from DP cells analyses could not identify any relevant sex differences in TCR generation.

TCR selection

We identified three genes that show differential usage between males and females. Whilst it is possible that some of these differences are merely random, resulting from multiple comparisons, the preferential use of TRBV6-5 in females has previously been observed in the peripheral TCR repertoire (45). Notably, this gene is overrepresented in female lupus patients compared to their controls (51,52). Further research is needed to validate this observation and then elucidate its significance.

Our most striking observation relates to significant sex-specific differences in inferred specificities of CD8 T cells, with a biased CD8 selection process in the thymus favoring sequences associated with autoimmune diseases and bacterial antigens in females. The fact that this bias does not affect DP cells indicates that it is related to TCR selection rather than generation. Its relevance is strongly supported by the fact that (i) there is no such bias for other self-antigens that are not associated with autoimmunity and (ii) the bias affects both the effectors and regulators of autoimmunity with a directionality compatible with the increased prevalence of autoimmune disease in females. TCRs associated with autoimmunity related self-antigens have been observed to beincreased in CD8 effector T cells, while decreased levels have been seen in the CD4 Tregs. It should be noted that this bias also affects some bacterial antigens, though not viral antigens, are also affected by this bias. This phenomenom could potentially be explained by the fact that bacterial antigens express numerous mimotopes of self-antigens, and are known to contribute to the shaping of the TCR repertoire. The observed bias towards bacterial antigens in female CD8 T cells could likewise provide more effector cells specific for bacterial antigens that mimic self-antigens, contributing to autoimmune disease development by molecular mimicry (48). In this line, antigens linked to celiac disease (CeD) and type 1 diabetes (T1D) are highly represented in the database and these two AIDs have been linked to bacterial components that may mimic self-antigens (49–53). As is well-established that the gut microbiota influences thymic development of T cells (54). Sexual hormones could influence this biased selection by affecting the composition of the gut microbiota (55) and/or by affecting the selection of the TCRs specific for these antigens. In addition to these specific patterns, we observed a significant increase in the usage of hydrophobic amino acids at the central p109 position of the TRB CDR3aa in female CD8 thymocytes, with similar trend at p110. Hydrophobic residues at these positions have been implicated in enhanced self-reactivity and cross-reactivity of CD8 T cells (26–28), suggesting that female CD8 thymocytes may be slightly enriched for TCRs with intrinsically higher potential for recognizing self-like or cross-reactive peptide MHC complexes. It is noteworthy that the positional hydrophobicity of CD8 T cells CDR3s is consistent with the over-representation of sequences annotated as recognizing autoimmune-associated self-antigens, pointing to a more autoreactive TCR repertoire.

Finally, we have recently described polyspecific TCRs that can respond to multiple unrelated viral antigens and hypothesized their possible involvement in autoimmunity (43). We defined a CDR3 as being polyspecific if the TCRs containing it have been experimentally associated with epitopes from at least two different antigen origins. We observed a higher usage of polyspecific sequences in the CD8 SP repertoire of females, although this was not statistically significant. In contrast, the frequency of polyspecific TCR sequence usage was lower in female CD4 Tregs SP cells. This could limit CD4 Treg antigen coverage and be detrimental to prevent autoimmunity. However, little is known about polyspecific TCRs and these observations warrant further investigation.

These inferences rely on a harmonized specificity compendium combining McPAS-TCR, IEDB and VDJdb, which provide complementary and only partially overlapping landscapes (pathology-enriched versus predominantly viral specificities). Although there is limited numerical overlap between our thymic repertoires and annotated sequences is limited, this is to be expected given that public databases currently capture only a small fraction of the potential TCR specificity space. In this context, the sex-biased enrichment of sequences annotated as autoimmune- and bacteria-associated should be viewed as a robust signal emerging from a very sparse annotation space. However, it should still be interpreted as hypothesis-generating rather than exhaustive.

To further evaluate the robustness of our findings, we performed additional analyses to assess whether donor age could confound the observed sex-specific differences. A potential limitation of our study is the relatively wide age range of donors. However, the absence of age-related clustering in repertoire features (data not shown), and the lack of age impact on CD8 and CD4 Treg bacterial-specific TRB usage and CD8 and CD4 Treg self-associated to autoimmune disease-specific TRB usage patterns (Supplemental Figure 13), argue against age as a confounding factor. These results support the hypothesis of an intrinsic sex bias in thymic TCR selection rather than an age-related effect.

While our findings provide valuable insights into sex-specific differences in thymic selection of TCR repertoires, our TCR specificity analyses were limited. We can confirm that our study focused exclusively on the TRB CDR3aa region of the TCR, which is widely acknowledge as the most critical element in defining TCR specificity (31, 32) while both the TRA and the TRB chains contribute to TCR specificity (31, 32). This limitation is intrinsic to the limited number of TRA with assigned specificity in databases and to the fact that, given the nature of our studies, there is a need to study large repertoires, which is currently not feasible by single cell TCR sequencing. However, as methods for inferring specificity continue to evolve, they will likely allow more refined analyses of our dataset in the future (56). Furthermore, the continuous development of the TCR database should enhance these analyses.

Despite the relatively small size of our cohort, which was outbred, we believe that this design is essential for capturing sex-associated differences. Inbred models would reduce inter-individual variability, but they would not reflect the complexity of human thymic selection, nor the interplay between sex and genetic diversity. Despite the modest sample size, the consistency of the observed patterns across the analytical approaches supports the robustness of our conclusions.

Altogether, our analyses of TCR specificity identified a biased selection in females towards sequences associated with autoimmune diseases, potentially underlying the higher prevalence of autoimmune diseases in females. Future research should focus on validating these observations in relevant independent datasets, which are currently lacking.

Methods

Samples

Thymus samples were obtained from twenty-two organ donors aged between 73 days and 64 years, without any specific pathologies (Figure 1). Adult samples were collected post-surgery at the Cardiac Surgery Departments of many hospitals in France, following approval by the Agence de Biomédecine and the Ministry of Research (#PFS14-009). Pediatric samples were obtained following total thymectomy during cardiac surgeries at Necker Hospital. The ratio of male to female across donors is maintained at 1:1, although it varies between 1.1:1 and 1.33:1 depending on the cell subtypes analyzed. The age distributions of male and female donors were compared using two-sample Kolmogorov–Smirnov tests, with no significant differences observed across thymic subsets (p > 0.46), indicating no age-related confounding (Supplementary Figure 1).

Thymocyte isolation and RNA extraction

Thymic cell suspensions were prepared through automated tissue dissociation using gentleMACS dissociators (Miltenyi®) or by mechanical disruption, followed by filtration through a 70-µm nylon mesh. The cells were stained with anti-CD3 (AF700), anti-CD4 (APC), anti-CD8 (FITC), and for some samples, anti-CD25 (PE) antibodies. Cell sorting was performed by fluorescent activated cell sorting (FACS) on a Becton Dickinson FACSAriaII, achieving a purity of >85%. The sorted populations included DP CD3+ (CD3+CD4+CD8+), SP CD8+ (CD3+CD4-CD8+), and SP CD4+ (CD3+CD4+CD8-). A further separation was made into SP CD4 Teff (CD3+CD4+CD8-CD25-) and SP CD4 Treg (CD3+CD4+CD8-CD25+) for sixteen of the samples (Figure 1). For the sorted SP CD4, the Treg percentage varied between 5.8% and 16% across samples, with an average of 9.63%. Six samples of SP CD4 samples were not sorted into Teffs and Tregs, we thus categorized them as SP CD4 Teffs, assuming a purity of Teffs within the total SP CD4 of around 90%. Please refer to the detailed sample metadata in Supplementary Table 2 of the supplementary material. This includes donor IDs and the number of sorted cells per thymic subset for each donor.

The RNA was isolated using the RNAqueous extraction kit (Invitrogen®), in accordance with the manufacturer’s protocol. The RNA samples were quantified and the integrity of each sample was determined using a Nanodrop (Thermo Fisher) or a Tapestation 4200 (Agilent, RNA ScreenTape).

TCR library preparation and sequencing

TCR library preparation and sequencing were performed as described by Barennes et al. (57). Briefly, RNA was processed using the SMARTer Human TCR a/b profiling v1 kit (Takarabio), in accordance with the manufacturer’s protocol. Amplicons were then purified using AMPure XP beads (Beckman Coulter) and their quantity was determined. The integrity of the amplicons was then assessed using the 2100 Bioanalyzer System (Agilent, DNA 1000 kit) or Tapestation 4200 (Agilent, D1000 screentape). Bulk Next-Generation Sequencing was performed using either a Hiseq 2500 (Illumina) with the SR-300 protocol plus 10% PhiX on the LIGAN-PM Genomics platform (Lille, France), or a Novaseq 6000 (Illumina) with the PE-250 protocol plus 10% PhiX on the LIGAN-PM Genomics platform or ICM platform (Paris, France).

Data processing

The raw FASTQ/FASTA files were aligned to TRA and TRB using MiXCR (v3.0.13, RNA-seq parameters), a software solution that is able to correct for both sequencing and PCR errors (58). Samples with fewer than 1,000 CDR3aa sequences, an imbalanced TRA/TRB ratio (<1/9), or a low sequence count relative to sorted cell counts (< 0.9 ratio) were removed. Sequences with CDR3aa lengths outside 6-23 amino acids were also discarded. To standardize clonotype counts and the number of unique clonotypes across sequencing techniques, the initial counts for each clonotype were reduced by 1 to exclude those with counts of 0 (singletons). Furthermore, due to the high similarity between certain TRBV sequences, which has the potential to influence the analysis depending on the sequencing method used, TRBV06-2 and TRBV06-3 were merged as TRBV06-2/3, and TRBV12-3 and TRBV12-4 were combined as TRBV12-3/4 (59).Please refer to the summary table provided in the supplementary material (Supplementary Table 2) for a comprehensive overview of the total number of TCR sequences retained after preprocessing, as well as the corresponding number of unique clonotypes, for each donor, chain and thymic subset.

Gene usage analysis

For both TRA and TRB chains across all cell subtypes, we analyzed the V and J gene usage and those of the VJ gene associations (Figure 1).

The usage of gene is calculated as follows:

Where:

Σ𝑖𝜖𝐺𝑐𝑖 represents the total number of counts for clonotypes associated with the specified gene or gene association.

Σ𝑗𝜖𝑆𝑐𝑗 represents the total sum of the counts of all clonotypes in the sample.

In this notation, 𝐺 is the set of clonotypes using the interested gene, 𝑆 is the set of all clonotypes in the sample, and ci and cj are the counts of the clonotypes i and j, respectively. The usage of the V gene and the VJ gene combinations was investigated through dimensional reduction using Principal Component Analysis (PCA) (MASS package v7.3-60, factoMineR package v2.8, factoextra package v1.0.7). Confidence ellipses were represented with a 5% alpha risk. Furthermore, for VJ gene combinations usage, a heatmap was constructed using the Jensen-Shannon Divergence (JSD) distance to measure the distance between samples based on their gene usage distribution (complexHeatmap package v2.16.0, philentropy package v0.7.0). Hierarchical clustering was performed using the Euclidean distance and the complete linkage method.

Diversity analysis

For the TCR repertoire diversity analysis, to correct for sequencing depth bias and differences in cellular richness between individuals, rarefactions were performed fifty times for each sample, using X clonotypes, where X represents the effective diversity number (60).

The effective diversity was calculated as the exponential of the Shannon index (𝑒𝐻) (61). Rényi indices (11 indices from 0 to Infinity) were calculated for each rarefaction (62).

The median Shannon index, Simpson index and Berger-Parker indices were then compared between the sexes.

Distribution of CDR3aa length analysis

The calculation of the usage of all CDR3 by the amino acid length as follows:

Where

Σ𝑖𝜖𝐿𝑐𝑖represents the sum of the counts of the CDR3 characterized by the amino acid length of interest.

Σ𝑗𝜖𝑆𝑐𝑗 represents the total sum of counts of all CDR3s per sample.

In this notation, 𝐿 is the set of CDR3s characterized by the amino acid length of interest, 𝑆 is the set of all CDR3s in the sample, and ci and cj are the counts of the CDR3s i and j, respectively.

Composition usage of amino acid within the CDR3 region

All CDR3aa sequence were aligned using the IMGT convention (positions p104-p118) (63). Three complementary scales were used in the study. Firstly, sequence level aa composition. Within the CDR3 loop region (IMGT positions p108 to p114), involved in MHC-peptide complex interactions (63), aa usage at the sequence level was quantified for CDR3aa sequences 9-21 aa long in both TRA and TRB chains. For each sequence, the percentage of a given aa was calculated as follows: Where 𝑛𝑎𝑎is the number of occurences of the amino acid of interest within the CDR3 loop, and L denotes the total length of the CDR3aa sequence. . Furthermore, amino acids were categorized into three hydropathy classifications (neutral, hydrophobic, and hydrophilic) using the IMGT classification that takes into account the Kyte-Doolittle hydropathy index (25,64)

Secondly, the position-specific aa usage at p109 and p110. To assess positional composition, for each individual, CDR3aa usage is examined at IMGT positions p109 and p110, corresponding to p6-p7 positions described by Stadinsky et al. and Khosravi-Maharlooei et al., which hydrophobic amino acids at these two central positions have been implicated in modulating TCR self-reactivity and cross-reactivity (26–28). For a hydrophobic amino acid 𝑕 (excluding alanine, due to its weak hydrophobicity) at position 𝑝, usage was defined: Where 𝑛 is the total number of unique CDR3 sequences and 𝑁(𝑕,𝑝) the number of unique CDR3 sequences carrying hydrophobic amino acid 𝑎 at position 𝑝. This analysis quantifies amino acid composition at fixed positions independently of clonotype frequency.

Probability of generation analysis

The probability of generation analysis of DP and CD8 cell subtypes consisted initially in the creation of a generative model of V(D)J recombination using 400,000 randomly selected non-productive sequences from all samples for both TRA and TRB chains, with the IGOR tool (65). The generation probabilities (Pgen) for all sequences in the dataset were subsequently calculated using the OLGA tool, based on the respective generative models for each cell subtype and TCR chain (66). Comparisons of Pgen values at the nucleotide level between males and females were performed. To ensure statistical robustness, control groups with equivalent male and female sample numbers were created by permuting the samples into two groups 20,000 times.

TCR network structure by sequence similarity

For each sample, 100 random sub-samplings of CDR3aa sequences were performed, using the minimum CDR3aa count per cell subtype. Two CDR3aa were connected if their Levenshtein distance was = 1 (Figure 1) (67). Network analysis focused on two metrics: the proportion of connected sequences and network density, defined as the ratio of actual connections to all possible connections. Median values from 100 subsampling iterations were then analyzed. Levenshtein distances were computed with the stringdist package (v0.9.10), networks were visualized using Cytoscape (v3.8.2), and network density was calculated with the igraph package (v1.5.1).

TRB CDR3aa motif search

Two types of structural motifs were searched: local motifs (strict sequential motifs) and global motifs (sequences with substitutions preserving a positive BLOSUM62 score) (Figure 1).

The selection of enriched motifs was carried out using the gliph2 function with the turboGliph package (v0.99.2), with specific parameters defined as follows:

- The input sequences were from the sex group of interest (male or female), with reference sequences matched for TRBV gene usage.

- It is imperative that CDR3aa sequences are longer than eight aa.

- Local motif lengths were restricted to 3-5 aa.

- Only motifs consisting of more than two unique CDR3aa were conserved.

- Local motif significance of Fisher’s test has not been boosted.

Additional filters were applied so that: (i) the inclusion of public CDR3aa sequences is a prerequisite for a motif, with these sequences shared by at least two individuals, (ii) a significant enrichment is required, as determined by Fisher’s test with a p-value p < 0.01 and (iii) a usage difference between groups of at least twofold is necessary, as evidenced by a Wilcoxon test with a p-value p < 0.05.

Motif validation was performed on an openly published pediatric thymic dataset from the Arstila team (35, 36) consisting in a bulk TCR dataset from infants aged between seven days and eight months. Here, we removed one male subject “Thymus A”, who was a twin of the “Thymus B” and the “Thymus 1” neonate sample, in order to analyze height samples with a male-to-female ratio of 5:3. Further validation was carried out on sorted peripheral blood TCR repertoires (CD8, CD4 Teff and CD4 Treg) from seventy-five healthy volunteers aged from eighteen to eighty-four years (male-to-female ratio of 1.03:1), whose libraries were generated under the same conditions as those of our thymic dataset. They were recruited within the Transimmunom observational trial (NCT02466217) and HEALTHIL-2 (NCT03837093) (68). PBMCs from healthy individuals were isolated using Ficoll density gradient centrifugation and enriched for CD4+ T cells via EasySep™ Human CD4+ T Cell magnetic beads (Stemcell). The samples were then separated into two groups: effector T cells (CD4+ CD25-) and Tregs (CD4+ CD25+).This was performed using EasySep™ Human Pan-CD25 magnetic beads (Stemcell). Purity was assessed by flow cytometry based on the expression of CD4 and FoxP3, with a purity threshold of >80%. From the frozen PBMC aliquots, CD8+ T cells (CD3+ CD8+) were sorted using a FACS ARIA II cell sorter.

TRB CDR3aa specificity analysis

We have established a curated and harmonized TCR specificity database by integrating data from three public TCR repositories: Mc-PAS, IEDB, and VDJdb (38, 39, 69) (data collected as of October 2023) (Figure 1). The harmonization procedure was carried out in accordance with the framework outlined in Jouannet et al. (70) and included :

- Uniform formatting of TRA/TRB TCR sequences and associated metadata,

- Consolidation of redundant entries (identical TRA/TRB TCR, epitope, organism, PubMed ID, and cell subset),

- Maintening separate entries when source databases are incompatible,

Reliability levels for sequences and specificity assignments were quantified using the Verified_score (VS) and Antigen-identification_score (AIS) as defined in (70). The VS, ranging from 0 to 2, reflects the concordance between calculated and curated CDR3 boundaries following the IEDB verification strategy (VS = 2: both TRA and TRB present and verified; VS = 1.1: only TRA verified; VS = 1.2: only TRB verified; VS = 0: no verified chain) and the AIS (0 to 5), which ranks the strength of antigen-identification methods (70).

In this study, we focused our analyses on high-reliability TRB sequences, defined as those with a verified TRB CDR3aa (VS ≥ 1.2) and an AIS corresponding to in vitro T cell stimulation with a pathogen, protein or peptide, or pMHC X-mer sorting (AIS > 3.2, excluding categories 4.1 and 4.2). This ensures that the specificity assignments relied on robust experimental evidence.

Specificity groups were categorized according to antigen class, including viral, bacterial, yeast, parasitic and self-antigens as along with antigens derived from wheat, plants or animals. Self-antigens were then annotated into three categories: autoimmune-associated antigens, cancer-associated antigens and self-antigens unrelated to disease. For autoimmune-associated antigens, assignments were based on explicit annotations present in McPAS-TCR, IEDB and VDJdb. These databases identify celiac disease (CeD) and type 1 diabetes (T1D) related antigens, which were retained as such in the harmonized dataset. For cancer-associated antigens, the following classifications were used: (i) information present in the original databases and (ii) cross-referencing with external cancer antigen resources, including the Cancer Antigenic Peptide Database (de Duve Institute), the Human Protein Atlas and the cancer database caAtlas (71). Please refer to Supplementary Table 3, which provides a comprehensive list of all self-antigens present in the harmonized database. This table also includes their associated disease category and their mapping into the specificity groups used in the manuscript.

This harmonized database was used to identify exact matches with the CDR3aa sequences of our TRB sequences. The enrichment of CDR3aa sequences associated with a given specificity group in male and female thymic repertoires for each cell subtype compared to the specificity distribution in the database was tested. Furthermore, we performed specificity group distributions between male and female samples, as well as was the overall usage of these sequences across entire TCR repertoires.

Additionally, our research concentrated on polyspecific sequences defined as CDR3aa TRB sequences recognizing at least two distinct microbial species or belonging to multiple specificity groups for non-microbial antigens. In this study, polyspecificity is therefore considered at the level of broader antigenic categories rather than individual epitopes and thus reflects functional cross-reactivity patterns aggregated across studies, instead of implying molecular-level multi-epitope recognition. These polyspecific TRB CDR3aa sequences underwent the same statistical analyses as monofunctional sequences.

Statistical analysis

For V-J gene usage, CDR3aa length usage, CDR3aa usage, network similarity and specificity analyses, the comparison between males and females was statistically analysed using the Wilcoxon test. The statistical tests were performed using the ggpubr library (v0.6.0) in R. The number of asterisks denotes the significance of the results: one asterisk indicates a p-value p ≤ 0.05, two asterisk indicate a p-value p ≤ 0.01, and so on until four asterisks, denoting a p-value p ≤ 0.0001. The comparisons of the Rényi curves and Pgen were compared using the Kolmogorov-Smirnov test. The Wilcoxon test was used to test the comparisons of the proportions of specific sequences in male and female subjects relative to those in the specificity database.

eLife Assessment

This study provides valuable insights into addressing the question of whether the prevalence of autoimmune disease could be driven by sex differences in the T cell receptor (TCR) repertoire, correlating with higher rates of autoimmune disease in females. The authors compared male and female TCR repertoires using bulk RNA sequencing, from sorted thymocyte subpopulations in pediatric and adult human thymuses; however, the analyses provided do not provide sufficient discrimination and incompletely support the central claims regarding sex differences in the TCR repertoire and potential autoimmune bias.

Reviewer #1 (Public review):

Summary:

The goal of this paper was to determine whether the T cell receptor (TCR) repertoire differs between a male or female human. To address this, this group sequenced TCRs from double-positive and single-positive thymocytes in male and female humans of various ages. Such an analysis on sorted thymocyte subsets has not been performed in the past. The only comparable dataset is a pediatric thymocyte dataset where total thymocytes were sorted.

They report on participant ages and sexes, but not on ethnicity, race, nor provide information about HLA typing of individuals. The experiments are heroic, yet do represent a relatively small sampling of diverse humans. They observed no differences in TCRbeta or TCRalpha usage, combinational diversity, or differences in the length of the CDR3 region, or amino acid usage in the CD3aa region between males or females. Though they observed some TCRbeta CD3aa sequence motifs that differed between males and females, these findings could not be replicated using an external dataset and therefore were not generalizable to the human population.

They also compared TCRbeta sequences against those identified in the past databases using computational approaches to recognize cancer-, bacterial-, viral-, or autoimmune-antigens. They found little overlap of their sequences with these annotated sequences (depending on the individual, ranged from 0.82-3.58% of sequences). Within the sequences that were in overlap, they found that certain sequences against autoimmune or bacterial antigens were significantly over-represented in female versus male CD8 SP cells. Since no other comparable dataset is available, they could not conclude whether this is a generalizable finding in the human population.

Strengths:

It is a novel dataset that attempts to understand sex differences in the T cell repertoire in humans. Overall, the methodologies are sound and are the current state-of-the-art. There was an attempt to replicate their findings in cases where an appropriate dataset was available. I agree that there are no gross differences in TCR diversity between males and females. This is an important negative result.

Weaknesses:

Overall, the sample size is small given that it is an outbred population. This reviewer recognizes the difficulty in obtaining samples for this experiment (which were from deceased donors), and this limitation was appropriately discussed. Their analysis was limited by the current availability of other TCR sequences. These weaknesses were appropriately discussed and considered.

Reviewer #2 (Public review):

Summary:

This study addresses the hypothesis that the strikingly higher prevalence of autoimmune diseases in women could be the result of biased thymic generation or selection of TCR repertoires. The biological question is important and the hypothesis is valuable. Although the topic is conceptually interesting and the dataset is rich, the study has a number of major issues. In particular, the majority of "autoimmunity-related TCRs" considered in this study are in fact specific to type 1 diabetes (T1D). Notably, T1D incidence is higher in males, which directly contradicts the stated objective of the study - to explain the higher prevalence of autoimmune diseases in women. Given this conceptual inconsistency, the evidence presented does not support the authors' conclusions.

Strengths:

The key strength of this work is the newly generated dataset of TCR repertoires from sorted thymocyte subsets (DP and SP populations). This approach enables the authors to distinguish between biases in TCR generation (DP) and thymic selection (SP). Bulk TCR sequencing allows deeper repertoire coverage than single-cell approaches, which is valuable here, although the absence of TRA-TRB pairing and HLA context limits the interpretability of antigen specificity analyses. Importantly, this dataset represents a valuable community resource and should be openly deposited rather than being "available upon request."

Weaknesses:

I thank the authors for their detailed responses to my previous comments. Several concerns were addressed satisfactorily; however, important issues remain unresolved, and a new major concern has emerged from the revised manuscript.

Major concerns:

(1) Autoimmune specificity is dominated by T1D, contradicting the study's premise. Newly added supplementary Table 3 shows that the authors considered only 14 autoimmune-related epitopes, of which 12 are associated with type 1 diabetes (T1D) and 2 with celiac disease (CeD). (I guess this is because identification of particular peptide autoantigens is an extremely difficult task and was only successful in T1D and CeD.) Thus conclusions of this work mostly relate to T1D. However, the incidence of T1D is higher in males than in females (e.g. doi:10.1111/j.1365-2796.2007.01896.x; doi:10.25646/11439.2). This directly contradicts the stated objective of the study - to explain the higher prevalence of autoimmune diseases in women. As a result, the authors' conclusions (a) cannot be generalized to autoimmune disease as a whole as the authors only considered T1D and CeD antigens and (b) are internally inconsistent with the stated objective of the study.

(2) By contrast, CeD does show a female bias (~60/40 female/male; doi:10.1016/j.cgh.2018.11.013). However, the manuscript does not allow evaluation of how much the reported "autoimmune TCR enrichment" derives from T1D versus CeD. Despite my previous request, the authors did not provide per-donor and per-epitope distributions of autoimmune-specific TCR matches. I therefore explicitly request a table in which: each row corresponds to a specific autoimmune antigen; each column corresponds to a donor (with metadata available including sex); each cell reports the number of unique TCRs specific to that antigen in that donor. Without such data, the conclusions cannot be evaluated.

(3) It is scientifically inappropriate to generalize findings to "autoimmune diseases" when only T1D and CeD were analyzed. Moreover, given that T1D and CeD show opposite directions of sex bias, combining them into a single "AID" category is misleading. All analyses presented in Figure 8 and Supplementary Figure 16 should be repeated and shown separately for T1D and CeD, rather than combined.

(4) The McPAS database contains TCRs associated with other autoimmune diseases (e.g., multiple sclerosis, rheumatoid arthritis), although the exact autoantigens in these contexts are unknown. Why didn't the authors perform the search for such TCRs? I believe disease association even without particular known antigen could still be insightful.

(5) Misuse of the concept of polyspecificity. I appreciate the authors' reference to Don Mason's work; however, the concept of polyspecificity discussed there is fundamentally different from the authors' usage. Mason, Sewell (doi:10.1074/jbc.M111.289488), Garcia (doi:10.1016/j.cell.2014.03.047), and others demonstrated that individual TCRs can recognize multiple peptides, possibly around 1 million. But importantly these peptides are not random but share some sequence motif. This is a general feature of TCRs, i.e. 100% of TCRs are polyspecific in this sense.

In contrast, the authors define polyspecificity as TRB sequences annotated as specific to unrelated epitopes in TCR databases such as VDJdb. These databases are well known to contain substantial numbers of false-positive annotations (see, e.g., Ton Schumacher's preprint https://www.biorxiv.org/content/10.1101/2025.04.28.651095.abstract). The authors acknowledge that, under their definition, polyspecificity has been experimentally validated for only one (!) TCR (Quiniou et al.). In the absence of robust experimental validation, use of the term "polyspecificity" in this context is misleading. I strongly recommend removing all analyses and conclusions related to polyspecificity from the manuscript unless supported by independent functional validation.

(6) I agree that comparing specificity enrichment between sexes is meaningful. However, enrichment relative to the database composition itself is not biologically interpretable, as acknowledged by the authors in their response. I therefore recommend removing Supplementary Figure 15, which is potentially misleading.

(7) In contrast, Supplementary Figure 16 represents the most convincing result of the study (keeping in mind that the AID group should be splitted to T1D and CeD with T1D and that T1D and CeD have opposing directions of sex biases) and should be shown as a main figure, replacing Figure 8A-B which is less convincing as it doesn't show per-donor distribution.

(8) The authors argue that applying mixed-effects modeling to Rényi entropy would require assuming a common sex effect across subsets. I do not find this assumption unreasonable. For example, if sex effects are mediated through AIRE-dependent negative selection, one would indeed expect a consistent direction of effect across subsets. The lack of statistical significance in Figure 3 may reflect limited sample size rather than true absence of the difference. Moreover, the title's phrasing "comparable TCR repertoire diversity" is vague: what is the statistical definition of "comparable"?

Author Response:

The following is the authors’ response to the previous reviews

Public Reviews:

Reviewer #1 (Public review):

Summary:

The goal of this paper was to determine whether the T cell receptor (TCR) repertoire differs between a male or female human. To address this, this group sequenced TCRs from doublepositive and single-positive thymocytes in male and female humans of various ages. Such an analysis on sorted thymocyte subsets has not been performed in the past. The only comparable dataset is a pediatric thymocyte dataset where total thymocytes were sorted.

They report on participant ages and sexes, but not on ethnicity, race, nor provide information about HLA typing of individuals. The experiments are heroic, yet do represent a relatively small sampling of diverse humans. They observed no differences in TCRbeta or TCRalpha usage, combinational diversity, or differences in the length of the CDR3 region, or amino acid usage in the CD3aa region between males or females. Though they observed some TCRbeta CD3aa sequence motifs that differed between males and females, these findings could not be replicated using an external dataset and therefore were not generalizable to the human population.

They also compared TCRbeta sequences against those identified in the past databases using computational approaches to recognize cancer-, bacterial-, viral-, or autoimmune-antigens. They found little overlap of their sequences with these annotated sequences (depending on the individual, ranged from 0.82-3.58% of sequences). Within the sequences that were in overlap, they found that certain sequences against autoimmune or bacterial antigens were significantly over-represented in female versus male CD8 SP cells. Since no other comparable dataset is available, they could not conclude whether this is a generalizable finding in the human population.

Strengths:

It is a novel dataset that attempts to understand sex differences in the T cell repertoire in humans. Overall, the methodologies are sound and are the current state-of-the-art. There was an attempt to replicate their findings in cases where an appropriate dataset was available. I agree that there are no gross differences in TCR diversity between males and females. This is an important negative result.

Weaknesses:

Weaknesses:

Overall, the sample size is small given that it is an outbred population. This reviewer recognizes the difficulty in obtaining samples for this experiment (which were from deceased donors), and this limitation was appropriately discussed. Their analysis was limited by the current availability of other TCR sequences. These weaknesses were appropriately discussed and considered.

We thank this reviewer for his appreciation of our work.

Reviewer #2 (Public review):

Summary:

This study addresses the hypothesis that the strikingly higher prevalence of autoimmune diseases in women could be the result of biased thymic generation or selection of TCR repertoires. The biological question is important and the hypothesis is valuable. Although the topic is conceptually interesting and the dataset is rich, the study has a number of major issues. In particular, the majority of "autoimmunity-related TCRs" considered in this study are in fact specific to type 1 diabetes (T1D). Notably, T1D incidence is higher in males, which directly contradicts the stated objective of the study - to explain the higher prevalence of autoimmune diseases in women. Given this conceptual inconsistency, the evidence presented does not support the authors' conclusions.

We disagree with the reviewer’s assertion that our findings create a conceptual inconsistency.

Autoimmune diseases are multifactorial conditions in which multiple biological layers, including thymic selection, peripheral immune regulation, hormonal effects, environmental exposures, and tissue-specific vulnerability, contribute to disease incidence. These layers may influence sex ratios in different directions. Therefore, observing a higher frequency of TCRs annotated as T1D-associated in females does not imply that T1D incidence must also be higher in females.

Actually, T1D incidence itself is not uniformly male-biased worldwide. Epidemiological analyses (reviewed in Qu and Hakonarson, Diabetes Obes Metab 2025) show that male predominance is mainly observed in high-incidence Northern European populations, whereas in several lowerincidence regions, including parts of East Asia and Africa, the sex ratio is balanced or even femalebiased. Furthermore, another recent study highlights that T1D incidence and prevalence in women and men varies depending on the study period (PMC12544016).

This heterogeneity indicates that disease incidence reflects context-dependent interactions between genetic load, environmental exposures, and sex-specific biological modifiers. Moreover, biological sex acts as a dynamic modifier of genetic risk and immune function in T1D, influencing central tolerance, peripheral immune activation, and β-cell intrinsic resilience (reviewed in Qu and Hakonarson, 2025). Experimental models further demonstrate estrogenmediated protection of pancreatic β-cells (Kim et al., Biochem Biophys Res Commun 2025), indicating that disease incidence reflects the integration of immune, hormonal, and tissuespecific layers rather than central autoreactive TCR release alone. Sex hormones may exert distinct and sometimes opposing effects on thymic selection and on target-organ vulnerability, while environmental factors such as vitamin D status, infections, and microbiota composition further shape disease expression.

Importantly, our study does not claim causality, nor does it aim to predict the epidemiology of any specific autoimmune disease. Our conclusions are limited to the observation that sexdependent differences exist in thymic TCR selection.

Strengths:

The key strength of this work is the newly generated dataset of TCR repertoires from sorted thymocyte subsets (DP and SP populations). This approach enables the authors to distinguish between biases in TCR generation (DP) and thymic selection (SP). Bulk TCR sequencing allows deeper repertoire coverage than single-cell approaches, which is valuable here, although the absence of TRA-TRB pairing and HLA context limits the interpretability of antigen specificity analyses. Importantly, this dataset represents a valuable community resource and should be openly deposited rather than being "available upon request."

We agree with the reviewer’s comment. As already stated in the previous revision and the "Data Availability" section of the manuscript, all raw sequencing data have been deposited and are publicly available on NCBI (BioProject PRJNA1379632): https://www.ncbi.nlm.nih.gov/sra/PRJNA1379632.

Weaknesses:

I thank the authors for their detailed responses to my previous comments. Several concerns were addressed satisfactorily; however, important issues remain unresolved, and a new major concern has emerged from the revised manuscript.

Major concerns:

(1) Autoimmune specificity is dominated by T1D, contradicting the study's premise. Newly added supplementary Table 3 shows that the authors considered only 14 autoimmune-related epitopes, of which 12 are associated with type 1 diabetes (T1D) and 2 with celiac disease (CeD). (I guess this is because identification of particular peptide autoantigens is an extremely difficult task and was only successful in T1D and CeD.) Thus conclusions of this work mostly relate to T1D. However, the incidence of T1D is higher in males than in females (e.g. doi:10.1111/j.13652796.2007.01896.x; doi:10.25646/11439.2). This directly contradicts the stated objective of the study - to explain the higher prevalence of autoimmune diseases in women. As a result, the authors' conclusions (a) cannot be generalized to autoimmune disease as a whole as the authors only considered T1D and CeD antigens and (b) are internally inconsistent with the stated objective of the study.

(2) By contrast, CeD does show a female bias (~60/40 female/male; doi: 10.1016/j.cgh.2018.11.013). However, the manuscript does not allow evaluation of how much the reported "autoimmune TCR enrichment" derives from T1D versus CeD. Despite my previous request, the authors did not provide per-donor and per-epitope distributions of autoimmune-specific TCR matches. I therefore explicitly request a table in which: each row corresponds to a specific autoimmune antigen; each column corresponds to a donor (with metadata available including sex); each cell reports the number of unique TCRs specific to that antigen in that donor. Without such data, the conclusions cannot be evaluated.

(3) It is scientifically inappropriate to generalize findings to "autoimmune diseases" when only T1D and CeD were analyzed. Moreover, given that T1D and CeD show opposite directions of sex bias, combining them into a single "AID" category is misleading. All analyses presented in Figure 8 and Supplementary Figure 16 should be repeated and shown separately for T1D and CeD, rather than combined.

We acknowledge that currently available antigen-annotated TCR databases remain limited. This reflects the considerable experimental difficulty of defining TCRs’ antigen specificities and is a widely recognized limitation in the field.

In the curated database used here, the autoimmune-associated entries correspond primarily to type 1 diabetes (T1D) and celiac disease (CeD), two autoimmune contexts for which antigen-specific TCRs have been experimentally characterized. However, focusing on the number of antigens alone does not accurately reflect the breadth of the dataset.

Specifically, our analysis is based on 48 epitopes and nearly 200 annotated TRB sequences, providing substantially broader antigenic representation than suggested by antigen count alone.

Importantly, our analytical framework does not attempt to interpret each epitope specificity individually. Instead, we examine whether TCRs annotated as autoimmune-associated are differentially represented between sexes at the level of thymic selection.

In our dataset we observe a stronger CD8⁺ thymic selection of TCRs annotated as autoimmune- associated in females. We interpret this as evidence that central tolerance mechanisms may contribute to sex-dependent differences in autoreactive repertoire composition, rather than as a determinant of any specific autoimmune disease pathophysiology.

(4) The McPAS database contains TCRs associated with other autoimmune diseases (e.g., multiple sclerosis, rheumatoid arthritis), although the exact autoantigens in these contexts are unknown. Why didn't the authors perform the search for such TCRs? I believe disease association even without particular known antigen could still be insightful.

For multiple sclerosis, the only antigen present in the database is myelin basic protein (MBP). In our thymic repertoire dataset, we could not detect any CDR3 sequence matching MPB annotated CDR3s from the database.

For rheumatoid arthritis, the database contains only a small number of TRA sequences without corresponding TRB chains. Because our specificity analysis is based on TRBs, these entries could not be used in our analyses.

(5) Misuse of the concept of polyspecificity. I appreciate the authors' reference to Don Mason's work; however, the concept of polyspecificity discussed there is fundamentally different from the authors' usage. Mason, Sewell (doi:10.1074/jbc.M111.289488), Garcia(doi:10.1016/j.cell.2014.03.047), and others demonstrated that individual TCRs can recognize multiple peptides, possibly around 1 million. But importantly these peptides are not random but share some sequence motif. This is a general feature of TCRs, i.e. 100% of TCRs are polyspecific in this sense.

In contrast, the authors define polyspecificity as TRB sequences annotated as specific to unrelated epitopes in TCR databases such as VDJdb. These databases are well known to contain substantial numbers of false-positive annotations (see, e.g., Ton Schumacher's preprint https://www.biorxiv.org/content/10.1101/2025.04.28.651095.abstract). The authors acknowledge that, under their definition, polyspecificity has been experimentally validated for only one (!) TCR (Quiniou et al.). In the absence of robust experimental validation, use of the term "polyspecificity" in this context is misleading. I strongly recommend removing all analyses and conclusions related to polyspecificity from the manuscript unless supported by independent functional validation.

We agree with the reviewer that the concept of TCR polyspecificity is complex, controversial and not uniformly defined in the literature.

For some, polyspecificity refers to the ability of individual TCRs to recognize multiple related peptides sharing structural motifs, as described by Mason, Sewell, Garcia, and others. With this definition, we agree that many/most TCRs exhibit some degree of cross-reactivity and would thus be defined as polyspecific.

In contrast, our definition of polyspecificity came from our observation arising from large-scale repertoire analyses that certain CDR3 sequences are repeatedly annotated across databases as recognizing distinct and unrelated antigenic categories. In our previous study (Quiniou et al.), we showed that these sequences display specific biochemical and repertoire features and may represent a particular class of TCRs involved in early or heterologous immune responses. A classic cross reactivity based on structural motif sharing could not explain these results.

We believe that the existence of such TCRs, rather than classic cross-reactive TCRs, has the potential to better explain why patients with extremely reduced TCR repertoires (around 3000 TCRs only) can respond well to various infectious challenges (https://doi.org/10.1073/pnas.97.1.274) or why there are T cells with memory phenotypes against viruses not previously encountered (https://pmc.ncbi.nlm.nih.gov/articles/PMC3626102/ ). We acknowledge that direct experimental validation of the function of such TCRs is currently limited; further work will help clarify the notion of polyspecificity, and hopefully to better understand the overlooked “heterologous immunity”.

Of note, a recent paper in Nature Machine Intelligence (https://doi.org/10.1038/s42256-02501096-6) described the in-silico generation of antigen-specific TCRs. Using our definition of polyspecificity (TCRs with higher generation probabilities, specific V/J gene preferences, shared CDR3s across individuals, and reactivity to multiple unrelated peptides), they showed that “multitask models preferentially sample polyspecific CDR3β sequences”. Therefore, we consider the debate on polyspecificity to be ongoing, and our discussion of polyspecificity in this paper to be part of this debate.

(6) I agree that comparing specificity enrichment between sexes is meaningful. However, enrichment relative to the database composition itself is not biologically interpretable, as acknowledged by the authors in their response. I therefore recommend removing Supplementary Figure 15, which is potentially misleading.

In the original manuscript, the comparison to the pooled database was intended as a descriptive assessment rather than as a biological enrichment analysis. Differences between an experimental thymic repertoire and a curated reference database are expected, given the structure and annotation biases inherent to the reference resource.

The purpose of Supplementary Figures 15B and 15C was therefore twofold: (i) to provide a descriptive overview of how specificity categories are distributed in our thymic dataset relative to the curated database, and (ii) to evaluate whether deviations from database proportions were of similar magnitude in males and females, ensuring that database composition did not differentially bias one sex over the other. In addition, the donor-resolved representations demonstrate that these patterns are consistent across individuals and are not driven by a single donor.

To avoid any potential misinterpretation, we have revised the manuscript to remove references to “enrichment” relative to database composition and eliminated quantitative comparisons to baseline database frequencies. The corresponding text and figure legends have been clarified to indicate that these analyses are descriptive and methodological in nature, while all biological interpretations rely exclusively on direct sex-specific comparisons within the thymic dataset.

(7) In contrast, Supplementary Figure 16 represents the most convincing result of the study (keeping in mind that the AID group should be splitted to T1D and CeD with T1D and that T1D and CeD have opposing directions of sex biases) and should be shown as a main figure, replacing Figure 8A-B which is less convincing as it doesn't show per-donor distribution.

(8) The authors argue that applying mixed-effects modeling to Rényi entropy would require assuming a common sex effect across subsets. I do not find this assumption unreasonable. For example, if sex effects are mediated through AIRE-dependent negative selection, one would indeed expect a consistent direction of effect across subsets. The lack of statistical significance in Figure 3 may reflect limited sample size rather than true absence of the difference. Moreover, the title's phrasing "comparable TCR repertoire diversity" is vague: what is the statistical definition of "comparable"?

The use of “comparable” in comparing TCR repertoire diversity is indeed “soft”, and aimed to indicate that there are no obvious dissimilarities.

Recommendations for the authors:

Reviewer #2 (Recommendations for the authors):

Minor comments:

(1) Available HLA typing data for selected donors should be included as a table in the manuscript.

The available low-resolution HLA typing data for the donors included in this study have been compiled and added as Supplementary Table 1 in the revised manuscript.

(2) The authors' explanation for why external validation of gene usage biases was not possible should be concisely incorporated into the Discussion.

We have incorporated a concise explanation in the Discussion clarifying why independent validation of the TRBV6-5 bias in external thymic datasets is currently not feasible, due to the absence of publicly available cohorts combining sorted thymic subsets, balanced sex representation, and sufficient sequencing depth.

(3) The clarification that considered sex-specific motifs are public should be included explicitly in the main text, not only figure legend and methods.

We now explicitly state in the main Results section that only public motifs, defined as motifs containing CDR3 sequences shared by at least two individuals, were retained in the analysis.

(4) The statement "Thymocytes expressing TCRs with insufficient or excessive avidity are eliminated (negative selection)" is strictly speaking incorrect. Thymocytes with insufficient avidity are eliminated by death by neglect during positive selection.

We thank the reviewer for pointing out this imprecision. The statement has been corrected.

(5) Figure 8C is unclear - what does "80% of unique polyspecific TCRs" mean? In any case, I strongly recommend removal of all polyspecificity-related analyses.

We apologize for the lack of clarity in the axis label of Figure 8C. To clarify, this analysis represents the proportion of polyspecific CDR3aa sequences among all sequences with an assigned specificity within an individual’s repertoire. Specifically, it measures how many unique TCR sequences, previously identified as having a known specificity in reference databases, are also categorized as polyspecific.

To address the reviewer’s concern, we have updated the Y-axis label of Figure 8C to: "Proportion of polyspecific CDR3aa among antigen-specific sequences (%)".

(6) "However, no significant sex-based differences were found in the usage of hydrophobic, hydrophilic, or neutral aa at the critical p109 and p110 positions in TRB" - this Discussion statement is inconsistent with the new analysis on Fig. 4C.

We regret that the Discussion still contained wording from a previous version of the analysis. The text has now been corrected to reflect the updated results showing a significant increase in hydrophobic amino acid usage at positions p109/p110.

(7) In the Discussion the authors write: "the absence of age-related clustering in repertoire features (data not shown)". What is the reasoning for not showing the data?

We understand the reviewer's point. This exploratory clustering analysis was performed on the data presented in the heatmaps (Figure 2B and Supplemental Figures 10-13). However, as it revealed no distinct patterns or clustering based on the donors' age (with samples from different age groups being interspersed throughout the clusters), we chose not to add an extra layer of annotation to Figure 2B to maintain clarity.

Version 3

Abstract

Women represent about 80% of patients with autoimmune diseases. This may partly result from sex-based differences in T cell receptor (TCR) selection during thymocyte development, potentially influenced by hormones and the lower expression of the Autoimmune Regulator (AIRE) transcription factor in females.

To investigate this, we analyzed sex-specific differences in TCR generation and selection. We examined TCR repertoires in double-positive thymocytes and single-positive thymic cells, including CD8⁺ and CD4⁺ effector T cells and regulatory T cells (Tregs), derived from male and female organ donors. Minimal sex-based differences were observed in V and J gene usage, and there were no notable differences in TCR repertoire diversity, complementarity-determining region 3 (CDR3) length, amino acid composition, or network structure. No TCR sequences were exclusive to either sex.

However, female effector T cells exhibited a significantly higher prevalence of TCRs specific to self-antigens implicated in autoimmunity compared to males, while female Tregs showed a reduced frequency of such TCRs. These differences were not observed for TCRs targeting self-antigens unrelated to autoimmunity or antigens associated with cancer or viruses.

Our findings identify a sex-specific imbalance in thymic selection of TCRs with autoimmunity-associated specificities, providing mechanistic insight into the increased susceptibility of women to autoimmune diseases.

Introduction

Several studies have highlighted a sex imbalance in diseases involving the immune system. This is particularly evident in the case of autoimmune diseases (AIDs), where 80% of affected individuals are women, and the severity of the condition varies between the sexes (1–3). In addition, men are more severely affected by infectious diseases such as COVID-19 and tuberculosis (4,5). The response to preventive immunotherapies such as vaccination, and to curative treatments, like immune checkpoint inhibitors or anti-TNF alpha, is also sex dependent (6). These observations indicate the presence of intrinsic biological differences in immune responses between males and females.

A variety of factors have been identified that could help explain the observed sex-based differences. Among these, sex hormones have been shown to influence immune responses by acting on immune cells that express specific receptors to these hormones (7–9). Sex hormones could directly influence the adaptive immune response at the level of thymocyte differentiation and selection by acting on the expression of Autoimmune Regulator (AIRE) in thymic epithelial cells. AIRE is responsible for the thymic expression of otherwise tissue-specific antigens, contributing to both the negative selection of effector cells and the positive selection of Tregs that recognize such antigens (10). Research has shown that testosterone increases AIRE expression in medullary thymic epithelial cells, whereas estradiol decreases it (11,12). Furthermore, a recent study has shown that the transcriptomic profiles of these specialized antigen-presenting cells in the thymus differ between males and females (13). This suggests that there may be differences in the selection of the T cell receptor (TCR) repertoire between males and females.

Despite these findings, there is still limited knowledge about potential differences in the TCR repertoire between the sexes. The generation of the TCR is a key process in T-cell development that takes place in the thymus. The TCR consists of two chains, alpha (TRA) and beta (TRB), each of which resulting from a random genetic rearrangement involving the V (variable), D (diversity) [for the TRB], and J (joining) genes. The combination of these gene segments is accompanied by insertions and deletions, creating a highly diverse region, the Complementarity Determining Region 3 (CDR3) (14). It is this region of the TCR that predominantly interacts with the peptide, a critical interaction for antigen recognition and subsequent T cell activation. The ability to specifically recognize diverse antigens via the CDR3 is central to the efficiency and specificity of the adaptive immune response.

TCRs are randomly generated, which means that while some may be useful, others may be harmful by recognizing self-antigens and causing autoimmunity. The current understanding is that during thymocyte development, selection is based on the avidity of their TCRs for antigens presented by cortical and medullary thymic epithelial cells. Thymocytes expressing TCRs with insufficient avidity fail to receive survival signals and undergo death by neglect during positive selection, whereas those with excessive avidity are actively eliminated via negative selection to ensure self-tolerance. In contrast, thymocytes with intermediate avidity are positively selected and mature into single-positive cells, either CD4+ or CD8+. Thymocytes with the highest of these intermediate affinities develop in Tregs. Given AIRE’s pivotal function in these processes, sex differences in AIRE expression may significantly influence thymic selection.

The TCR repertoire of female and male peripheral T cells have shown that their diversity, particularly that of the TRB repertoire, is influenced not only by the age of the individuals but also by their sex (15–18). However, another study showed no association between the sex of individuals and the diversity of their TCR repertoire (19). Nevertheless, the study of peripheral blood TCRs reflects that of an immune system that has undergone multiple immune responses. For example, observed differences between males and females in their responses to self-antigens associated with autoimmunity could indicate either the cause (the TCRs drive the disease) or the consequence (the TCRs are expanded because of the disease).

To assess whether there are significant differences in the generation and thymic selection of the TCR repertoire between males and females, we therefore focused on analyzing the TCR repertoires of thymocytes. These cells represent the latest stage in T cell development and have not yet been involved in immune responses. Their repertoires thus represent the mechanisms of TCR generation for the analysis of CD4/CD8 double positive (DP) cells, and of TCR selection for CD4 and CD8 simple positive (SP) T cells. We generated a unique data set comprising the sequences of the TRA and TRB chains of DP cells, CD8 and CD4 SP effector cells, and CD4 SP Tregs from organ donors, both infants and adults. The analysis of TRA and TRB repertoires using state-of-the-art strategies revealed no sex differences in repertoire generation (DP cell repertoire), but an enrichment in females for CD8 SP T cells bearing CDR3 associated with specificities related to autoimmunity.

Results

We analyzed thymic TCRαβ repertoires from 22 human organ donors (11 females, 11 males, aged between 73 days and 64 years) (Figure 1, Supplementary Figure 1). For each donor, CD3+CD4+CD8+ double positive (DP), CD3+CD4-CD8+ CD8 single positive (SP), and CD3+CD4+CD8-CD4 SP thymocytes were sorted by flow cytometry. In 16 donors, CD4 SP cells were further subdivided into CD25-conventional CD4 Teff and CD25+ CD4 Treg subsets, whereas in the remaining six donors, total CD4 SP cells were analyzed as Teff-enriched populations. This resulted in 20 DP, 21 CD8 SP, 22 CD4 Teff SP and 14 CD4 Treg SP samples (Figure 1, Supplementary Figure 1). Bulk TCR sequencing was then performed on each cell subset, with the aim to characterizing the TRA and TRB repertoires. Rank–frequency plots of clonotype abundances displayed comparable heavy-tailed distributions across donors, without any obvious systematic differences between males and females (Supplementary Figure 2).

Schematic overview of the generation of the thymic TCR dataset and the analytical pipeline.

Top Panel: Generation of the thymic TCR dataset. From deceased human thymuses of males and females, we isolated key T-cell subtypes through cell sorting. These subtypes included double-positive (DP) cells (CD3+ CD4+ CD8+), single-positive (SP) CD8+ cells (CD3+ CD4-CD8+), SP CD4+ cells (CD3+ CD4+ CD8-) that were further separated into T effector (Teff, CD3+ CD4+ CD8-CD25-) and Treg (CD3+ CD4+ CD8-CD25+) cells. TCR libraries were generated from the RNA of each cell population using rapid amplification of cDNA ends by PCR (5’RACE PCR). Following sequencing, data preprocessing involved quality sequencing checks, contig alignment, and quality control. The final dataset comprised 20 DP samples (male-to-female ratio of 1:1), 21 SP CD8+ samples (1.1:1), 6 SP CD4+ samples (1:5), 16 SP CD4 Teff samples (1.67:1), and 14 SP CD4 Treg samples (1.33:1). Males are depicted in violet and females in orange. Bottom Panel: Analytical pipeline. We compared the TCR repertoires of males and females across various dimensions. We evaluated general aspects of the TCR repertoire were evaluated, including diversity, gene usage, CDR3aa length distribution, and aa usage within the CDR3 region. Additionally, we analyzed the probability of sequence generation and the TCR repertoire structure based on CDR3aa sequence similarity. We identified differentially expressed TRB CDR3aa motifs between sexes and analyzed TRB CDR3aa sequence specificity. Created with BioRender.com.

Minimal sex-based differences in TCR V and J gene usage across thymic cell subsets

The first level of diversity in generating TCRs is the usage of V and J genes (20). We analyzed the usage of TRA and TRB V and J genes and their combinations in each thymocyte population (Figure 1).

Males and females had a relatively similar V gene usage, as evidenced by the lack of a clear separation in the Principal Component Analysis (PCA), for both TRA and TRB, and across all cell subtypes studied (Figure 2A). However, we observed some individual differences in V and J gene usage between males and females, with some being specific to certain cell subtypes (Supplemental Figures 3-6), and others observed across multiple cell subtypes (Supplemental Figures 3-6). For example, TRBV6-5 was found to be more highly expressed in females compared to males in DP cells (Supplemental Figure 3C) and CD8 SP cells (Supplemental Figure 4C).

Comparable overall TCR gene usage between males and females.

(A) Principal Component Analysis (PCA) derived from the distribution of TRAV (left) and TRBV (right) gene usage frequencies across sex groups (males vs females), showing results for DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and CD4 Treg (N = 14) cells (displayed from top to bottom). Each point on the graph represents an individual. Ellipses indicate 95% confidence intervals. (B) Heatmap showing the Jensen-Shannon Divergence (JSD) score between samples, derived from the distributional usage of TRAV-TRAJ (left) and TRBV-TRBJ (right) gene associations in DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and CD4 Treg (N = 14) cells (displayed from top to bottom). Hierarchical clustering was performed using the Euclidean distance and the complete linkage method. Males are shown in violet; females in orange.

In the following step, we compared the usage of TRAV-TRAJ and TRBV-TRBJ gene combinations across all individuals using the Jensen-Shannon Divergence (JSD) score (Figure 2B). The hierarchical clustering showed no clear separation between male and female individuals for any of the chain or cell subtype, indicating no significant differences in the gene combination usage between males and females (Figure 2B). A similar observation was made when directly analyzing the frequency of these V-J gene combinations (Supplemental Figures 3E-6E, 3F-6F).

Comparable TCR repertoire diversity between males and females

A high diversity of the TCR repertoire is essential for generating a broad potential reactivity against a variety of antigens. We thus compared the TCR repertoire diversity between males and females. To achieve this objective, we analyzed Rényi diversity curves (from zero to infinite 𝛼 parameters) that measure various aspects of TCR repertoire diversity. At different 𝛼 values, the Rényi index places more or less weight on the contribution of the most frequent clonotypes. Higher 𝛼 values highlight the most prevalent clonotypes, thereby revealing the effect of expanded clones on overall diversity.

Rényi curves showed comparable diversities between sexes in DP and CD4 Teff cells, for both TRA and TRB, and a difference in CD8 and CD4 Treg cells (Supplemental Figure 7). In CD8 cells, the curves begin to diverge at low 𝛼 values (Supplemental Figure 7), with an earlier inflection in males, indicating a less balanced distribution of TCR clones. This suggests that in males, fewer clones dominate the repertoire at lower diversity thresholds compared to females. In Tregs, while the inflection point is similar, the difference seems to be characterized by higher diversity and richness in females compared to males (Supplemental Figure 7). However, no significant differences were observed for the Shannon diversity index (𝛼 = 1), the Simpson index (𝛼 = 2), and the Berger-Parker index (𝛼 = infinite), for both chains and all cell subtypes (Figure 3). These results indicate that the diversity of thymic TCR repertoire is similar between males and females.

Comparable thymic TCR repertoire diversity between males and females.

Boxplots display Shannon (A), Simpson (B), and Berger-Parker (C) index values for TRA (left) and TRB (right), across thymic T cell subtypes, displayed from top to bottom, in DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and CD4 Treg (N = 14). Each point on the graph represents the median value from 50 rarefactions per sample. Statistical analysis (Wilcoxon test) showed no significant sex bias in TCR repertoire diversity (p > 0.05). Males are shown in violet; females in orange.

Comparable CDR3aa length distribution in male and female TCR repertoires

Most of the repertoire diversity is generated by the random addition of nucleotides within the VDJ junctions, resulting in the hypervariable CDR3 region that interacts with the peptide-MHC complex, and is mostly responsible for the TCR specificity. The CDR3 length has been shown to directly influence the TCR’s ability to interact with the peptide (21). We thus studied the distribution of CDR3 aa (CDR3aa) sequence lengths between males and females to identify potential TCR generation and/or selection biases (Figure 1).

Our results showed that males and females have a comparable distribution of CDR3aa lengths for both TRA (Supplemental Figure 8) and TRB (Figure 4A). CDR3aa sequences in TRA were shorter than those in TRB, as previously described in peripheral blood T cells (22,23). Although there were subtle differences in certain CDR3aa lengths between males and females, these variations were minimal, involving sequences that represent less than 2% of the total TCR repertoire (Figure 4A and Supplemental Figure 8). Altogether, the overall distribution of CDR3aa length is comparable between males and females.

CDR3aa length and amino acid composition of TRB CDR3s in males and females.

(A) Distribution of TRB CDR3aa length usage in DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and Treg (N = 14) SP cells (displayed from top to bottom). Stars indicate statistical differences between males and females based on the p-value of the Wilcoxon test (*: p < 0.05, **: p < 0.01). (B) The data on amino acid (aa) usage between males and females is presented as the log2 fold change of the median per-donor usage in females over males for each aa in the p108 to p114 CDR3aa region for TRB. A line at log2 fold change = 0 is indicative of the direction of the difference in usage frequency. The bars are colour-coded to the hydropathy class of the aa as defined by the Kyte-Doolittle-based IMGT classification (neutral aa by gray, hydrophilic aa by blue-green and hydrophobic aa by gold). Stars indicate statistical differences of usage between males and females based on the p-value of the Wilcoxon test (*: p < 0.05, **: p < 0.01). (C) Position-specific usage of hydrophobic aa (excluding alanine, due to its weak hydrophobicity) at IMGT positions p109 and p110 in TRB across thymic T cell subtypes, including DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and CD4 Treg (N = 14). For each donor, the values represent the proportion of unique TRB CDR3aa sequences carrying a hydrophobic amino acid at the indicated position. Stars indicate significant sex differences based on the p-value of the Wilcoxon test (*: p < 0.05), with a significant increase in hydrophobic usage at p109 in female CD8 SP cells. Males are depicted in violet; females in orange.

Subtle differences in the composition of CDR3aa in the thymic TCR repertoires of males and females

We analyzed the aa usage within the FG loop of the CDR3, which is in contact with the peptide-MHC complex (24). The classification of aa was determined based on their hydropathy properties, categorized as follows: hydrophilic, hydrophobic, or neutral (25). We observed significant differences in the usage of some aa between males and females for both TRA (Supplemental Figure 9) and TRB (Figure 4B). In DP cells, these differences predominantly involved hydrophobic aa: females exhibited higher alanine (A) usage in TRA and lower in TRB compared with males (Figure 4B and Supplemental Figure 9). Additionally, females showed increased usage of phenylalanine (F) and isoleucine (I) in TRB (Figure 4B). Furthermore, other aa with varying hydropathy properties were differentially used between males and females in CD4 Teff and Treg populations (Figure 4B and Supplemental Figure 9).

We then examined the usage of aa specifically at positions p109 and p110 of the TRB CDR3aa region, as hydrophobic aa at these two positions have been associated with a stronger recognition of self-antigens (26–28). For each individual, we calculated the proportion of unique TRB CDR3aa carrying a hydrophobic aa (excluding alanine, which is only weakly hydrophobic) at these two positions. This positional analysis revealed a significant increase in the usage of hydrophobic amino acid at position p109 in female CD8 cells compared with males, with similar trends observed at p109 in DP and CD4 Teff subsets and at p110 in CD8 cells (Figure 4C). Taken together, these positional patterns suggest a subtle sex-biased enrichment of hydrophobic amino acids at key TRB CDR3 positions that could be compatible with a slightly increased potential for self-reactivity and cross-reactivity in female CD8 T cells (28).

Slight sex-based variations in thymic TCR generation probability distribution

Another approach to evaluating biases in the generation and selection of the TCR repertoire is to compare the probability of generation (Pgen) of CDR3 nucleotide sequences (Figure 1). Pgen represents the likelihood of a sequence being produced, according to models of V(D)J recombination (14,29). For DP and CD8 cells only and for each TCR chain, a sequence generation model was created using 100,000 non-productive nucleotide sequences. Using this model, the Pgen of all productive TCR sequences that make up each repertoire was calculated.

When comparing the distribution of Pgen between males and females, we observe a very slight difference of distribution, characterized by a nearly null Kolmogorov-Smirnov score (DK-S < 0.05) (Figure 5). This difference is more pronounced in the TRB compared to the TRA chain (Figure 5). As this minor discrepancy might be indicative of variations among individuals irrespective of their sex, we have generated 20,000 permuted mixed-sex groups utilizing our dataset. Our findings revealed a significant distribution difference of Pgen sequences between these mixed groups population (DP – TRA: DK-S = 0.0183 ± 0.0145 (µ ± sd), DP – TRB: DK-S = 0.0190 ± 0.0161, CD8 – TRA: DK-S = 0.0227 ± 0.0153, and CD8 – TRB: DK-S = 0.0259 ± 0.0140). However, the observed difference in the Pgen distribution between males and females is higher than that between the mixed groups in the DP TRB.

Probabilities of generation of TCRs in males and females.

The figure shows the distribution of Pgen (probability of generation) sequences between males and females for TRA and TRB in DP and CD8 cells. A V(D)J recombination model was created using 400,000 non-productive random sequences derived from the nonproductive sequences of all individuals for each chain, both for DP and CD8 cells (65). This model was then used to calculate the Pgen of sequences for each individual (66). The overall distribution comparison between males and females was tested using the Kolmogorov-Smirnov test, with the D value and associated p-value indicated. Males are depicted in violet and females in orange.

Comparable TCR repertoire network structure based on CDR3 amino acid sequence similarity between males and females.

For each sample, 100 random subsamplings were performed on the minimum number of CDR3aa per cell subtype. Two CDR3aa are linked if they have a Levenshtein distance of one. (A) TRA Network structure of subsampled TCR repertoire of a male subject for DP, CD8, CD4 Teff and CD4 Treg SP (from left to right). Each point on the graph represents a CDR3aa. (B-C) Comparison of the proportion of linked sequences (B) and network density (C) between male and female samples, in DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and CD4 Treg (N = 14). Each point on the graph represents the median value from 100 subsampling iterations for each sample. Statistical analysis using the Wilcoxon test revealed no significant sex differences for these two metrics (p > 0.05). Males are depicted in violet; females in orange.

Similar network structure of thymic TCR repertoires in males and females

Next, we evaluated whether thymic selection in males and females is directed towards similar or distinct sequences. For each TCR repertoire, CDR3aa sequences are connected if they differ by one aa, i.e. have a Levenshtein distance of =1 (Figure 1). Such linked sequences have a high probability of sharing the same specificities (30). A random downsampling process to the smallest sample was performed to compare samples of comparable sizes. This process was repeated 100 times, and the results are detailed in the Supplementary Table 1. The shapes of the network are depicted in Figure 6A.

To analyze these clusterings with greater precision, we used various metrics. When comparing the number of clustered sequences in each individual network between males and females, no significant differences were observed for any cell type in either chain (Figure 6B). Similarly, the degree of sequence similarity, as indicated by the density of edges between sequences within an individual’s TCR repertoire, was comparable between males and females (Figure 6C). These findings indicate that there are no significant sex-specific differences in the similarity of CDR3aa sequences, suggesting that males and females produce CDR3aa sequences with the same degree of variability within their TCR repertoires.

Absence of sex-specific CDR3aa motifs in thymic TCR repertoires of males and females

We then sought to determine whether we could detect CDR3aa sequence motifs that were more prevalent in one group compared to the other, in order to evaluate whether the selection of CDR3aa was biased towards particular sequences differently between males and females (Figure 1). To identify sequences with shared characteristics that might indicate a probable common specificity among individuals of the same group, the analysis focused exclusively on CDR3aa sequences of the TRB, as this region has been shown to have a greater influence on TCR specificity (31,32). Local motifs were defined using Gliph2 (33), which are a sequences of three to five aa, as well as global motifs, which are sequences of more than three aa where, at one position, an aa can be substituted by another aa if it has a positive BLOSUM62 score (34). Importantly, only motifs containing public CDR3aa sequences shared by at least two individuals were retained, thereby excluding private or individual-specific sequences from the analysis.

We identified several hundred motifs that were differentially expressed between males and females for each cell subpopulation. A greater number of motifs were specific to females, ranging from 426 motifs for DP to 278 motifs for CD4 Tregs, compared to males, ranging from 328 motifs for CD4 Teffs to 34 motifs for CD4 Tregs (Figure 7A). The identified motifs are mainly local motifs and largely confined to their respective sex group, indicating that motifs identified in females are predominantly found in females, and vice versa (Supplemental Figure 10-13).

Thymic TRB TCR sex-associated motifs.

Different structural motifs found differentially expressed between males and females in our dataset. We distinguish local motifs as distinct aa sequences, and global motifs as motif regions with one variable aa position maintaining a BLOSUM62 score of ≥ 0. (A) Number of male and female associated motifs by cell subset. (B) Euler diagram illustrating the distribution and overlap between all sex-associated motifs. The numbers indicate the number of motifs in overlap zones. (C-D) Validation of these sex-associated TRB CDR3aa motifs. The following heatmap illustrates the usage of all the TRB CDR3aa motif in the external thymic pediatric dataset (35,36) (C) and those of TRB CD8 in the peripheral dataset (D). Sex and total CDR3aa number are depicted by sample. Males are depicted in violet and females in orange then local motifs in blue-green and global motifs in magenta.

As illustrated in Figure 7B, we observed minimal motif overlap between different cell subtypes. Only two female associated motifs overlapped: one between CD4 Teffs and CD4 Tregs, and another between CD4 Teffs and CD8 cells (Figure 7B). In contrast, no overlap was found between male-associated motifs (Figure 7B). Furthermore, it was observed that there was an overlap between one motif associated with female CD8 cells and a motif associated with male CD4 Teff cells (Figure 7B).

We then proceeded to evaluate the usage of the differentially expressed motifs in external datasets. There is no publicly available thymic TCR dataset that reports repertoires according to cell subsets. We thus tested these motifs on TCRs from pediatric bulk thymocytes, which contains TCR data from children aged between two and eight months, with a male-to-female ratio of 5:3 (35,36). We calculated the usage of these differentially expressed motifs in each sample of this dataset. We were unable to separate individuals by sex using these motif usages (Figure 7C).

We then evaluated these motifs using a peripheral blood TCR dataset from healthy individuals, where CD8, CD4 Teff, and CD4 Treg cells had been sorted. The usage of these motifs could not distinguish males from females across all cell subsets (Figure 7D and Supplemental Figure 14).

Sex-biased enrichment of TCRs associated with autoimmune diseases and bacterial antigens

We sought to identify the usage of TCRs with known specificity as represented in the IEDB, Mc-PAS and VDJdb public databases (37–39). We compiled the TCRs from these three databases, retaining only those with high sequence reliability scores (see Materials and Methods). We focused on TRB sequences exclusively, due to their greater representation in the databases (31,40). We identified 55,368 unique TRB CDR3aa sequences with high reliability and specificity assignment scores for at least one specificity.

We classified the TCR specificities based on the antigen they recognize: viral or bacterial peptides, or human peptides overexpressed in cancers, autoimmune diseases (AIDs), or neither of these diseases (Supplemental Figure 15A). We sought to identify the TRB CDR3aa sequences in our thymocyte dataset. Depending on the individuals and cell subtypes, they represented between 0.82% and 3.58% of the TRB repertoires.

We observed that unique TRB CDR3aa sequences specific to self-antigens not associated with pathologies and those associated with cancers, are proportionally more represented in our thymic TCR dataset than in the pooled reference database, in both females and males (Supplemental Figure 15B). This pattern was consistent across all cell subtypes (Supplemental Figure 15B). These comparisons to the reference database are presented for descriptive purposes only, as differences between an experimental thymic repertoire and a curated database are expected given the structure of the reference resource. We then compared the distribution of these CDR3aa between males and females for each cell subtype. Strikingly, in the TCR repertoires with one specificity only, we observed a significantly higher proportion of unique TRB CDR3aa sequences associated with AIDs in female CD8 SP cells compared to male CD8 SP cells (Figure 8A). Furthermore, in the specific TCR repertoires only, there was a trend towards a higher proportion of thymic sequences with specificity associated to bacterial compounds in female CD8 SP cells compared to males, and an opposite trend was observed for CD4 Tregs SP (Figure 8A). These differences observed in CD8 SP cells persisted when examining the usage of these sequences (i.e. cumulative usage frequency), with significantly higher usage of sequences with specificity associated to bacterial compounds and those associated with AIDs in females compared to males (Figure 8B). Of particular note are the donor-resolved analyses (Supplementary Figures 15B and 16), in which each individual is represented separately. These confirm that these patterns are not driven by a single donor. Donor age had no significant effect on the sex-biased enrichment of self-and bacteria-specific TRB CDR3 sequences in CD8 SP and CD4 Treg SP cells (p ≥ 0.52), indicating that the observed differences are age-independent (Supplemental Figure 17). In addition, we investigated for the presence of polyspecific TCRs, which are capable of recognizing multiple antigens from different organisms (41–43). These sequences were found to be enriched across all cell subtypes, in comparison to their representation in the reference database (Supplemental Figure 15C). However, no significant differences were observed in the proportion or usage of these polyspecific sequences between males and females. Interestingly, an inverse usage trend was observed between CD8 and CD4 Treg SP cells: females exhibited higher usage of these sequences in the CD8 SP repertoire but a lower usage in the CD4 Treg SP repertoire compared to males (Figure 8C-8D).

Sex-biased enrichment of TCRs specific for known antigens.

From a pooled and curated database, an exact match with this database infers the specificity of TRB CDR3aa of our thymic dataset. Many specificity groups are defined according to the nature of the antigen peptide targeted (bacteria, virus, autoimmune disease [AID], cancer and self-peptide no associated to disease [human]). This analysis compares the distributions of the proportion of unique TRB CDR3aa sequences with a specific specificity (A) and their usage (B) between females and males across cell subtypes, using the log2 fold change of the median values (females over males), following each specificity group. These groups of specificity are additionally classified as microorganism in the top section (bacteria in gold and virus in light blue) and self at the bottom (AID in magenta, cancer in red, human in blue-green). Polyspecific CDR3aa are defined here as CDR3aa capable of recognizing multiple antigens from different organisms (for no self-antigens) or from different specificity groups (e.g. categorized microorganisms, categorized self-antigens, allergens…). The proportion of polyspecific CDR3aa among antigen-specific sequences (C) and their usage (D) is compared between males and females, in DP (N = 22), CD8 (N = 23), CD4 Teff (N = 24) and CD4 Treg (N = 14) cells. Stars indicate statistical differences between males and females based on the p-value of the Wilcoxon test (*: p < 0.05, **: p < 0.01). Males are depicted in violet and females in orange.

In conclusion, our analysis of TCR specificity has identified sex-specific differences in the thymic selection of CD8 effector T cells associated to autoimmune and bacterial antigens that are overrepresented in female versus males.

Discussion

It is noteworthy that there is a paucity of research on potential differences in the TCR repertoire between the sexes, despite the much higher prevalence of autoimmune diseases in females, a topic that remains to be fully elucidated. To explore the hypothesis that the generation and/or the selection of the TCR repertoires may be different between sexes requires, it is necessary to study the TCR repertoire where they are formed and selected, in the thymus. We thus generated a unique and valuable dataset comprising thymic samples collected from deceased organ donors of various ages, and young children who have undergone cardiac surgery. Firstly, we separated the various thymocytes populations in order to study both TCR repertoire generation at the level of DP cells, and repertoire selection at the level of CD8, CD4 Teff and Treg SP cells. We performed bulk sequencing of the TCRs from these cells, as this is the only method that yet generates enough sequences for sensitive comparison at this time. Single-cell sequencing is an analyzing technique that it is limited in its scope. It is only capable of analyzing a few thousand cells and would not have enabled the analyses that can be performed with hundreds of thousands of cells.

TCR generation

Our analyses initially revealed similarities rather than dissimilarities in TCR generation. Our detailed analyses revealed no significant sex differences in relation to the majority of the examined variables when comparing males and females DP cells’ repertoires, except for differences in the usage of small number of V and J genes from both TCR chains. However, these minor discrepancies are likely attributable to random variation, considering the extensive number of comparisons conducted. This finding is further supported by the observation that there are no significant differences in the usage of TRBV-TRBJ and TRAV-TRAJ gene associations between males and females. We also observed no differences in diversity indexes in DP cells’ repertories between males and females. Altogether, the overall combinatorial diversity of the TCR repertoire remains comparable between sexes.

A detailed analysis of the CDR3aa region in DP cells showed that some hydrophobic aa in the FG loop region are differently found for both TCR chains. However, no significant sex-based differences were found in the usage of hydrophobic aa at the critical p109 and p110 positions in TRB that have been described as influencing self-antigen recognition due to the impact of hydrophobic aa on TCR self-reactivity (26–28). These findings suggest that the subtle sex-based differences in the FG loop’s aa usage are minor and do not significantly impact the overall characteristics of the CDR3 region in a sex-dependent manner.

The Pgen analysis of DP cells revealed only minimal differences in the distribution of Pgen between males and females, which were similar to those observed in control groups. This indicates that the mechanisms governing TCR generation are largely consistent across sexes, with only minor, likely random, deviations.

Additionally, although we identified TRB CDR3aa sequence motifs that were differentially expressed between males and females at the DP cell stage in our dataset, these motifs could not reliably differentiate individuals by sex in external datasets. This indicates that they are more dataset-specific than sex-specific. This issue may also be explained by the different characteristics of our data and that of the other dataset. For example, in the Arstila dataset, the analyses were performed on unsorted cells from young subjects. Younger individuals tend to have a more diverse and richer TCR repertoire, with greater repertoire sharing in the periphery, which could potentially obscure sex-specific patterns (16,44). This suggests that these motifs might be context-specific or influenced by other factors such as age, genetic background, or environmental exposures. Moreover, we did not observe any sex-specific differences in CDR3 TRB specificity usage in DP cells’ repertoires. In summary, our results from DP cells analyses could not identify any relevant sex differences in TCR generation.

TCR selection

We identified three genes that show differential usage between males and females. Whilst it is possible that some of these differences are merely random, resulting from multiple comparisons, the preferential use of TRBV6-5 in females has previously been observed in the peripheral TCR repertoire (45). Notably, this gene is overrepresented in female lupus patients compared to their controls (51,52). Although the preferential usage of TRBV6-5 in females is consistent with previous peripheral observations, external validation of this thymic bias remains challenging. Publicly available TCR datasets rarely provide (i) sorted thymocyte subsets, (ii) balanced sex representation, and (iii) sufficient sequencing depth within each subset to allow reliable comparison of gene usage patterns. Moreover, differences in tissue origin (peripheral blood versus thymus), age distribution, and technical protocols may substantially influence TRBV usage frequencies. For these reasons, direct replication in independent thymic cohorts is currently not feasible. Our permutation-based internal control analysis was therefore implemented to ensure that the observed signal was not attributable to random donor grouping or individual-level variation.

Our most striking observation relates to significant sex-specific differences in inferred specificities of CD8 T cells, with a biased CD8 selection process in the thymus favoring sequences associated with autoimmune diseases and bacterial antigens in females. The fact that this bias does not affect DP cells indicates that it is related to TCR selection rather than generation. Its relevance is strongly supported by the fact that (i) there is no such bias for other self-antigens that are not associated with autoimmunity and (ii) the bias affects both the effectors and regulators of autoimmunity with a directionality compatible with the increased prevalence of autoimmune disease in females. TCRs associated with autoimmunity-related self-antigens have been observed to be increased in CD8 effector T cells, while decreased levels have been seen in CD4 Tregs. It should be noted that this bias also affects certain bacterial antigens, though not viral antigens. This phenomenon could be explained by the fact that bacterial antigens express numerous mimotopes of self-antigens, and are known to contribute to the shaping of the TCR repertoire. The observed bias towards bacterial antigens in female CD8 T cells could also provide more effector cells specific for bacterial antigens that mimic self-antigens, contributing to autoimmune disease development by molecular mimicry (48). In this line, antigens linked to celiac disease (CeD) and type 1 diabetes (T1D) are highly represented in the database and these two AIDs have been linked to bacterial components that may mimic self-antigens (49–53). As is well-established, the gut microbiota influences thymic development of T cells (54). Sexual hormones could influence this biased selection by affecting the composition of the gut microbiota (55) and/or by affecting the selection of the TCRs specific for these antigens. In addition to these specific patterns, we observed a significant increase in the usage of hydrophobic amino acids at the central p109 position of the TRB CDR3aa in female CD8 thymocytes, with similar trend at p110. Hydrophobic residues at these positions have been implicated in enhanced self-reactivity and cross-reactivity of CD8 T cells (26–28), suggesting that female CD8 thymocytes may be slightly enriched for TCRs with intrinsically higher potential for recognizing self-like or cross-reactive peptide MHC complexes. It is noteworthy that the positional hydrophobicity of CD8 T cells CDR3s is consistent with the over-representation of sequences annotated as recognizing autoimmune-associated self-antigens, pointing to a more autoreactive TCR repertoire.

Finally, we have recently described polyspecific TCRs that can respond to multiple unrelated viral antigens and hypothesized their possible involvement in autoimmunity (43). We defined a CDR3 as being polyspecific if the TCRs containing it have been experimentally associated with epitopes from at least two different antigen origins. We observed a higher usage of polyspecific sequences in the CD8 SP repertoire of females, although this was not statistically significant. In contrast, the frequency of polyspecific TCR sequence usage was lower in female CD4 Tregs SP cells. These observations, which could be indicative of improved anti-infectious responses and worse autoimmune responses in women, warrant further exploration.

These inferences rely on a harmonized specificity compendium combining McPAS-TCR, IEDB and VDJdb, which provide complementary and only partially overlapping landscapes (pathology-enriched versus predominantly viral specificities). Although there is limited numerical overlap between our thymic repertoires and annotated sequences is limited, this is to be expected given that public databases currently capture only a small fraction of the potential TCR specificity space. In this context, the sex-biased enrichment of sequences annotated as autoimmune-and bacteria-associated should be viewed as a robust signal emerging from a very sparse annotation space. However, it should still be interpreted as hypothesis-generating rather than exhaustive.

To further evaluate the robustness of our findings, we performed additional analyses to assess whether donor age could confound the observed sex-specific differences. A potential limitation of our study is the relatively wide age range of donors. However, the absence of distinct age-related clustering in repertoire features, the lack of age impact on CD8 and CD4 Treg bacterial-specific TRB usage and CD8 and CD4 Treg self-associated to autoimmune disease-specific TRB usage patterns ((Supplemental Figure 17), argue against age as a confounding factor. These results support the hypothesis of an intrinsic sex bias in thymic TCR selection rather than an age-related effect.

While our findings provide valuable insights into sex-specific differences in thymic selection of TCR repertoires, our TCR specificity analyses were limited. We can confirm that our study focused exclusively on the TRB CDR3aa region of the TCR, which is widely acknowledged as the most critical element in defining TCR specificity (31,32) while both the TRA and the TRB chains contribute to TCR specificity (31,32). This limitation is intrinsic to the limited number of TRA with assigned specificity in databases and to the fact that, given the nature of our studies, there is a need to study large repertoires, which is currently not feasible by single cell TCR sequencing. However, as methods for inferring specificity continue to evolve, they will likely allow more refined analyses of our dataset in the future (56). Furthermore, the continuous development of the TCR database should enhance these analyses.

Despite the relatively small size of our cohort, which was outbred, we believe that this design is essential for capturing sex-associated differences. Inbred models would reduce inter-individual variability, but they would not reflect the complexity of human thymic selection, nor the interplay between sex and genetic diversity. Despite the modest sample size, the consistency of the observed patterns across the analytical approaches supports the robustness of our conclusions.

Altogether, our analyses of TCR specificity identified a biased selection in females towards sequences associated with autoimmune diseases, supporting the idea that early central tolerance mechanisms may represent one layer contributing to sex differences in autoimmune susceptibility, in interaction with peripheral and environmental factors. Future research should focus on validating these observations in relevant independent datasets, which are currently lacking.

Methods

Samples

Thymus samples were obtained from twenty-two organ donors aged between 73 days and 64 years, without any specific pathologies (Figure 1). Individual partial HLA typing is detailed in Supplementary Table 2. Adult samples were collected post-surgery at the Cardiac Surgery Departments of many hospitals in France, following approval by the Agence de Biomédecine and the Ministry of Research (#PFS14-009). Pediatric samples were obtained following total thymectomy during cardiac surgeries at Necker Hospital. The ratio of male to female across donors is maintained at 1:1, although it varies between 1.1:1 and 1.33:1 depending on the cell subtypes analyzed. The age distributions of male and female donors were compared using two-sample Kolmogorov–Smirnov tests, with no significant differences observed across thymic subsets (p > 0.46), indicating no age-related confounding (Supplementary Figure 1).

Thymocyte isolation and RNA extraction

Thymic cell suspensions were prepared through automated tissue dissociation using gentleMACS dissociators (Miltenyi®) or by mechanical disruption, followed by filtration through a 70-µm nylon mesh. The cells were stained with anti-CD3 (AF700), anti-CD4 (APC), anti-CD8 (FITC), and for some samples, anti-CD25 (PE) antibodies. Cell sorting was performed by fluorescent activated cell sorting (FACS) on a Becton Dickinson FACSAriaII, achieving a purity of >85%. The sorted populations included DP CD3+ (CD3+CD4+CD8+), SP CD8+ (CD3+CD4-CD8+), and SP CD4+ (CD3+CD4+CD8-). A further separation was made into SP CD4 Teff (CD3+CD4+CD8-CD25-) and SP CD4 Treg (CD3+CD4+CD8-CD25+) for sixteen of the samples (Figure 1). For the sorted SP CD4, the Treg percentage varied between 5.8% and 16% across samples, with an average of 9.63%. Six samples of SP CD4 samples were not sorted into Teffs and Tregs, we thus categorized them as SP CD4 Teffs, assuming a purity of Teffs within the total SP CD4 of around 90%. Please refer to the detailed sample metadata in Supplementary Table 3 of the supplementary material. This includes donor IDs and the number of sorted cells per thymic subset for each donor.

The RNA was isolated using the RNAqueous extraction kit (Invitrogen®), in accordance with the manufacturer’s protocol. The RNA samples were quantified and the integrity of each sample was determined using a Nanodrop (Thermo Fisher) or a Tapestation 4200 (Agilent, RNA ScreenTape).

TCR library preparation and sequencing

TCR library preparation and sequencing were performed as described by Barennes et al. (57). Briefly, RNA was processed using the SMARTer Human TCR a/b profiling v1 kit (Takarabio), in accordance with the manufacturer’s protocol. Amplicons were then purified using AMPure XP beads (Beckman Coulter) and their quantity was determined. The integrity of the amplicons was then assessed using the 2100 Bioanalyzer System (Agilent, DNA 1000 kit) or Tapestation 4200 (Agilent, D1000 screentape). Bulk Next-Generation Sequencing was performed using either a Hiseq 2500 (Illumina) with the SR-300 protocol plus 10% PhiX on the LIGAN-PM Genomics platform (Lille, France), or a Novaseq 6000 (Illumina) with the PE-250 protocol plus 10% PhiX on the LIGAN-PM Genomics platform or ICM platform (Paris, France).

Data processing

The raw FASTQ/FASTA files were aligned to TRA and TRB using MiXCR (v3.0.13, RNA-seq parameters), a software solution that is able to correct for both sequencing and PCR errors (58). Samples with fewer than 1,000 CDR3aa sequences, an imbalanced TRA/TRB ratio (<1/9), or a low sequence count relative to sorted cell counts (< 0.9 ratio) were removed. Sequences with CDR3aa lengths outside 6-23 amino acids were also discarded. To standardize clonotype counts and the number of unique clonotypes across sequencing techniques, the initial counts for each clonotype were reduced by 1 to exclude those with counts of 0 (singletons).

Furthermore, due to the high similarity between certain TRBV sequences, which has the potential to influence the analysis depending on the sequencing method used, TRBV06-2 and TRBV06-3 were merged as TRBV06-2/3, and TRBV12-3 and TRBV12-4 were combined as TRBV12-3/4 (59). Please refer to the summary table provided in the supplementary material (Supplementary Table 3) for a comprehensive overview of the total number of TCR sequences retained after preprocessing, as well as the corresponding number of unique clonotypes, for each donor, chain and thymic subset.

Gene usage analysis

For both TRA and TRB chains across all cell subtypes, we analyzed the V and J gene usage and those of the VJ gene associations (Figure 1).

The usage of gene is calculated as follows:

Where:

Σ𝑖𝜖𝐺Ci represents the total number of counts for clonotypes associated with the specified gene or gene association.

Σ𝑗𝜖𝑆cj represents the total sum of the counts of all clonotypes in the sample.

In this notation, 𝐺 is the set of clonotypes using the interested gene, 𝑆 is the set of all clonotypes in the sample, and ci and cj are the counts of the clonotypes i and j, respectively. The usage of the V gene and the VJ gene combinations was investigated through dimensional reduction using Principal Component Analysis (PCA) (MASS package v7.3-60, factoMineR package v2.8, factoextra package v1.0.7). Confidence ellipses were represented with a 5% alpha risk. Furthermore, for VJ gene combinations usage, a heatmap was constructed using the Jensen-Shannon Divergence (JSD) distance to measure the distance between samples based on their gene usage distribution (complexHeatmap package v2.16.0, philentropy package v0.7.0). Hierarchical clustering was performed using the Euclidean distance and the complete linkage method.

Diversity analysis

For the TCR repertoire diversity analysis, to correct for sequencing depth bias and differences in cellular richness between individuals, rarefactions were performed fifty times for each sample, using X clonotypes, where X represents the effective diversity number (60).

The effective diversity was calculated as the exponential of the Shannon index (𝑒!) (61). Rényi indices (11 indices from 0 to Infinity) were calculated for each rarefaction (62).

The median Shannon index, Simpson index and Berger-Parker indices were then compared between the sexes.

Distribution of CDR3aa length analysis

The calculation of the usage of all CDR3 by the amino acid length as follows:

Where

Σ𝑖𝜖𝐿ci represents the sum of the counts of the CDR3 characterized by the amino acid length of interest.

Σ𝑗𝜖𝑆cj represents the total sum of counts of all CDR3s per sample.

In this notation, 𝐿 is the set of CDR3s characterized by the amino acid length of interest, 𝑆 is the set of all CDR3s in the sample, and ci and cj are the counts of the CDR3s i and j, respectively.

Composition usage of amino acid within the CDR3 region

All CDR3aa sequences were aligned using the IMGT convention (positions p104-p118) (63).

Three complementary scales were used in the study. Firstly, sequence level amino acid composition. Within the CDR3 loop region (IMGT positions p108-p114), involved in MHC-peptide complex interaction (63), amino acid usage at the sequence level was quantified for CDR3aa sequences 9-21 aa long in both TRA and TRB chains. For each sequence, the percentage usage of a given amino acid was calculated as follows: Where 𝑛aa is the number of occurrences of the amino acid of interest in the CDR3 loop, and L denotes the total length of the CDR3aa sequence. Furthermore, amino acids were categorized into three hydropathy classifications (neutral, hydrophobic, and hydrophilic) using the IMGT classification that takes into account the Kyte-Doolittle hydropathy index (25,64)

Secondly, the position-specific amino acid usage at p109 and p110. To assess positional composition, for each individual, CDR3aa usage is examined at IMGT positions p109 and p110, corresponding to p6-p7 positions described by Stadinsky et al. and Khosravi-Maharlooei et al., which hydrophobic amino acids at these two central positions have been implicated in modulating TCR self-reactivity and cross-reactivity (26–28). For a hydrophobic amino acid 𝑕 (excluding alanine, due to its weak hydrophobicity) at position 𝑝, usage was defined: Where 𝑛 is the total number of unique CDR3 sequences and 𝑁(h,p) the number of unique CDR3 sequences carrying hydrophobic amino acid 𝑎 at position 𝑝. This analysis quantifies amino acid composition at fixed positions independently of clonotype frequency.

Probability of generation analysis

The probability of generation analysis of DP and CD8 cell subtypes consisted initially in the creation of a generative model of V(D)J recombination using 400,000 randomly selected non-productive sequences from all samples for both TRA and TRB chains, with the IGOR tool (65). The generation probabilities (Pgen) for all sequences in the dataset were subsequently calculated using the OLGA tool, based on the respective generative models for each cell subtype and TCR chain (66). Comparisons of Pgen values at the nucleotide level between males and females were performed. To ensure statistical robustness, control groups with equivalent male and female sample numbers were created by permuting the samples into two groups 20,000 times.

TCR network structure by sequence similarity

For each sample, 100 random sub-samplings of CDR3aa sequences were performed, using the minimum CDR3aa count per cell subtype (refer to Supplementary Table 1). Two CDR3aa were connected if their Levenshtein distance was = 1 (Figure 1) (67). Network analysis focused on two metrics: the proportion of connected sequences and network density, defined as the ratio of actual connections to all possible connections. Median values from 100 subsampling iterations were then analyzed. Levenshtein distances were computed with the stringdist package (v0.9.10), networks were visualized using Cytoscape (v3.8.2), and network density was calculated with the igraph package (v1.5.1).

TRB CDR3aa motif search

Two types of structural motifs were searched: local motifs (strict sequential motifs) and global motifs (sequences with substitutions preserving a positive BLOSUM62 score) (Figure 1).

The selection of enriched motifs was carried out using the gliph2 function with the turboGliph package (v0.99.2), with specific parameters defined as follows:

-The input sequences were from the sex group of interest (male or female), with reference sequences matched for TRBV gene usage.

-It is imperative that CDR3aa sequences are longer than eight aa.

-Local motif lengths were restricted to 3-5 aa.

-Only motifs consisting of more than two unique CDR3aa were conserved.

-Local motif significance of Fisher’s test has not been boosted.

Additional filters were applied so that: (i) the inclusion of public CDR3aa sequences is a prerequisite for a motif, with these sequences shared by at least two individuals, (ii) a significant enrichment is required, as determined by Fisher’s test with a p-value p < 0.01 and (iii) a usage difference between groups of at least twofold is necessary, as evidenced by a Wilcoxon test with a p-value p < 0.05.

Motif validation was performed on an openly published pediatric thymic dataset from the Arstila team (35,36) consisting in a bulk TCR dataset from infants aged between seven days and eight months. Here, we removed one male subject, “Thymus A”, who was a twin of the “Thymus B” and “Thymus 1” neonate sample, in order to analyze height samples with a male-to-female ratio of 5:3. Further validation was conducted on sorted peripheral blood TCR repertoires (CD8, CD4 Teff and CD4 Treg) from seventy-five healthy volunteers aged from eighteen to eighty-four years (male-to-female ratio of 1.03:1), whose libraries were generated under the same conditions as those of our thymic dataset. They were recruited within the Transimmunom observational trial (NCT02466217) and HEALTHIL-2 (NCT03837093) (68). PBMCs from healthy individuals were isolated using Ficoll density gradient centrifugation and enriched for CD4+ T cells via EasySep™ Human CD4+ T Cell magnetic beads (Stemcell). The samples were then separated into two groups: effector T cells (CD4+ CD25-) and Tregs (CD4+ CD25+). This was done using EasySep™ Human Pan-CD25 magnetic beads (Stemcell). Purity was assessed by flow cytometry based on the expression of CD4 and FoxP3, with a purity threshold of >80%. From the frozen PBMC aliquots, CD8+ T cells (CD3+ CD8+) were sorted using a FACS ARIA II cell sorter.

TRB CDR3aa specificity analysis

We have established a curated and harmonized TCR specificity database by integrating data from three public TCR repositories: Mc-PAS, IEDB, and VDJdb (38,39,69) (data collected as of October 2023) (Figure 1). The harmonization procedure was carried out in accordance with the framework outlined in Jouannet et al. (70) and included:

-Uniform formatting of TRA/TRB TCR sequences and associated metadata,

-Consolidation of redundant entries (identical TRA/TRB TCR, epitope, organism, PubMed ID, and cell subset),

-Maintening separate entries when source databases are incompatible,

Reliability levels for sequences and specificity assignments were quantified using the Verified_score (VS) and Antigen-identification_score (AIS) as defined in (70). The VS, ranging from 0 to 2, reflects the concordance between calculated and curated CDR3 boundaries following the IEDB verification strategy (VS = 2: both TRA and TRB present and verified; VS = 1.1: only TRA verified; VS = 1.2: only TRB verified; VS = 0: no verified chain) and the AIS (0 to 5), which ranks the strength of antigen-identification methods (70).

In this study, we focused our analyses on high-reliability TRB sequences, defined as those with a verified TRB CDR3aa (VS ≥ 1.2) and an AIS corresponding to in vitro T cell stimulation with a pathogen, protein or peptide, or pMHC X-mer sorting (AIS > 3.2, excluding categories 4.1 and 4.2). This ensures that the specificity assignments relied on robust experimental evidence.

Specificity groups were categorized according to antigen class, encompassing viral, bacterial, yeast, parasitic and self-antigens along with antigens derived from wheat, plants or animals. Self-antigens were then annotated into three categories: autoimmune-associated antigens, cancer-associated antigens and self-antigens unrelated to disease. For autoimmune-associated antigens, assignments were based on explicit annotations present in McPAS-TCR,

IEDB and VDJdb. These databases identify celiac disease (CeD) and type 1 diabetes (T1D) related antigens, which were retained as such in the harmonized dataset. For cancer-associated antigens, the following classifications were used: (i) information present in the original databases and (ii) cross-referencing with external cancer antigen resources, including the Cancer Antigenic Peptide Database (de Duve Institute), the Human Protein Atlas and the cancer database caAtlas (71). Please refer to Supplementary Table 4, which provides a comprehensive list of all self-antigens present in the harmonized database. This table also includes their associated disease category and their mapping into the specificity groups used in the manuscript.

This harmonized database was used to identify exact matches with the CDR3aa sequences of our TRB sequences. The enrichment of CDR3aa sequences associated with a given specificity group in male and female thymic repertoires for each cell subtype compared to the specificity distribution in the database was tested. Furthermore, we performed specificity group distributions between male and female samples, as well as was the overall usage of these sequences across entire TCR repertoires.

Additionally, our research concentrated on polyspecific sequences defined as CDR3aa TRB sequences recognizing at least two distinct microbial species or belonging to multiple specificity groups for non-microbial antigens. In this study, polyspecificity is therefore considered at the level of broader antigenic categories rather than individual epitopes and thus reflects functional cross-reactivity patterns aggregated across studies, instead of implying molecular-level multi-epitope recognition. These polyspecific TRB CDR3aa sequences underwent the same statistical analyses as monofunctional sequences.

Statistical analysis

For V-J gene usage, CDR3aa length usage, CDR3aa usage, network similarity and specificity analyses, the comparison between males and females was statistically analysed using the Wilcoxon test. The statistical tests were performed using the ggpubr library (v0.6.0) in R. The number of asterisks denotes the significance of the results: one asterisk indicates a p-value p ≤ 0.05, two asterisk indicate a p-value p ≤ 0.01, and so on until four asterisks, denoting a p-value p ≤ 0.0001. The comparisons of the Rényi curves and Pgen were compared using the Kolmogorov-Smirnov test. The Wilcoxon test was used to test the comparisons of the proportions of specific sequences in male and female subjects relative to those in the specificity database.